---
title: "Site-Specific Shannon Entropy Dynamics at SARS-CoV-2 Spike Protein Variant-Defining Mutation Sites: Temporal Partitioning and Gaussian Mixture Model Classification Across Alpha, Beta, Gamma, Delta, and Omicron Variants of Concern"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    includes:
      in_header: meta_spike_temporal_eda.html
bibliography: references_partitioning.bib
nocite: '@*'
---

# Overview

This script performs a systematic temporal analysis of Shannon entropy dynamics
at SARS-CoV-2 Spike protein mutation sites across six variant groups: Alpha
(B.1.1.7), Beta (B.1.351), Gamma (P.1), Delta (B.1.617.2), Omicron
(B.1.1.529), and D614G. Two independent post-processed datasets are analysed in
parallel — unaligned NCBI US sequences (January 2020 to September 2021,
109,536 sequences; see the `preprocessing_pipeline` vignette for the full
preprocessing workflow) and aligned GISAID US sequences (March 2020 to June
2022, 129,371 sequences; see `analysis/GISAID_data_preprocessing/` for the
GISAID preprocessing pipeline) — to assess consistency of findings across data
sources.

Three temporal partitioning strategies are applied to each dataset: cumulative
windows expanding by one month from a fixed origin, 2-month sliding windows
with one-month overlap, and non-overlapping 2-month disjoint windows. The
cumulative strategy emulates the sequential accumulation of sequences in public
databases, where each partition incorporates all data collected from the study
origin to a given point in time. Under this scheme, entropy estimates are
progressively shaped by the cumulative sequence composition up to each time
point, with earlier and more prevalent variants contributing disproportionately
as the window grows, making this approach well suited for identifying sustained,
long-term shifts in amino acid diversity. The sliding and disjoint strategies,
by contrast, restrict each partition to a fixed 2-month span, thereby limiting
the influence of early-pandemic sequences on later time points. This temporal
localisation enhances sensitivity to transient entropy changes associated with
variant emergence and replacement events, although at the cost of increased
variability due to smaller sample sizes within each window. This multi-strategy
design allows robust findings to be distinguished from artefacts of a particular
temporal aggregation scheme.

Within each partition, per-site Shannon entropy is computed and sites are
assigned to entropy classes via Gaussian mixture model (GMM) clustering. Class
labels are relabeled so that class 1 consistently denotes the highest-entropy 
group. The temporal evolution of these class assignments is the primary object 
of study: a site transitioning into the highest-entropy class reflects the 
emergence and coexistence of multiple amino acid states, where novel 
substitutions initially coexist with ancestral residues before increasing in 
prevalence and, in some cases, approaching fixation in the circulating 
population.

Four output types are produced per variant per partition strategy: (1) grouped
entropy trajectory plots distinguishing defining amino acid substitution sites 
(SNPs) from other mutation sites; (2) per-site GMM entropy class trajectory 
plots on the raw Shannon entropy scale; (3) amino acid frequency tables 
recording the observed residue distribution at each defining SNP site across all
partitions; and (4) GMM class assignment tables summarizing the class label of 
every site at every partition, with total sequence counts (`Nseq`) and fitted 
GMM classes counts (`Nclust`) as summary rows.

Memory usage is substantial with estimated peak consumption at approximately
15 GB RAM when partitioning the full GISAID dataset. To support
reproducibility and facilitate re-runs on memory-constrained systems, partition
objects are serialized to disk immediately after construction and reloaded on
demand; intermediate objects are explicitly removed via `rm()` and `gc()`
throughout. Wall-clock runtime estimates for each partitioning call are
recorded via `system.time()` and printed to the session log.

<!--
  Notes:
  - Estimated peak RAM usage is substantial. Run on the server, not a local machine.
  - Partitioning is checkpoint-based: if RDS files already exist they are loaded
    directly, skipping the expensive re-partitioning step.
  - The main analysis loop processes both datasets (NCBI US, GISAID US) and all
    three windowing strategies for all VOCs (Alpha, Beta, Gamma, Delta, Omicron,
    D614G). Relabeling is performed once per partition key and reused across
    all variants to minimise peak memory consumption.
  - D614G (site 614) is treated as a standalone variant with first US detection
    hardcoded to March 2020. No entropy trajectory plot is generated for D614G
    (single site, nothing to compare); only class trajectory, frequency table,
    and class assignment table are produced.
  - All other sites: detection lines are drawn for every VOC whose Mutation_Sites
    include that site.
  - Omicron entropy trajectory plots show defining SNP sites only (no
    by-group stratification), due to the high number of SNP sites.
  - Class trajectory plots use raw Shannon Entropy on the y-axis throughout.
  - Entropy trajectory plots (grouped VOC sites) use 10 × ln(Shannon Entropy).
  - All source() calls, require() calls, and references to legacy custom
    functions have been eliminated. Only ViralEntropR package functions and
    standard CRAN packages are used.
-->

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  comment = "#>"
)
```

## Libraries

```{r libraries, message=FALSE, warning=FALSE}

library(ViralEntropR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(kableExtra)
library(scales)
library(here)

```

## Output Directories

```{r output-dirs}

out_root <- here("Temporal_Site_EDA", "outputs")

# All partition and variant label combinations are generated programmatically
# to ensure the folder structure is complete and consistent.
partition_labels_all <- c(
  "NCBI_US_cumulative_1m",   "NCBI_US_sliding_2m",   "NCBI_US_disjoint_2m",
  "GISAID_US_cumulative_1m", "GISAID_US_sliding_2m", "GISAID_US_disjoint_2m"
)
voc_labels_all <- c("Alpha", "Beta", "Gamma", "Delta", "Omicron", "D614G")

dirs <- c(
  file.path(out_root, "partitions"),
  file.path(out_root, "Entropy_Trajectories", partition_labels_all),
  unlist(lapply(partition_labels_all, function(p)
    file.path(out_root, "Class_Trajectories", p, voc_labels_all))),
  file.path(out_root, "Class_Assignments", partition_labels_all),
  unlist(lapply(partition_labels_all, function(p)
    file.path(out_root, "Frequency_Tables", p, voc_labels_all)))
)

invisible(lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE))

```

## Log Transformation for Entropy Trajectory Visualisation

```{r transformation}

# Defines the ln(x) * 10 transformation applied to the y-axis of grouped
# entropy trajectory plots (VOC mutation sites). Multiplying the natural
# logarithm of Shannon Entropy by 10 compresses the dynamic range and
# improves visual separation of low-entropy sites while preserving
# monotonicity. Class trajectory plots use raw Shannon Entropy.

log10p_trans <- scales::trans_new(
  name      = "ln_x10",
  transform = function(x) log(x) * 10,
  inverse   = function(x) exp(x / 10),
  breaks    = scales::log_breaks()
)

```

## Site Colour Map

```{r site-colours}

# Fixed color assignments for all defining SNP sites across Alpha, Beta,
# Gamma, Delta, Omicron, and D614G. Sites are keyed by index (character);
# values are color strings. All entries are consistent across all figures.

site_colour_map <- c(

  # Alpha, Beta, Gamma, Delta SNP sites
  "18"   = "#E5C494",
  "19"   = "#FFED6F",
  "20"   = "#8DD3C7",
  "26"   = "#BEBADA",
  "80"   = "#80B1D3",
  "138"  = "#FDB462",
  "190"  = "#FCCDE5",
  "215"  = "#D9D9D9",
  "417"  = "#B3DE69",
  "452"  = "#FF7F00",
  "478"  = "#FFFFB3",
  "484"  = "#FB8072",
  "501"  = "#E41A1C",
  "570"  = "#CCEBC5",
  "614"  = "#8B2500",
  "655"  = "#377EB8",
  "681"  = "#984EA3",
  "701"  = "#4DAF4A",
  "716"  = "#BC80BD",
  "950"  = "#F781BF",
  "982"  = "#FFFF33",
  "1027" = "#A65628",
  "1118" = "#A6D854",

  # Omicron-specific SNP sites
  "76"   = "#1B9E77",
  "95"   = "#D95F02",
  "212"  = "#7570B3",
  "339"  = "#E7298A",
  "371"  = "#66A61E",
  "373"  = "#E6AB02",
  "375"  = "#A6761D",
  "440"  = "#666666",
  "446"  = "#A6CEE3",
  "477"  = "#56B4E9",
  "493"  = "#B2DF8A",
  "496"  = "#33A02C",
  "498"  = "#FB9A99",
  "505"  = "#CAB2D6",
  "547"  = "#6A3D9A",
  "679"  = "#FF6347",
  "764"  = "#20B2AA",
  "796"  = "#DAA520",
  "856"  = "#8B4513",
  "954"  = "#00CED1",
  "969"  = "#9ACD32",
  "981"  = "#FF1493"

)

```

## VOC Colour Map

```{r voc-colours}

# Colours for variant-specific detection vertical lines on class trajectory
# plots, and for column highlighting in GMM class assignment tables.

voc_colour_map <- c(
  Alpha   = "#1F78B4",   # cerulean blue
  Beta    = "#E31A1C",   # red
  Gamma   = "#33A02C",   # dark green
  Delta   = "#FF7F00",   # orange
  Omicron = "#6A3D9A",   # deep violet
  D614G   = "#8B2500"    # dark red
)

voc_table_colours <- c(
  Alpha   = "#AED6F1",   # light blue
  Beta    = "#F1948A",   # light salmon
  Gamma   = "#A9DFBF",   # light green
  Delta   = "#FFD700",   # gold
  Omicron = "#D7BDE2",   # light purple
  D614G   = "#F4A460"    # sandy brown
)

```

# Variant Metadata

```{r variants}

# Extract metadata for all five VOCs from the ViralEntropR package dataset.
# D614G is added as a standalone entry with hardcoded March 2020 US detection.

alpha_idx   <- which(unlist(sarscov2_variants$WHO_Label) == "Alpha")
beta_idx    <- which(unlist(sarscov2_variants$WHO_Label) == "Beta")
gamma_idx   <- which(unlist(sarscov2_variants$WHO_Label) == "Gamma")
delta_idx   <- which(unlist(sarscov2_variants$WHO_Label) == "Delta")
omicron_idx <- which(unlist(sarscov2_variants$WHO_Label) == "Omicron")

voc_list <- list(

  Alpha = list(
    label     = "Alpha",
    pango     = sarscov2_variants$Pango_Lineage[[which(unlist(sarscov2_variants$WHO_Label=="Alpha"))]],
    mut_sites = sarscov2_variants$Mutation_Sites[[alpha_idx]],
    snp_sites = sarscov2_variants$Defining_SNP_Sites[[alpha_idx]],
    snps      = sarscov2_variants$Defining_SNPs[[alpha_idx]],
    us_date   = sarscov2_variants$Date_First_Detected_US[[alpha_idx]]
  ),

  Beta = list(
    label     = "Beta",
    pango     = sarscov2_variants$Pango_Lineage[[which(unlist(sarscov2_variants$WHO_Label=="Beta"))]],
    mut_sites = sarscov2_variants$Mutation_Sites[[beta_idx]],
    snp_sites = sarscov2_variants$Defining_SNP_Sites[[beta_idx]],
    snps      = sarscov2_variants$Defining_SNPs[[beta_idx]],
    us_date   = sarscov2_variants$Date_First_Detected_US[[beta_idx]]
  ),

  Gamma = list(
    label     = "Gamma",
    pango     = sarscov2_variants$Pango_Lineage[[which(unlist(sarscov2_variants$WHO_Label=="Gamma"))]],
    mut_sites = sarscov2_variants$Mutation_Sites[[gamma_idx]],
    snp_sites = sarscov2_variants$Defining_SNP_Sites[[gamma_idx]],
    snps      = sarscov2_variants$Defining_SNPs[[gamma_idx]],
    us_date   = sarscov2_variants$Date_First_Detected_US[[gamma_idx]]
  ),

  Delta = list(
    label     = "Delta",
    pango     = sarscov2_variants$Pango_Lineage[[which(unlist(sarscov2_variants$WHO_Label=="Delta"))]],
    mut_sites = sarscov2_variants$Mutation_Sites[[delta_idx]],
    snp_sites = sarscov2_variants$Defining_SNP_Sites[[delta_idx]],
    snps      = sarscov2_variants$Defining_SNPs[[delta_idx]],
    us_date   = sarscov2_variants$Date_First_Detected_US[[delta_idx]]
  ),

  Omicron = list(
    label     = "Omicron",
    pango     = sarscov2_variants$Pango_Lineage[[which(unlist(sarscov2_variants$WHO_Label=="Omicron"))]],
    mut_sites = sarscov2_variants$Mutation_Sites[[omicron_idx]],
    snp_sites = sarscov2_variants$Defining_SNP_Sites[[omicron_idx]],
    snps      = sarscov2_variants$Defining_SNPs[[omicron_idx]],
    us_date   = sarscov2_variants$Date_First_Detected_US[[omicron_idx]]
  ),

  # D614G: standalone entry. First detected in the US in March 2020, predating
  # all named VOCs. No entropy trajectory plot is generated (single site).
  D614G = list(
    label     = "D614G",
    pango     = NA_character_,
    mut_sites = 614L,
    snp_sites = 614L,
    snps      = "D614G",
    us_date   = "Mar-2020"
  )

)

cat("VOC metadata summary:\n")
for (vl in names(voc_list)) {
  v <- voc_list[[vl]]
  pango_str <- if (is.na(v$pango)) "standalone" else v$pango
  cat(sprintf("  %-8s (%s), US detection: %s\n", vl, pango_str, v$us_date))
  cat("    SNP sites:", paste(sort(v$snp_sites), collapse = ", "), "\n")
}

```

# Data Partitioning (Checkpoint-Based)

Partitioning is the most memory-intensive step. Each dataset is loaded,
partitioned under all three windowing strategies, saved to disk as an RDS
file, and immediately removed from memory before the next dataset is loaded.
On subsequent runs the checkpoint files are detected and re-partitioning is
skipped entirely.

Three complementary windowing strategies are applied:

- **Cumulative (1-month increments):** the start date is fixed at the study
  origin and the window end expands by one calendar month per partition.
- **Sliding (2-month windows):** a 2-month window advances one month at a
  time, producing overlapping consecutive partitions.
- **Disjoint (2-month windows):** consecutive non-overlapping 2-month
  partitions covering the full study period without redundancy.

## NCBI US Data

```{r ncbi-us-partition}

ncbi_rds <- file.path(out_root, "partitions", "NCBI_US_Part_List.rds")

if (!file.exists(ncbi_rds)) {

  message("Partitioning NCBI US data ...")
  AL_df_NCBI <- readRDS(here("NCBI_US_unaligned_feature_matrix_1273aa.rds"))

  # --- 1. Cumulative partitioning: 1-month increments ------------------------
  t_ncbi_cumulative_1m <- system.time({
    ncbi_us_cumulative_1m <- partition_time_windows(
      data          = AL_df_NCBI,
      n_sites       = 1273L,
      window_length = 1L,
      window_type   = 1L,
      start_date    = "2020-01-01",
      end_date      = "2021-09-01",
      verbose       = FALSE
    )
  })
  message("NCBI cumulative_1m partitioning: ",
          round(t_ncbi_cumulative_1m["elapsed"], 1), "s")

  # --- 2. Sliding window partitioning: 2-month windows -----------------------
  t_ncbi_sliding_2m <- system.time({
    ncbi_us_sliding_2m <- partition_time_windows(
      data          = AL_df_NCBI,
      n_sites       = 1273L,
      window_length = 2L,
      window_type   = 2L,
      start_date    = "2020-01-01",
      end_date      = "2021-09-01",
      verbose       = FALSE
    )
  })
  message("NCBI sliding_2m partitioning:    ",
          round(t_ncbi_sliding_2m["elapsed"], 1), "s")

  # --- 3. Disjoint window partitioning: 2-month windows ----------------------
  t_ncbi_disjoint_2m <- system.time({
    ncbi_us_disjoint_2m <- partition_time_windows(
      data          = AL_df_NCBI,
      n_sites       = 1273L,
      window_length = 2L,
      window_type   = 3L,
      start_date    = "2020-01-01",
      end_date      = "2021-09-01",
      verbose       = FALSE
    )
  })
  message("NCBI disjoint_2m partitioning:   ",
          round(t_ncbi_disjoint_2m["elapsed"], 1), "s")

  rm(AL_df_NCBI); gc()

  NCBI_US_Part_List <- list(
    cumulative_1m = ncbi_us_cumulative_1m,
    sliding_2m    = ncbi_us_sliding_2m,
    disjoint_2m   = ncbi_us_disjoint_2m
  )
  rm(ncbi_us_cumulative_1m, ncbi_us_sliding_2m, ncbi_us_disjoint_2m,
     t_ncbi_cumulative_1m, t_ncbi_sliding_2m, t_ncbi_disjoint_2m); gc()

  t_ncbi_save <- system.time({
    saveRDS(NCBI_US_Part_List, ncbi_rds)
  })
  message("NCBI US serialisation:           ",
          round(t_ncbi_save["elapsed"], 1), "s")
  message("NCBI US partition list saved")
  rm(NCBI_US_Part_List, t_ncbi_save); gc()

} else {
  message("NCBI US partition RDS found, skipping re-partitioning.")
}

```

## GISAID US Data

```{r gisaid-us-partition}

gisaid_rds <- file.path(out_root, "partitions", "GISAID_US_Part_List.rds")

if (!file.exists(gisaid_rds)) {

  message("Partitioning GISAID US data ...")
  AL_df_GISAID <- readRDS(here("GISAID_US_aligned_feature_matrix_1273aa.rds"))

  # --- 1. Cumulative partitioning: 1-month increments ------------------------
  t_gisaid_cumulative_1m <- system.time({
    gisaid_us_cumulative_1m <- partition_time_windows(
      data          = AL_df_GISAID,
      n_sites       = 1273L,
      window_length = 1L,
      window_type   = 1L,
      start_date    = "2020-03-01",
      end_date      = "2022-06-01",
      verbose       = FALSE
    )
  })
  message("GISAID cumulative_1m partitioning: ",
          round(t_gisaid_cumulative_1m["elapsed"], 1), "s")

  # --- 2. Sliding window partitioning: 2-month windows -----------------------
  t_gisaid_sliding_2m <- system.time({
    gisaid_us_sliding_2m <- partition_time_windows(
      data          = AL_df_GISAID,
      n_sites       = 1273L,
      window_length = 2L,
      window_type   = 2L,
      start_date    = "2020-03-01",
      end_date      = "2022-06-01",
      verbose       = FALSE
    )
  })
  message("GISAID sliding_2m partitioning:    ",
          round(t_gisaid_sliding_2m["elapsed"], 1), "s")

  # --- 3. Disjoint window partitioning: 2-month windows ----------------------
  t_gisaid_disjoint_2m <- system.time({
    gisaid_us_disjoint_2m <- partition_time_windows(
      data          = AL_df_GISAID,
      n_sites       = 1273L,
      window_length = 2L,
      window_type   = 3L,
      start_date    = "2020-03-01",
      end_date      = "2022-06-01",
      verbose       = FALSE
    )
  })
  message("GISAID disjoint_2m partitioning:   ",
          round(t_gisaid_disjoint_2m["elapsed"], 1), "s")

  rm(AL_df_GISAID); gc()

  GISAID_US_Part_List <- list(
    cumulative_1m = gisaid_us_cumulative_1m,
    sliding_2m    = gisaid_us_sliding_2m,
    disjoint_2m   = gisaid_us_disjoint_2m
  )
  rm(gisaid_us_cumulative_1m, gisaid_us_sliding_2m, gisaid_us_disjoint_2m,
     t_gisaid_cumulative_1m, t_gisaid_sliding_2m, t_gisaid_disjoint_2m); gc()

  t_gisaid_save <- system.time({
    saveRDS(GISAID_US_Part_List, gisaid_rds)
  })
  message("GISAID US serialisation:           ",
          round(t_gisaid_save["elapsed"], 1), "s")
  message("GISAID US partition list saved")
  rm(GISAID_US_Part_List, t_gisaid_save); gc()

} else {
  message("GISAID US partition RDS found, skipping re-partitioning.")
}

```

# VOC Detection Partition Indices and Analysis Configurations

```{r detection-indices-and-configs}

ncbi_labels   <- lapply(readRDS(ncbi_rds),   `[[`, "Dates_Labels")
gisaid_labels <- lapply(readRDS(gisaid_rds), `[[`, "Dates_Labels")

# Returns the index of the first partition whose label string contains the
# target month-year string. Returns NA_integer_ when not found.
find_partition_index <- function(labels, date_str) {
  if (is.null(date_str) || is.na(date_str) || !nzchar(date_str))
    return(NA_integer_)
  idx <- which(vapply(labels, function(x) grepl(date_str, x, fixed = TRUE),
                      logical(1L)))
  if (length(idx) == 0L) NA_integer_ else idx[1L]
}

# Compute detection partition indices for all VOCs (including D614G) across
# all three partition types for a given dataset's Dates_Labels list.
compute_voc_detection <- function(labels_list) {
  lapply(voc_list, function(v) {
    lapply(labels_list, function(lbl) find_partition_index(lbl, v$us_date))
  })
}

voc_detection_ncbi   <- compute_voc_detection(ncbi_labels)
voc_detection_gisaid <- compute_voc_detection(gisaid_labels)

build_partition_configs <- function(dataset_id, voc_detection, titles) {
  partition_keys <- c("cumulative_1m", "sliding_2m", "disjoint_2m")
  lapply(partition_keys, function(pk) {
    list(
      label       = paste0(dataset_id, "_", pk),
      key         = pk,
      title       = titles[[pk]],
      voc_indices = vapply(voc_detection, function(v) {
        idx <- v[[pk]]
        if (is.null(idx) || length(idx) == 0L) NA_integer_
        else as.integer(idx[1L])
      }, NA_integer_)
    )
  })
}

ncbi_titles <- list(
  cumulative_1m = "NCBI US, Cumulative Partitioning, 1-Month Increments",
  sliding_2m    = "NCBI US, Sliding Window Partitioning, 2-Month Windows",
  disjoint_2m   = "NCBI US, Disjoint Window Partitioning, 2-Month Windows"
)
gisaid_titles <- list(
  cumulative_1m = "GISAID US, Cumulative Partitioning, 1-Month Increments",
  sliding_2m    = "GISAID US, Sliding Window Partitioning, 2-Month Windows",
  disjoint_2m   = "GISAID US, Disjoint Window Partitioning, 2-Month Windows"
)

ncbi_configs   <- build_partition_configs("NCBI_US",   voc_detection_ncbi,   ncbi_titles)
gisaid_configs <- build_partition_configs("GISAID_US", voc_detection_gisaid, gisaid_titles)

cat("VOC detection partition indices (cumulative | sliding | disjoint):\n\n")
cat("NCBI US:\n")
for (vl in names(voc_list)) {
  vals <- vapply(ncbi_configs, function(c) {
    xi <- c$voc_indices[[vl]]
    if (is.na(xi)) "NA" else as.character(xi)
  }, character(1L))
  cat(sprintf("  %-8s: %s\n", vl, paste(vals, collapse = " | ")))
}
cat("\nGISAID US:\n")
for (vl in names(voc_list)) {
  vals <- vapply(gisaid_configs, function(c) {
    xi <- c$voc_indices[[vl]]
    if (is.na(xi)) "NA" else as.character(xi)
  }, character(1L))
  cat(sprintf("  %-8s: %s\n", vl, paste(vals, collapse = " | ")))
}

rm(ncbi_labels, gisaid_labels,
   voc_detection_ncbi, voc_detection_gisaid); gc()

```

# Helper Functions

## Partition Relabeling

```{r relabel-helper}

# Relabels all non-sentinel, non-empty cluster DataFrames within a
# partition_time_windows() output so that class label 1 consistently denotes
# the highest-entropy group. Max_Entropy[i] is updated to 1L accordingly.
# Called once per partition key before iterating over all VOCs.

relabel_part_data <- function(part_data) {
  for (i in seq_len(part_data$N_partitions)) {
    df_i <- part_data$Clusters[[i]]$DataFrame
    if (!is.null(df_i) && nrow(df_i) > 0L &&
        !all(df_i$class == 999L, na.rm = TRUE)) {
      part_data$Clusters[[i]]$DataFrame <- relabel_entropy_classes(df_i)
      part_data$Max_Entropy[i]           <- 1L
    }
  }
  part_data
}

```

## VOC Detection Vertical Lines

```{r voc-vlines-helper}

# Returns a list of geom_vline layers marking VOC first-US-detection
# partitions on a class trajectory plot. One pair of dotted flanking lines
# is drawn per VOC whose Mutation_Sites include the site, coloured by variant.
# VOCs whose detection index is NA are silently skipped.
# D614G is handled uniformly alongside the five named VOCs.
# Note: this helper is retained for reference but is not called in the
# class trajectory loop (detection lines have been removed from those plots).

get_site_vlines <- function(site, voc_list, voc_partition_indices,
                             voc_colour_map) {
  layers <- list()
  for (vl in names(voc_list)) {
    if (!(vl %in% names(voc_partition_indices))) next
    if (as.integer(site) %in% as.integer(voc_list[[vl]]$mut_sites)) {
      xi  <- voc_partition_indices[[vl]]
      if (!is.na(xi)) {
        col <- voc_colour_map[[vl]]
        layers <- c(layers, list(
          ggplot2::geom_vline(xintercept = xi - 0.15,
                              colour = col, linetype = "dotted", linewidth = 3),
          ggplot2::geom_vline(xintercept = xi + 0.15,
                              colour = col, linetype = "dotted", linewidth = 3)
        ))
      }
    }
  }
  layers
}

```

## SNP Name Lookup

```{r snp-name-helper}

# Returns the amino acid substitution string (e.g. "P681H") for a given
# site within a VOC entry. Returns the site index as a string if not found.

get_snp_name <- function(site, voc) {
  idx <- which(as.integer(voc$snp_sites) == as.integer(site))
  if (length(idx) == 0L) return(as.character(site))
  voc$snps[idx[1L]]
}

```

## GMM Class Assignment Table

```{r classes-table-helper}

# Constructs a wide-format site-by-partition GMM entropy class assignment
# table and saves it as a styled HTML file.
#
# Layout:
#   Rows    = Spike protein sites (sorted numerically), followed by Nseq and
#             Nclust summary rows.
#   Columns = Partition windows labelled by date range.
#   Values  = GMM class labels after relabeling (class 1 = highest entropy).
#             Em-dashes indicate partitions in which a site was absent.
#   Nseq    = Total number of sequences in each partition window.
#   Nclust  = Number of fitted GMM entropy components in each partition.
#
# Highlighting:
#   VOC detection columns : each column highlighted in its variant colour.
#   Defining SNP rows     : bold, marked with * in the Site column.
#                           No background highlighting (avoids colour overload).
#   Nseq row              : bold, light grey (#E8E8E8).
#   Nclust row            : bold, medium grey (#D0D0D0).
#
# A footnote is appended listing the variant-to-colour correspondence and
# noting that Beta and Gamma were detected in the US concurrently in
# January 2021, causing their detection columns to coincide.

make_classes_table <- function(data_frame, part_data,
                               voc_indices = NULL,
                               snp_sites   = NULL,
                               voc_colours = NULL,
                               save_path, table_name) {

  df        <- data_frame
  df$sites  <- as.integer(as.character(df$sites))
  df$class  <- as.integer(as.character(df$class))
  df$period <- as.integer(df$period)

  wide <- tidyr::pivot_wider(
    df[, c("sites", "period", "class")],
    names_from  = "period",
    values_from = "class"
  )

  period_cols    <- as.integer(colnames(wide)[-1L])
  col_labels     <- part_data$Dates_Labels[period_cols]
  colnames(wide) <- c("Site", col_labels)
  wide           <- wide[order(wide$Site), ]

  wide         <- as.data.frame(wide, stringsAsFactors = FALSE)
  wide[, -1L]  <- lapply(wide[, -1L], as.character)
  wide[is.na(wide)] <- "\u2014"

  # Mark defining SNP sites with asterisk in the Site column.
  # Applied before Nseq/Nclust rows are appended so that only site rows
  # receive the marker. Rows are bolded (not highlighted) after kbl() is built.
  if (!is.null(snp_sites) && length(snp_sites) > 0L) {
    snp_site_chars    <- as.character(snp_sites)
    is_snp            <- wide$Site %in% snp_site_chars
    wide$Site[is_snp] <- paste0(wide$Site[is_snp], "*")
  }

  # Nseq: total sequences per partition.
  nseq_vals <- vapply(period_cols, function(i) {
    nrow(part_data$Partitions[[i]])
  }, integer(1L))

  # Nclust: number of fitted GMM components per partition.
  nclust_vals <- vapply(period_cols, function(i) {
    df_i <- part_data$Clusters[[i]]$DataFrame
    if (is.null(df_i) || nrow(df_i) == 0L) return(0L)
    cl <- df_i$class
    if (all(cl == 999L, na.rm = TRUE)) return(1L)
    length(unique(cl[cl != 999L]))
  }, integer(1L))

  nseq_row_df <- as.data.frame(
    matrix(c("Nseq",   as.character(nseq_vals)),   nrow = 1L),
    stringsAsFactors = FALSE
  )
  colnames(nseq_row_df) <- colnames(wide)

  nclust_row_df <- as.data.frame(
    matrix(c("Nclust", as.character(nclust_vals)), nrow = 1L),
    stringsAsFactors = FALSE
  )
  colnames(nclust_row_df) <- colnames(wide)

  wide        <- rbind(wide, nseq_row_df, nclust_row_df)
  n_all_rows  <- nrow(wide)
  n_site_rows <- n_all_rows - 2L

  tbl <- kableExtra::kbl(
    wide,
    align   = "c",
    escape  = FALSE,
    caption = paste("GMM Entropy Class Assignments by Partition:", table_name)
  ) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width        = TRUE,
      font_size         = 11
    )

  # Highlight each VOC detection column in its variant colour.
  if (!is.null(voc_indices) && length(voc_indices) > 0L) {
    for (vl in names(voc_indices)) {
      xi <- voc_indices[[vl]]
      if (!is.na(xi)) {
        voc_col_pos <- which(period_cols == xi) + 1L
        if (length(voc_col_pos) == 1L) {
          bg <- if (!is.null(voc_colours) && vl %in% names(voc_colours))
            voc_colours[[vl]] else "#FFD700"
          tbl <- kableExtra::column_spec(tbl, voc_col_pos,
                                         background = bg, bold = TRUE)
        }
      }
    }
  }

  # Bold defining SNP site rows (marked with *).
  snp_rows <- grep("\\*$", wide$Site[seq_len(n_site_rows)])
  if (length(snp_rows) > 0L)
    tbl <- kableExtra::row_spec(tbl, snp_rows, bold = TRUE)

  # Style Nseq and Nclust summary rows.
  tbl <- tbl |>
    kableExtra::row_spec(n_all_rows - 1L,
                         bold = TRUE, background = "#E8E8E8") |>
    kableExtra::row_spec(n_all_rows,
                         bold = TRUE, background = "#D0D0D0")

  # Colour legend footnote.
  if (!is.null(voc_colours) && length(voc_colours) > 0L) {
    colour_items <- vapply(names(voc_colours), function(vl) {
      sprintf(
        paste0(
          '<span style="display:inline-block;width:14px;height:14px;',
          'background-color:%s;border:1px solid #999;margin-right:4px;',
          'vertical-align:middle;"></span>&nbsp;%s'
        ),
        voc_colours[[vl]], vl
      )
    }, character(1L))
    footnote_text <- paste0(
      "Column highlight colours by variant: ",
      paste(colour_items, collapse = " &nbsp; "),
      ". Defining SNP sites are indicated with * and shown in bold. ",
      "Note: Beta (B.1.351) and Gamma (P.1) were first detected in the ",
      "US concurrently in January 2021; their detection columns coincide ",
      "and one highlight colour overlays the other."
    )
    tbl <- tbl |>
      kableExtra::footnote(general        = footnote_text,
                           escape         = FALSE,
                           general_title  = "")
  }

  if (!dir.exists(save_path))
    dir.create(save_path, recursive = TRUE, showWarnings = FALSE)

  file_name <- file.path(save_path,
                          paste0(table_name, "_GMM_class_assignments.html"))
  kableExtra::save_kable(tbl, file = file_name)
  message("GMM class assignment table saved: ", file_name)

  invisible(tbl)
}

```

# Main Analysis: Entropy Trajectories, Class Trajectories, Frequency Tables,
# and Class Assignment Tables

All VOCs (Alpha, Beta, Gamma, Delta, Omicron, D614G) are analysed for both
NCBI US and GISAID US data under all three windowing strategies.

For each VOC the following outputs are generated:

- **Entropy trajectory plot** (Alpha, Beta, Gamma, Delta, Omicron only): all
  mutation sites, with defining SNPs as Group 1 and remaining sites as Group
  2; for Omicron, defining SNP sites only without by-group stratification.
  Y-axis:  10 x ln(Shannon Entropy) . Not produced for D614G (single site).
- **Class trajectory plots** (all VOCs including D614G): one per defining SNP
  site. Y-axis: raw Shannon Entropy.
- **Amino acid frequency tables** (all VOCs including D614G): one HTML table
  per defining SNP site, showing per-partition amino acid count distributions.
- **GMM class assignment table** (all VOCs including D614G): wide-format
  site-by-partition class label matrix with Nseq and Nclust summary rows and
  a color legend footnote.

```{r main-analysis, fig.width=20, fig.height=14, message=FALSE, warning=FALSE}

analysis_datasets <- list(
  list(label   = "NCBI_US",
       rds     = ncbi_rds,
       configs = ncbi_configs),
  list(label   = "GISAID_US",
       rds     = gisaid_rds,
       configs = gisaid_configs)
)

for (ds in analysis_datasets) {

  message("\n========================================")
  message("Dataset: ", ds$label)
  message("========================================")

  Part_List <- readRDS(ds$rds)

  for (cfg in ds$configs) {

    message("\n  Partition strategy: ", cfg$label)

    # Relabel once per partition key, reused across all VOCs.
    part_data_rl <- relabel_part_data(Part_List[[cfg$key]])

    for (voc_label in names(voc_list)) {

      voc <- voc_list[[voc_label]]
      pango_str <- if (is.na(voc$pango)) "" else paste0(" (", voc$pango, ")")
      message("    VOC: ", voc_label, pango_str)

      # ------------------------------------------------------------------
      # Entropy trajectory parameters.
      # D614G: no entropy trajectory plot.
      # Omicron: SNP sites only, no by-group stratification.
      # All others: mutation sites with SNPs as Group 1.
      # ------------------------------------------------------------------
      skip_entropy_plot <- voc_label == "D614G"

      if (voc_label == "Omicron") {
        et_sites   <- voc$snp_sites
        et_bygroup <- FALSE
        et_groups  <- NULL
        et_ltypes  <- NULL
        et_lsizes  <- NULL
        et_lsize   <- 1.5
        et_title   <- paste0(
          "Omicron (", voc$pango, "): Defining SNP Sites\n", cfg$title
        )
      } else if (voc_label == "D614G") {
        et_sites   <- voc$snp_sites
        et_bygroup <- FALSE
        et_groups  <- NULL
        et_ltypes  <- NULL
        et_lsizes  <- NULL
        et_lsize   <- 1.5
        et_title   <- NULL   # not used
      } else {
        et_sites   <- voc$mut_sites
        et_bygroup <- TRUE
        et_groups  <- list(voc$snp_sites)
        et_ltypes  <- c("solid", "dashed")
        et_lsizes  <- c(3, 1.5)
        et_lsize   <- 1.5
        et_title   <- paste0(
          voc_label, pango_str,
          ": Defining SNPs (Group 1) vs. All Mutation Sites (Group 2)\n",
          cfg$title
        )
      }

      # Call plot_entropy_trajectories to obtain traj$Data_Frame,
      # traj$XBreaks, and traj$XLabels regardless of VOC — these are needed
      # for class trajectories, frequency tables, and class assignment tables.
      traj <- tryCatch(
        plot_entropy_trajectories(
          part_data        = part_data_rl,
          sites            = et_sites,
          site_colors      = site_colour_map[as.character(et_sites)],
          by_group         = et_bygroup,
          groups_list      = et_groups,
          line_type_groups = et_ltypes,
          line_size_groups = et_lsizes,
          line_size        = et_lsize,
          transformation   = if (skip_entropy_plot) NULL else log10p_trans,
          legend           = TRUE,
          # Increased legend text size for better legibility of
          # site and group labels on the right-hand side of entropy plots.
          legend_text_size = 14,
          plot_title       = et_title
        ),
        error = function(e) {
          message("      Skipping ", voc_label, " (", cfg$label, "): ",
                  conditionMessage(e))
          NULL
        }
      )

      if (is.null(traj)) next

      # ------------------------------------------------------------------
      # Entropy trajectory plot: save for all VOCs except D614G.
      # ------------------------------------------------------------------
      if (!skip_entropy_plot) {

        voc_xi       <- cfg$voc_indices[[voc_label]]
        entropy_plot <- traj$Plot +
          suppressWarnings(
            ggplot2::scale_y_continuous(
              trans  = log10p_trans,
              name   = "10 \u00d7 ln(Shannon Entropy)",
              expand = ggplot2::expansion(mult = c(0.05, 0.05))
            )
          ) +
          ggplot2::xlab("Partition Window") +
          ggplot2::theme(
            plot.title   = ggplot2::element_text(size = 18, hjust = 0.5),
            axis.title.x = ggplot2::element_text(size = 20),
            axis.title.y = ggplot2::element_text(size = 20),
            axis.text.x  = ggplot2::element_text(size = 14),
            axis.text.y  = ggplot2::element_text(size = 14)
          )

        if (!is.na(voc_xi)) {
          entropy_plot <- entropy_plot +
            ggplot2::geom_vline(
              xintercept = voc_xi,
              colour     = voc_colour_map[[voc_label]],
              linetype   = "longdash",
              linewidth  = 5
            )
        }

        ggplot2::ggsave(
          filename = file.path(
            out_root, "Entropy_Trajectories", cfg$label,
            paste0(voc_label, "_entropy_trajectory.png")
          ),
          plot   = entropy_plot,
          width  = 20, height = 15, dpi = 300
        )
        rm(entropy_plot); gc()
      }

      # ------------------------------------------------------------------
      # Class trajectory plots: one per defining SNP site.
      # Raw Shannon Entropy on y-axis (no log transform applied).
      # y-axis label forced via suppressWarnings(scale_y_continuous)
      # to override any label set inside plot_site_class_trajectory(),
      # ensuring "Shannon Entropy" is displayed regardless of the installed
      # function version.
      # ------------------------------------------------------------------
      df_cls <- traj$Data_Frame

      for (site in voc$snp_sites) {

        snp_name <- get_snp_name(site, voc)
        safe_snp <- gsub("[^A-Za-z0-9]", "", snp_name)

        p <- tryCatch(
          plot_site_class_trajectory(
            data_frame = df_cls,
            site       = site,
            site_color = site_colour_map[as.character(site)],
            xbreaks    = traj$XBreaks,
            xlabels    = traj$XLabels,
            label_size = 5,
            line_size  = 3,
            plot_title = paste0(
              "GMM Entropy Class Trajectory: Spike Site ", site,
              " (", snp_name, "), ", voc_label, "\n", cfg$title
            )
          ) +
            # Complete scale replacement — suppresses the expected
            # "scale for y already present" warning produced by ggplot2 when
            # a second scale_y_continuous is added. This guarantees the label
            # reads "Shannon Entropy" regardless of the installed package version.
            suppressWarnings(
              ggplot2::scale_y_continuous(
                name   = "Shannon Entropy",
                expand = ggplot2::expansion(mult = c(0.4, 0.4))
              )
            ) +
            ggplot2::xlab("Partition Window") +
            ggplot2::theme(
              plot.title   = ggplot2::element_text(size = 22, hjust = 0.5),
              axis.title.x = ggplot2::element_text(size = 24),
              axis.title.y = ggplot2::element_text(size = 24),
              axis.text.x  = ggplot2::element_text(size = 16),
              axis.text.y  = ggplot2::element_text(size = 16)
            ),
          error = function(e) {
            message("      Skipping class trajectory site ", site, ": ",
                    conditionMessage(e))
            NULL
          }
        )

        if (is.null(p)) next

        ggplot2::ggsave(
          filename = file.path(
            out_root, "Class_Trajectories", cfg$label, voc_label,
            paste0("Site_", site, "_", safe_snp, "_class_trajectory.png")
          ),
          plot   = p,
          width  = 20, height = 15, dpi = 300
        )
        rm(p, snp_name, safe_snp); gc()
      }

      # ------------------------------------------------------------------
      # Amino acid frequency tables.
      # ------------------------------------------------------------------
      for (site in voc$snp_sites) {
        tabulate_site_evolution(
          partitions     = part_data_rl$Partitions,
          site_index     = site,
          labels         = part_data_rl$Dates_Labels,
          alphabet_size  = 25L,
          zeros          = FALSE,
          return_table   = FALSE,
          save           = TRUE,
          save_path      = file.path(out_root, "Frequency_Tables",
                                     cfg$label, voc_label),
          save_extension = ".html"
        )
      }

      # ------------------------------------------------------------------
      # GMM class assignment table.
      # ------------------------------------------------------------------
      make_classes_table(
        data_frame  = traj$Data_Frame,
        part_data   = part_data_rl,
        voc_indices = cfg$voc_indices,
        snp_sites   = voc$snp_sites,
        voc_colours = voc_table_colours,
        save_path   = file.path(out_root, "Class_Assignments", cfg$label),
        table_name  = paste0(voc_label, "_", cfg$label)
      )

      rm(traj, df_cls); gc()

    }  # end VOC loop

    rm(part_data_rl); gc()

  }  # end partition config loop

  rm(Part_List); gc()

}  # end dataset loop

```

# Session Information

```{r session-info}

sessionInfo()

```

# References

