---
title: "ViralEntropR: Functional Clustering of Per-Site Shannon Entropy Curves and Pairwise Hellinger Trajectories Around WHO Variant Emergence Dates"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
    df_print: kable
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: kable
bibliography: references_fda.bib
always_allow_html: true
nocite: |
  @*
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
library(kableExtra)
```

# Overview

This study treats per-site Shannon entropy and pairwise Hellinger
distance, computed across temporal partitions of the SARS-CoV-2 Spike
protein, as **functional data** [@RamsaySilverman2005] and clusters
the resulting per-site curves and trajectories around the US first-
detection dates of the WHO-labelled SARS-CoV-2 variants. The
clustering is performed in a Functional Data Analysis (FDA) framework
using agglomerative hierarchical clustering [@Sangalli2010] as
implemented in the `fdacluster` package, with the number of clusters
chosen by silhouette width [@Rousseeuw1987] over the grid
$k \in \{2, 3, 4, 5, 6\}$.

The design is factorial. **Twelve** WHO-labelled variants — five
Variants of Concern (Alpha, Beta, Gamma, Delta, Omicron) and seven
Variants of Interest (Epsilon, Zeta, Eta, Iota, Theta, Kappa,
Lambda) — are crossed with two independent surveillance datasets
(NCBI US and GISAID US) and three temporal-partitioning strategies
(2-month sliding, 2-month disjoint, 1-month cumulative). A cell is
one `(variant, dataset, strategy)` combination. The cross is
$12 \times 2 \times 3 = 72$ cells in design.

Three terminal cell statuses occur in production:

- **`ok`** (52 cells) — full clustering pipeline ran end-to-end.
- **`underpowered_bin`** (17 cells) — at least one bin in the cell's
  window contains fewer than `MIN_SEQUENCES_PER_BIN = 30L` sequences,
  so the per-site entropy and Hellinger estimates would be unreliable
  for that bin and the cell is excluded from clustering. Concentrated
  on low-prevalence VOIs (Epsilon, Eta, Iota, Zeta), where regionally
  sparse US deposition cannot support the bin-density threshold.
- **`out_of_coverage`** (3 cells) — the variant's US detection date
  lies outside the dataset's coverage. Affects only Omicron × NCBI US:
  Omicron was first detected in the US in December 2021, but the NCBI
  US feature matrix ends in September 2021.

Each `ok` cell yields **three** functional clustering solutions —
one on Shannon entropy curves, one on Hellinger trajectories anchored
at the window-start partition $T_1$, and one on Hellinger
trajectories anchored at the partition containing the *predecessor*
variant's US detection date (denoted $T_{\text{pred}}$). The study
therefore reports $52 \times 3 = 156$ cell-level clustering
solutions, together with the corresponding static publication
figures and animated visualisations.

The animation pipeline is **per-frame**: rather than fitting a single
cell-level clustering and rendering the same cluster assignments
across every bin, the pipeline fits a fresh GMM and fresh `fdahclust`
on every five-partition fitting sub-window (centred at each interior
partition of the cell). Each frame thus shows the clustering as it
would have been concluded *if a researcher had stopped sampling at
that moment in time*. The resulting GIFs reveal when the variant's
defining mutations converge into a single cluster and how stable
that convergence is across time. See §[Methodology] for details.

The headline question is descriptive rather than predictive: **do the
Spike positions retained by Gaussian Mixture Model (GMM) entropy site
selection organise into temporally coherent groups around each
variant's emergence, and is that organisation reproducible across
the three windowing strategies and two independent surveillance
datasets?** The evaluation comprises four metric families covering
internal clustering quality (silhouette width), robustness to site
resampling (bootstrap Jaccard stability), methodological concordance
(cross-strategy Adjusted Rand Index), and biological plausibility
(Fisher exact enrichment of defining-SNP sites within clusters).

# Terminology

The following terms appear throughout the document. Definitions are
written for readers approaching the work from either the statistical,
virological, or functional-data-analysis side; cross-references to
standard sources are given where applicable.

| Term | Definition |
|------|------------|
| Spike protein | The 1,273-amino-acid surface glycoprotein of SARS-CoV-2. The primary target of immune surveillance and of this study. Reference: GenBank accession YP_009724390. |
| Variant | A SARS-CoV-2 lineage carrying defining amino-acid substitutions, classified by the WHO. This study analyses all twelve WHO-labelled variants with non-missing `Date_First_Detected_US` in the `sarscov2_variants` catalogue: the five Variants of Concern (VOCs: Alpha, Beta, Gamma, Delta, Omicron) and seven Variants of Interest (VOIs: Epsilon, Zeta, Eta, Iota, Theta, Kappa, Lambda). |
| Surveillance dataset | A real-world feature matrix of dated Spike-protein sequences. This study uses NCBI US [@Sayers2022] (137,132 raw sequences, 109,536 post-processed) and GISAID US [@ShuMcCauley2017] (129,371 post-processed). |
| Feature matrix | A `data.frame` with one row per dated sequence, one column for the collection date, and 1,273 integer-coded columns for the 1,273 Spike positions, encoded under a 25-symbol amino-acid alphabet (20 standard residues, the three IUPAC ambiguous codes B, Z, X, the stop codon `*`, and the alignment gap `-`). |
| Cell | One `(variant, dataset, strategy)` combination. The study has 72 cells in design; 69 are attempted (3 are `out_of_coverage`), of which 52 reach `ok` status and 17 are marked `underpowered_bin`. |
| Detection date $d_v$ | The first US detection date of variant $v$, taken from `sarscov2_variants$Date_First_Detected_US` and parsed to a first-of-month `Date` by `parse_month_year()`. |
| Per-cell window | The contiguous date range $[d_v - K \cdot b_{\max}, \; d_v + K \cdot b_{\max}]$ snapped to first-of-month boundaries, where $K = 5$ is the number of bins on each side and $b_{\max} = 2$ months is the largest bin width used. The window is clipped to the dataset's coverage at runtime. When clipping at one edge reduces the realised $K$ on that side, the cell records both `K_pre_realised` and `K_post_realised`. |
| Partition / bin | A contiguous date sub-range within the per-cell window over which residue frequencies and the per-site entropy are computed. The three strategies differ in how the window is partitioned into bins (see below). Bins are denoted $T_1, T_2, \ldots, T_n$ in temporal order. |
| Strategy 1: 2-month sliding | A 2-month bin advances one month at a time. A 20-month window yields $n = 19$ overlapping bins. Adjacent bins share 1 month of data and differ by 1 month. `partition_time_windows(window_type = 2L, window_length = 2L)`. |
| Strategy 2: 2-month disjoint | Non-overlapping 2-month bins. A 20-month window yields $n = 10$ bins, each containing 2 months of distinct data. `partition_time_windows(window_type = 3L, window_length = 2L)`. |
| Strategy 3: 1-month cumulative | An expanding bin that begins at the window's start and grows by one month at each step. Bin $T_k$ contains all sequences from window-start through month $k$. A 20-month window yields $n = 20$ nested bins. `partition_time_windows(window_type = 1L, window_length = 1L)`. |
| Reference bin $T_1$ | The first bin of the window. For Hellinger trajectories its residue distribution serves as one of two anchor distributions (see *Dual anchors* below). |
| Predecessor anchor $T_{\text{pred}}$ | The partition containing the US detection date of the focal variant's *predecessor* in the canonical Spike-evolution lineage (Alpha → D614G, Beta/Gamma/Theta/Kappa/Lambda → Alpha, Delta → Beta, Omicron → Delta, Epsilon/Zeta/Eta/Iota → D614G). When the predecessor's detection date falls outside the cell window, the anchor is snapped to the nearest in-window partition and the cell records `anchor_Tpred$snapped = TRUE`. |
| Shannon entropy $H(s, T_k)$ | The Shannon entropy of the residue distribution at Spike position $s$ in bin $T_k$, in bits: $H(s, T_k) = -\sum_{a} \hat p_a \log_2 \hat p_a$, where $\hat p_a$ is the empirical proportion of residue $a$ at position $s$ among the bin's sequences. Computed by `ViralEntropR::calculate_entropy()`. |
| Per-site entropy curve | The vector $\mathbf{e}_s = (H(s, T_1), H(s, T_2), \ldots, H(s, T_n))$. Functional clustering treats $\mathbf{e}_s$ as the discrete sampling of a continuous function of time and clusters sites with similar functions. |
| Hellinger distance $D_H(P, Q)$ | A bounded metric on probability distributions: $D_H(P, Q) = \frac{1}{\sqrt 2}\sqrt{\sum_a (\sqrt{p_a} - \sqrt{q_a})^2}$. Bounded in $[0, 1]$ when normalised; this study uses the unnormalised form, bounded in $[0, \sqrt 2]$. Reference: [@vanDerVaart1998, ch. 14]. |
| Per-site Hellinger trajectory | The vector $\mathbf{h}_s = (D_H(P_{s, T_2}, P_{s, T_a}), \ldots, D_H(P_{s, T_n}, P_{s, T_a}))$, of length $n - 1$, where $T_a \in \{T_1, T_{\text{pred}}\}$ is the chosen anchor. Computed by `ViralEntropR::calculate_hellinger_matrix()` with the anchor passed as `reference_idx`. |
| GMM (Gaussian Mixture Model) | A probabilistic clustering model that fits a weighted sum of $G$ Gaussian components to a univariate sample. This study uses `mclust::Mclust` [@Scrucca2016] via `ViralEntropR::cluster_sites_by_entropy()` over $G \in \{1, \ldots, 15\}$ components and selects the best fit by Bayesian Information Criterion. |
| Class 1 (highest-entropy sites) | After `relabel_entropy_classes()` standardises labels so that class 1 = highest mean entropy, this is the set of Spike positions in the highest-entropy GMM component for the cell's window. Functional clustering is applied only to these class-1 sites; sites in classes 2 to $G$ are not analysed. |
| FDA (Functional Data Analysis) | A statistical framework that treats observations indexed by an ordered variable (here, time) as samples of an underlying smooth function. References: [@RamsaySilverman2005; @KokoszkaReimherr2017]. |
| `fdahclust` | The hierarchical agglomerative clustering routine in the `fdacluster` package [@Sangalli2010], which builds a dendrogram of functional curves under a user-specified linkage and metric, with optional curve alignment. This study uses `warping_class = "none"` (no curve alignment) and the default $L^2$ metric. |
| log1p transformation | The transformation $x \mapsto \log(1 + x)$, applied to both entropy curves and Hellinger trajectories before functional clustering. Reduces the leverage of high-value tails while preserving order, maps zero to zero, and avoids $\log 0$ divergence. Reference: [@BoxCox1964]. |
| Silhouette width $s_i$ | A standard internal clustering quality measure. For each clustered curve $i$, $s_i = (b_i - a_i) / \max(a_i, b_i)$ where $a_i$ is its mean within-cluster distance and $b_i$ is the smallest mean between-cluster distance. Average $\bar s$ ranges in $[-1, 1]$; $> 0.5$ decisive, $0.25$--$0.5$ reasonable, $< 0.25$ weak. Reference: [@Rousseeuw1987]. |
| Bootstrap Jaccard stability | The stability of a clustering under site resampling [@Hennig2007]. Sites are resampled with replacement $B = 100$ times, the clustering is recomputed on each resample, and for each original cluster the maximum Jaccard similarity against any resample cluster is recorded. Per-cluster mean Jaccard $\geq 0.75$ indicates a stable cluster; $0.60$--$0.75$ a patterned cluster; $< 0.60$ instability. |
| ARI (Adjusted Rand Index) | A measure of agreement between two clusterings of the same items, corrected for chance. ARI $= 1$ indicates perfect agreement; $0$ indicates agreement at the chance level; $< 0$ indicates worse-than-chance agreement [@HubertArabie1985]. This study reports cross-strategy ARI to quantify whether the choice of partitioning strategy materially changes the clustering. |
| SNP enrichment (Fisher) | A one-tailed Fisher exact test of independence between cluster assignment and defining-SNP-site membership, computed on the cell's class-1 sites only [@Agresti1992]. Tests whether the variant's defining mutations concentrate disproportionately in one cluster. Multiple-comparison correction uses Bonferroni across `N_TOTAL_ENRICHMENT_TESTS = 220` (12 variants × 2 datasets × 3 strategies × 3 metrics minus the 9 metric-level slots in the 3 `out_of_coverage` cells, plus a small safety margin). |
| `underpowered_bin` | Cell status assigned when at least one bin in the per-cell window contains fewer than `MIN_SEQUENCES_PER_BIN = 30L` sequences. The cell records the worst offending bin in `status_reason` and is excluded from cell-level FDA, all downstream metric tables, and animations. |
| `out_of_coverage` | Cell status assigned when the variant's US detection date $d_v$ lies outside the dataset's coverage range. Affects only Omicron × NCBI US. |

# Research questions

- **Q1.** Around each variant's US first-detection date, do the GMM-
  selected class-1 Spike positions partition into temporally coherent
  groups of entropy curves and Hellinger trajectories under
  agglomerative functional clustering?
- **Q2.** How sensitive is the resulting clustering to the choice of
  temporal-partitioning strategy (2-month sliding vs 2-month disjoint
  vs 1-month cumulative)? The cross-strategy ARI quantifies the
  answer per cell.
- **Q3.** Is the clustering reproducible across two independent
  surveillance datasets (NCBI US and GISAID US) for the variants
  whose detection dates fall within both datasets' coverage?
- **Q4.** Does the clustering organise the variant's defining-SNP
  sites into a single cluster, providing biological plausibility for
  the cluster structure?
- **Q5.** How robust are the clusters to perturbation of the input
  curves by site resampling? Bootstrap Jaccard stability addresses
  the question per cluster within each cell.
- **Q6.** *When* do the defining sites converge into a single
  cluster? The per-frame design records, for each `(cell, metric)`,
  the first frame at which the SNP / Mutation / class-1 convergence
  predicates become TRUE; reported in
  `tab_01_internal_quality.csv$convergence_first_*`.

# Datasets

```{r dataset-table}
datasets_tbl <- data.frame(
  Dataset = c("NCBI US", "GISAID US"),
  `Date range` = c("January 2020 -- September 2021",
                   "January 2020 -- March 2024"),
  Months = c(21L, 51L),
  `Sequences (post-processed)` = c("109,536", "129,371"),
  Source = c("Zenodo DOI 10.5281/zenodo.19040165",
             "GISAID EpiCov, processed via `analysis/GISAID_data_preprocessing/`"),
  check.names = FALSE
)
kbl(datasets_tbl, caption = "Surveillance datasets analysed.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

Both feature matrices are integer-encoded under the 25-symbol
alphabet, sorted ascending by collection date, and date-clipped to
the first of each collection month at preprocessing time
[@Sayers2022; @ShuMcCauley2017]. Site columns occupy positions 1
through 1,273; the `Date` column follows. The clipping is automatic
via `AL_df$Date <- as.Date(format(corrected_dates, "%Y-%m-01"))` in
both preprocessing pipelines, which means every sequence's `Date`
is unambiguously assignable to a bin in any window whose start and
end are themselves month boundaries.

# Ground truth

The ground-truth catalogue is the `sarscov2_variants` dataset bundled
with `ViralEntropR`, restricted to the twelve WHO-labelled variants
with a non-missing `Date_First_Detected_US`. For each variant, three
pieces of catalogue metadata enter the analysis:

- **Detection date $d_v$** = `Date_First_Detected_US` after
  `parse_month_year()` conversion to a first-of-month `Date`. Used
  as the centre of the per-cell window.
- **Defining-SNP sites** = `Defining_SNP_Sites[[v]]`, the integer
  vector of Spike positions whose amino-acid substitutions define
  the variant in standard nomenclature. Used as the biological
  reference set in the SNP enrichment test (Q4).
- **Mutation sites** = `Mutation_Sites[[v]]`, the broader set of
  positions reported as mutated in the variant's defining lineage.
  Used as a secondary convergence target in the per-frame analysis.

```{r truth-table}
truth_tbl <- data.frame(
  Variant     = c("Alpha", "Beta", "Gamma", "Delta", "Omicron",
                  "Epsilon", "Zeta", "Eta", "Iota", "Theta",
                  "Kappa", "Lambda"),
  Class       = c("VOC", "VOC", "VOC", "VOC", "VOC",
                  "VOI", "VOI", "VOI", "VOI", "VOI",
                  "VOI", "VOI"),
  `US first detection` = c("Dec 2020", "Jan 2021", "Jan 2021",
                           "Mar 2021", "Dec 2021",
                           "Mar 2020", "Mar 2020", "Nov 2020",
                           "Nov 2020", "Mar 2021",
                           "Apr 2021", "Aug 2021"),
  Predecessor = c("D614G", "Alpha", "Alpha", "Beta", "Delta",
                  "D614G", "D614G", "D614G", "D614G",
                  "Alpha", "Alpha", "Alpha"),
  `In NCBI US window`   = c("yes", "yes", "yes", "yes",
                            "no (out-of-coverage)",
                            "yes", "yes", "yes", "yes",
                            "yes", "yes", "yes"),
  `In GISAID US window` = c("yes", "yes", "yes", "yes", "yes",
                            "yes", "yes", "yes", "yes",
                            "yes", "yes", "yes"),
  check.names = FALSE
)
kbl(truth_tbl,
    caption = "Variant detection dates, predecessor map, and dataset-coverage status. Omicron's December 2021 first-US-detection date lies outside NCBI's September 2021 coverage, dropping three cells (3 strategies × Omicron × NCBI) from the 72-cell design.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

# Methodology

## Per-cell window construction

For each `(variant, dataset)` pair, a candidate window is constructed
as $[d_v - K \cdot b_{\max}, \; d_v + K \cdot b_{\max}]$, where
$K = 5$ and $b_{\max} = 2$ months, giving a 20-month nominal window.
The candidate window is then snapped to first-of-month boundaries
and clipped to the dataset's available date range. A cell is marked
`out_of_coverage` and excluded from production when the detection
date itself lies outside the dataset's coverage; this affects only
Omicron × NCBI US.

A further bin-density check is applied at the strategy level. Once
the strategy has partitioned the (clipped) window into bins, any cell
with at least one bin containing fewer than
`MIN_SEQUENCES_PER_BIN = 30L` sequences is marked `underpowered_bin`
and excluded from cell-level FDA and all downstream tables. The
`status_reason` field records the worst offending bin (e.g., *"2 / 16
bins have < 30 seqs (worst = 2)"*), which lets downstream readers
distinguish a near-miss from a structural failure.

## Three windowing strategies

Within the per-cell window, three partitioning strategies are
applied in parallel. Each strategy generates its own bin grid, its
own per-site entropy curves, and its own Hellinger trajectories.
The intent is to make the methodological knob "how do we discretise
time?" an *explicit experimental factor* rather than a hidden
assumption.

```{r strategy-table}
strategy_tbl <- data.frame(
  Strategy = c("2-month sliding",
               "2-month disjoint",
               "1-month cumulative"),
  `partition_time_windows args` = c(
    "window_length = 2L, window_type = 2L",
    "window_length = 2L, window_type = 3L",
    "window_length = 1L, window_type = 1L"
  ),
  `Bins per 20-mo window` = c("19", "10", "20"),
  Properties = c(
    "Overlapping; adjacent bins share 1 month; high temporal resolution at the cost of dependence between adjacent bins.",
    "Independent; non-overlapping 2-month chunks; the cleanest statistical interpretation.",
    "Nested; each bin contains all previous data; gives a smoothed, monotone-coverage trajectory."
  ),
  check.names = FALSE
)
kbl(strategy_tbl,
    caption = "Three windowing strategies applied to each (variant, dataset) cell.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Per-window GMM site reduction

For each cell, the GMM site-reduction step of the ViralEntropR
pipeline is applied *within the cell's window only* — no oracle
leakage from the rest of the dataset is permitted. The procedure
mirrors `analysis/Changepoint_Detection_Study/helpers_hellinger.R::fit_window_gmm()`:

1. Per-site Shannon entropy is computed across all sequences in the
   window (one scalar per Spike position).
2. Invariant sites (entropy $\approx 0$) and singleton sites (entropy
   $= H(\frac{1}{n_w}, \frac{n_w - 1}{n_w})$, where $n_w$ is the
   number of sequences in the window) are filtered.
3. `mclust::Mclust` is fitted over $G \in \{1, \ldots, 15\}$ with
   model selection by BIC.
4. `relabel_entropy_classes()` standardises labels so that class 1 =
   the GMM component with the highest mean entropy.
5. The class-1 site set is extracted. The cell is marked
   `reduced_skipped` (and excluded from the metric tables) when the
   GMM returns the all-identical 999 sentinel, when class 1 is
   empty, or when class 1 contains fewer than
   `MIN_CLASS1_SITES_FOR_FDA = 4` sites. The minimum-site threshold
   is set so that silhouette over $k \in \{2, \ldots, 6\}$ is well-
   defined at the upper end of the grid.

The GMM is fit *once* per cell (not once per strategy), because the
GMM operates on across-window entropy and is therefore strategy-
agnostic. The class-1 site set is then propagated to all three
strategies of that cell.

## Per-site entropy curves and Hellinger trajectories

For each strategy of each cell, with the class-1 site set fixed:

- **Per-site entropy curve** $\mathbf{e}_s$ is the vector of
  per-bin Shannon entropies at position $s$. Computed by `apply()`-ing
  `ViralEntropR::calculate_entropy()` over the bin's sequence
  matrix. Shape: $(n_{\text{class-1 sites}}, n_{\text{bins}})$ matrix
  of entropy values.
- **Per-site Hellinger trajectory** $\mathbf{h}_s$ is the vector
  of pairwise Hellinger distances from a designated *anchor bin* to
  every other bin in temporal order, at position $s$. Computed by
  `ViralEntropR::calculate_hellinger_matrix()` with the anchor bin
  passed as `reference_idx` and the class-1 sites passed as the
  `sites` argument. Shape: $(n_{\text{class-1 sites}}, n_{\text{bins}} - 1)$.

**Dual anchors.** Each cell produces *two* Hellinger trajectories
rather than one:

1. **$T_1$ anchor.** The anchor is the window-start partition. This
   is the conventional choice and answers "how has the residue
   distribution at site $s$ evolved relative to the start of the
   observation window?".
2. **$T_{\text{pred}}$ anchor.** The anchor is the partition
   containing the predecessor variant's US detection date. The
   predecessor map encodes the canonical Spike-evolution lineage:
   Alpha and the early VOIs (Epsilon, Zeta, Eta, Iota) anchor on
   D614G (April 2020, [@Korber2020]); Beta, Gamma, Theta, Kappa, and
   Lambda anchor on Alpha; Delta anchors on Beta; Omicron anchors
   on Delta. When the predecessor's detection date lies outside the
   cell's window, the anchor is snapped to the nearest in-window
   partition and the cell records `anchor_Tpred$snapped = TRUE`.

Both matrices are then log1p-transformed elementwise prior to
clustering. The transformation $x \mapsto \log(1 + x)$ leaves zeros
fixed, is order-preserving, and substantially reduces the leverage
of high-entropy / high-Hellinger sites on the $L^2$ distance between
curves.

## Functional clustering via fdahclust

Each $(n_{\text{class-1 sites}} \times n_{\text{time points}})$
matrix is then passed to `fdacluster::fdahclust()` with the
following parameters:

- `warping_class = "none"`: no curve alignment is performed. The
  time axis is bin index, not biological time within a bin, and we
  do not want clusters to be defined by within-bin temporal warping.
- Default $L^2$ metric between curves.
- `centroid_type = "mean"`, `linkage = "complete"`.
- Number of clusters $k$ chosen by average silhouette width over
  $k \in \{2, 3, 4, 5, 6\}$.

`fdahclust` returns a hierarchical agglomerative clustering object
with cluster assignments per site. The cell's clustering result is
serialised together with the entropy and Hellinger matrices, the
chosen $k$, the silhouette value, and the GMM and window metadata.

## Per-frame animation pipeline

In addition to the cell-level (full-window) clustering described
above, the pipeline computes a *per-frame* clustering used by the
animation system. A **frame** corresponds to a five-partition fitting
sub-window centred at an interior partition of the cell's window.
With $h =$ `FITTING_WINDOW_HALFWIDTH = 2`, the set of valid frame
centres is $\{h+1, \ldots, n_{\text{bins}} - h\}$, giving 15 frames
for sliding-2m cells (19 bins), 6 frames for disjoint-2m cells (10
bins), and 16 frames for cumulative-1m cells (20 bins).

For each frame, three independent re-fits run for the three metrics
(entropy, Hellinger $T_1$, Hellinger $T_{\text{pred}}$):

1. **Per-frame GMM.** All sequences in the five partitions centred at
   the frame's central bin are pooled, per-site Shannon entropy is
   recomputed, and a fresh `Mclust` fit selects the per-frame class-1
   site set. The cell-level class-1 set is *not* propagated.
2. **Per-frame curve matrix.** Per-site entropy curves or Hellinger
   trajectories are computed for the frame's class-1 sites *over the
   full cell window* (not just the five partitions). The fitting
   window selects which sites enter clustering; the rendered curves
   span the entire window so the reader can see what happens outside
   the fitting slice.
3. **Per-frame `fdahclust` + silhouette.** $k$ is selected by mean
   silhouette over the grid $\{2, \ldots, 6\}$, exactly as in the
   cell-level fit. Per-site silhouette widths are recorded so that
   marginal sites can be rendered separately in the visualisation
   (state-1 sites with $s_i < 0$).
4. **Convergence record.** Three boolean predicates are evaluated:
   `converged_snp` (all variant-defining SNP sites that are present
   in this frame's class-1 share a single cluster), `converged_mut`
   (same for all variant-defining mutation sites), and
   `converged_class1` (all well-clustered class-1 sites of this frame
   share a single cluster). The "first frame at which the predicate
   becomes TRUE" is the headline number reported per cell.

### Three clustering-quality guards

A frame's clustering is treated as informative only when all three of
the following guards pass:

- **Guard A.** The per-frame GMM must produce $G > 1$. When $G = 1$
  the frame is *null*: there is no class-1 site set and the bottom
  panel renders all variant-defining sites red.
- **Guard B.** The frame's mean silhouette width must be $\geq 0.25$,
  the Kaufman–Rousseeuw "reasonable structure" threshold
  [@KaufmanRousseeuw1990]. Frames with $\bar s < 0.25$ are marked
  `structurally_null` and convergence predicates return FALSE
  regardless of cluster assignment patterns.
- **Guard C.** Per-site silhouette $s_i \geq 0$. Class-1 sites with
  $s_i < 0$ are marginal — they straddle clusters and would be drawn
  to a different cluster centroid on a small input perturbation.
  Such sites are rendered red and excluded from the convergence
  predicates.

### Oversized-cluster visualisation gate

A separate visualisation-only gate handles the case where the
clustering produces a single very-large cluster that would otherwise
dominate the plot. A cluster is **oversized** when its size exceeds
$M =$ `MAX_CLUSTER_SIZE = 33` (Omicron's defining mutation count,
the largest among the 12 WHO variants and therefore a biologically
motivated upper bound on what a real variant signature could look
like). Sites assigned to an oversized cluster are rendered in light
grey in both panels. The convergence predicates also require all
qualifying sites to share a cluster that is *not* oversized — i.e.,
an "all sites in one giant cluster" event does not count as
convergence.

The gate is purely visual. The underlying RDS payload and the
`tab_02_cluster_membership.csv` table record the unfiltered cluster
assignments. Readers and downstream analyses see the full picture;
the plot is just rendered without the visually overwhelming cluster.

### Dual-panel rendering

Each frame is rendered as a two-panel composite. The **top panel**
shows the entropy or Hellinger curves over the full cell window: all
1,273 Spike positions appear as faint light-grey background lines,
the frame's class-1 sites are drawn over them in cluster colours
(red if marginal, light grey if oversized), a vertical dashed line
marks the variant's detection bin, and two open triangles
($\triangleright$ / $\triangleleft$) mark the fitting sub-window
boundaries. The **bottom panel** is a two-row track of the variant's
defining sites along positions 1–1,273: row "SNP" is
`Defining_SNP_Sites`, row "Mutation" is `Mutation_Sites`, and each
site is coloured by its frame state (cluster colour if well-clustered
class-1, light grey if oversized cluster, red otherwise). The frames
are composed into an animated GIF per `(cell, metric)` at
`ANIMATION_FPS = 1L`, and a static "summary" PNG using the full-
window cell-level fit is also written per `(cell, metric)`.

## Evaluation metrics

Four metric families are computed per cell, per metric type (entropy,
Hellinger $T_1$, Hellinger $T_{\text{pred}}$).

```{r metric-table}
metric_tbl <- data.frame(
  Family = c("Internal quality",
             "Stability under resampling",
             "Cross-strategy concordance",
             "Biological plausibility"),
  Metric = c("Silhouette width $\\bar s$",
             "Bootstrap Jaccard stability",
             "Adjusted Rand Index (ARI)",
             "Fisher exact SNP enrichment"),
  Reference = c("[@Rousseeuw1987]",
                "[@Hennig2007]",
                "[@HubertArabie1985]",
                "[@Agresti1992]"),
  Granularity = c(
    "One value per `(cell, metric)`. Reported in `tab_01_internal_quality.csv`.",
    "One value per `(cell, metric, cluster)`. Reported in `tab_03_stability.csv`.",
    "One value per `(variant, dataset, metric, strategy_pair)`. Reported in `tab_04_cross_strategy_ari.csv`.",
    "One Fisher $p$-value and Bonferroni-adjusted $p$-value per `(cell, metric)`. Reported in `tab_05_snp_enrichment.csv`."
  ),
  check.names = FALSE
)
kbl(metric_tbl,
    caption = "Four metric families and their cross-cell granularity.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

A fifth table, `tab_02_cluster_membership.csv`, is descriptive (not
a metric) and lists each cluster's site composition and SNP overlap;
it is the bridge between the clustering and its biological
interpretation. A sixth table, `tab_06_per_frame.csv`, surfaces the
per-frame diagnostics used by the animation pipeline. All six CSVs
are written by `aggregate_tables.R` from the per-cell RDS files in
`outputs/cells/`. Their column-level data dictionaries are given in
§[Supplementary tables (data dictionary)].

The choice of metrics targets four independent failure modes. A
clustering can be:

- *Decisively separated but unstable* (high silhouette, low Jaccard) —
  the cluster boundaries are well-resolved on the observed data but
  would re-organise under modest perturbation.
- *Stable but uninformative* (high Jaccard, low silhouette) — the
  algorithm consistently produces the same partition, but the
  partition is poorly motivated.
- *Internally good but strategy-dependent* (high silhouette and
  Jaccard, low cross-strategy ARI) — the chosen partitioning
  strategy materially determines the conclusion.
- *Internally good but biologically incongruent* (high silhouette
  and Jaccard, non-significant SNP enrichment) — the clusters are
  real statistically but do not align with the variant's known
  mutation profile.

Reporting all four allows reviewers to triangulate cluster meaning
rather than relying on any single quality index.

# Computational budget and runtime

The numbers below are **measured** from the production run on the
shared server `tcell` (125 GiB RAM, 60-core class; Ubuntu Linux),
not pre-run estimates.

```{r runtime-table}
runtime_tbl <- data.frame(
  Phase = c(
    "Phase 1a: cell production (first pass)",
    "Phase 1b: cell production (OOM-recovery pass)",
    "Phase 2: aggregate_tables.R",
    "Phase 3: plot_results.R",
    "Phase 4: render_existing.R (post-hoc cluster_membership CSVs)",
    "Phase 5: animate_results.R"
  ),
  `Wall-clock time (measured)` = c(
    "11 h 49 min",
    "5 h 59 min",
    "6 s",
    "≈ 40--50 min",
    "40 min 32 s",
    "single-pass; runtime recorded in the run log"
  ),
  `N workers` = c("12 (callr::r_bg)", "5 (callr::r_bg)", "1", "1", "8 (xargs -P 8)", "1"),
  `Outcome` = c(
    "67 / 72 cells written; 5 SIGKILL (OOM) on Omicron / Alpha / Beta / Kappa / Lambda × heavy strategies",
    "5 / 5 OOM-affected cells re-run cleanly",
    "6 supplementary CSVs written (216 / 5058 / 469 / 141 / 216 / 1764 rows)",
    "1,920 PNGs written (52 ok cells × 3 metrics × ≈ 12 frames + summary panels)",
    "156 `cluster_membership_<metric>.csv` files written (52 ok cells × 3 metrics)",
    "Up to 156 `anim_<metric>.gif` files written"
  ),
  check.names = FALSE
)
kbl(runtime_tbl,
    caption = "Measured wall-clock budget per analysis phase. Phase 1 totals 17 h 48 min, dominated by Omicron × `cumulative_1m` and Delta × `cumulative_1m` on GISAID US.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Memory footprint per worker.** The feature matrix is loaded once
at subprocess startup (≈ 140–165 MB), and the per-cell working set
is bounded by the window sub-matrix (≈ 10–40 MB depending on
coverage) plus the GMM fit, the three `fdacluster` cell-level
objects, the per-frame `fdacluster` / `Mclust` intermediates, and
the full 1273-site entropy and Hellinger matrices (≈ 200 KB each).
Sustained worker RSS during the first-pass run ranged 3–6.5 GB,
with transient spikes substantially above that on
Omicron × cumulative_1m × GISAID US and a handful of other heavy
cells — five of these spikes exceeded the 10 GB/worker headroom
implied by `N_WORKERS = 12` on the 125 GB server and were terminated
by the kernel OOM-killer. These five cells were re-run in the
Phase 1b recovery pass at `N_WORKERS = 5` (≈ 25 GB/worker headroom),
which succeeded for all five. **Recommended production setting on
hosts with ≥ 64 GB RAM: `N_WORKERS = 6` or `N_WORKERS = 8`**, which
both gives every cell enough headroom for the worst case and avoids
the over-subscription regime that caused the OOM kills.

**Storage footprint per cell (RDS):** ≈ 600 KB for the three full-
1273-site matrices, ≈ 150 KB for the three class-1 matrices,
≈ 100 KB for the per-frame records (~15 frames × 3 metrics × ~25
fields each), ≈ 50 KB for the cell-level fits / stability /
enrichment. Total ≈ 1 MB per cell × 72 cells ≈ 70 MB of cell RDS
files. The OOM-recovered cells observed in the production run
ranged 270 KB (NCBI × Alpha × cumulative_1m, terminal status
`underpowered_bin` after the recovery) to 1.06 MB (GISAID × Lambda ×
sliding_2m).

**Figure / animation storage:** ≈ 1,920 frame PNGs × 500 KB +
156 summary PNGs × 500 KB + up to 156 GIFs × 1 MB ≈ 1.2 GB total in
`outputs/plots/` and `outputs/gifs/`.

# Reproducibility

- **Deterministic seeding.** Each cell receives a deterministic seed:
  `BASE_SEED + dataset_id*1e8 + variant_id*1e6`. With
  `BASE_SEED = 2025L`, `dataset_id ∈ {1, 2}`, and
  `variant_id ∈ {1, …, 12}`, the maximum seed is well within
  `.Machine$integer.max` and is used to seed both the cell-level
  GMM EM initialisation, the per-frame GMM fits (each with an
  additional `+100 + centre_bin` offset for the frame), and the
  bootstrap Jaccard resampling.
- **Atomic writes.** All RDS and CSV files are written via
  `save_rds_atomic()` / `write_csv_atomic()`, which rename the file
  from a per-PID temporary, ensuring no partially written file can
  be loaded by a downstream step.
- **Schema-aware resume.** On restart, `fda_analysis.R` scans
  `outputs/cells/cell_*.rds`, loads each, and skips any cell whose
  `schema_version` matches the current run. Older v1 RDS files are
  flagged stale and re-run. Downstream scripts (`plot_results.R`,
  `animate_results.R`, `aggregate_tables.R`, `render_existing.R`)
  regenerate their outputs from the cell RDS files and are
  therefore cheap to re-run.
- **No oracle leakage.** Per-cell GMM is fit on the cell's window
  only, not on the full dataset. Per-frame GMM is fit on the
  frame's five partitions only. Class-1 site selection is therefore
  locally determined at every level; differences across cells
  reflect biology, not training protocol.
- **Version pinning.** R and `ViralEntropR` version strings are
  recorded in every per-cell RDS and reproduced in the session-
  information block at the end of this README.

# Implementation summary

The implementation follows the same module conventions established
in `Changepoint_Detection_Study/`:

- **Concurrency.** `N_WORKERS = 1L` runs sequentially in-process
  with `tryCatch` per cell; `N_WORKERS > 1L` dispatches via
  `callr::r_bg` with a worker pool; `N_WORKERS = NULL` auto-detects
  with a RAM-budgeted cap of 64.
- **I/O.** Atomic RDS and CSV writes; the per-cell schema is
  documented under §[Per-cell output schema] below.
- **Resume.** On restart, the orchestrator scans `outputs/cells/`
  for existing files and re-runs only the missing ones.
- **Logging.** Structured log via `log_msg()` to
  `outputs/logs/orchestrator_<timestamp>.log`; per-cell errors
  caught and recorded in `outputs/logs/error_log.txt` without
  interrupting the orchestrator. All `stop()` and `warning()` calls
  use `call. = FALSE` per package convention.

Environment overrides at orchestrator startup:

```bash
N_WORKERS=8 VIRAL_FDA_STUDY_DIR=/path/to/FDA_Analysis Rscript fda_analysis.R
```

The production runs used `N_WORKERS = 12` for the first pass and
`N_WORKERS = 5` for the OOM-recovery pass. On hosts with ≥ 64 GB
RAM, `N_WORKERS = 6`–`8` is recommended as the steady-state setting
(see §[Computational budget and runtime] for the empirical
justification).

# Per-cell output schema

Each `outputs/cells/cell_<dataset>__<variant>__<strategy>.rds`
contains a named list with the following 61 fields:

```r
list(
  schema_version,                       # 2L

  # Cell identifiers
  dataset, variant, strategy,
  dataset_id, variant_id, strategy_id,

  # Window construction
  detection_date,                       # parse_month_year(...)
  window_start, window_end,             # Date, snapped to month boundary
  K_pre_realised, K_post_realised,      # may differ from K = 5 at edges
  n_seqs_window,                        # total sequences in window

  # Predecessor metadata
  predecessor_name,                     # "D614G", another variant, or NA
  predecessor_us_date,                  # Date or NA
  predecessor_status,                   # "ok" / "no_predecessor" / "snapped" / ...

  # Bin grid (strategy-dependent)
  bin_starts, bin_ends, bin_labels, bin_counts,
  n_bins, detection_bin,

  # Hellinger anchor indices
  anchor_T1    = list(bin_idx, snapped, distance_months),
  anchor_Tpred = list(bin_idx, snapped, distance_months),

  # GMM site reduction (full-window, shared across metrics)
  gmm_meta = list(G, modelName, n_class1_sites,
                  class1_sites, reduced_skipped, reason),
  snp_sites_truth,                      # integer vector
  mutation_sites_truth,                 # integer vector

  # FULL 1273-site matrices for background panels
  entropy_full_matrix,                  # 1273 x n_bins
  hellinger_T1_full_matrix,             # 1273 x (n_bins - 1)
  hellinger_Tpred_full_matrix,          # 1273 x (n_bins - 1) or NULL

  # Class-1 matrices, raw and log1p-transformed (three metrics)
  entropy_class1_matrix,
  hellinger_T1_class1_matrix,
  hellinger_Tpred_class1_matrix,
  entropy_class1_matrix_transformed,
  hellinger_T1_class1_matrix_transformed,
  hellinger_Tpred_class1_matrix_transformed,
  transformation,                       # "log1p" or "identity"

  # Cell-level FDA fits (three metrics)
  fda_entropy             = list(fit, n_clusters, silhouette_mean,
                                  silhouette_per_k, cluster_assignments,
                                  transformation, status, status_reason),
  fda_hellinger_T1        = list(...),
  fda_hellinger_Tpred     = list(...),

  # Bootstrap Jaccard stability (three metrics)
  stability_entropy        = list(n_bootstrap, per_cluster_jaccard,
                                   n_stable_clusters, n_failed_bootstrap),
  stability_hellinger_T1   = list(...),
  stability_hellinger_Tpred = list(...),

  # SNP Fisher enrichment (three metrics)
  snp_enrichment_entropy        = list(fisher_p, fisher_p_bonferroni,
                                        n_snp_in_class1, n_class1_total,
                                        snp_overlap_per_cluster,
                                        max_overlap_cluster, max_overlap_OR,
                                        contingency_table, status),
  snp_enrichment_hellinger_T1   = list(...),
  snp_enrichment_hellinger_Tpred = list(...),

  # Per-frame records
  frame_centres,                        # integer vector of centre bin indices
  frames_entropy          = list(...),  # list of per-frame records
  frames_hellinger_T1     = list(...),
  frames_hellinger_Tpred  = list(...),

  # Convergence summary: first centre_bin at which each predicate holds
  convergence_entropy            = list(first_snp, first_mut, first_class1),
  convergence_hellinger_T1       = list(first_snp, first_mut, first_class1),
  convergence_hellinger_Tpred    = list(first_snp, first_mut, first_class1),

  # Status and run metadata
  status,                               # "ok", "out_of_coverage",
                                        # "underpowered_bin", "reduced_skipped",
                                        # "too_few_sites", "error"
  status_reason,                        # character or NA
  walltime_s, seed_cell, seed_strategy,
  r_version, package_version, run_timestamp
)
```

Each element of `frames_<metric>` is itself a named list with shape
defined in `helpers_frames.R::run_one_frame()` and includes the
per-frame class-1 set, curve matrix (raw and transformed), cluster
assignments, per-site silhouette widths, oversized-cluster flags,
convergence flags, and the GMM status of the frame.

The cross-strategy ARI is computed by `aggregate_tables.R` from the
three same-`(dataset, variant, metric)` cell files; it is not stored
per cell.

# Supplementary tables (data dictionary)

`aggregate_tables.R` rolls the 72 per-cell RDS files into six
supplementary CSVs under `outputs/tables/`. The six tables differ
in granularity: some have one row per cell (`tab_01`, `tab_05`),
some one row per cell × cluster (`tab_03`), one per cell × site
(`tab_02`), one per `(dataset, variant, metric, strategy_pair)`
(`tab_04`), and one per cell × frame (`tab_06`). Row counts reported
below are from the production run on `tcell`.

The per-cell `cluster_membership_<metric>.csv` files written by
`render_existing.R` under `outputs/plots/<dataset>/<variant>/<strategy>/`
are a different artefact — they expose the per-site cluster
assignment as it is *rendered* in the GIF, including the per-frame
attribution. The aggregate `tab_02_cluster_membership.csv` is the
cell-level analogue.

## tab_01_internal_quality.csv

**Granularity:** one row per `(cell, metric)`. **Row count:** 216 =
72 cells × 3 metrics. **Purpose:** the headline cell-level quality
table — silhouette width, GMM model summary, chosen $k$, per-metric
status, and the three convergence-first-frame indices. This is the
table to consult first when interpreting a cell.

```{r tab01-dict}
tab01_dict <- data.frame(
  Column = c("dataset", "variant", "strategy", "metric",
             "cell_status", "cell_status_reason",
             "predecessor_name",
             "n_class1_sites",
             "gmm_G", "gmm_modelName",
             "n_bins", "n_frames",
             "k_selected", "silhouette_mean", "metric_status",
             "convergence_first_snp",
             "convergence_first_mut",
             "convergence_first_class1",
             "walltime_s"),
  Type = c("chr", "chr", "chr", "chr",
           "chr", "chr",
           "chr",
           "int",
           "int", "chr",
           "int", "int",
           "int", "num", "chr",
           "int", "int", "int",
           "num"),
  Description = c(
    "Dataset name: `NCBI_US` or `GISAID_US`.",
    "WHO variant label, one of the twelve variants in `sarscov2_variants`.",
    "Windowing strategy: `sliding_2m` (2-month bins, 1-month step), `disjoint_2m` (non-overlapping 2-month bins), or `cumulative_1m` (expanding bins, 1-month step).",
    "Curve type clustered. `entropy` = per-site Shannon-entropy curve (one entropy value per partition, in bits). `hellinger_T1` = per-site Hellinger trajectory anchored at the window-start partition `T1` (distance from each subsequent partition's residue distribution to `T1`'s). `hellinger_Tpred` = per-site Hellinger trajectory anchored at `T_pred`, the partition containing the predecessor variant's US detection date.",
    "Cell-level terminal status. `ok` = clustering ran end-to-end. `underpowered_bin` = at least one bin had fewer than `MIN_SEQUENCES_PER_BIN = 30L` sequences. `out_of_coverage` = the variant's US detection date is outside the dataset's coverage. `reduced_skipped` = the GMM fit was degenerate. `too_few_sites` = class-1 had fewer than `MIN_CLASS1_SITES_FOR_FDA = 4L` sites. `error` = an uncaught exception.",
    "Free-text reason matching the status (e.g., `\"2 / 16 bins have < 30 seqs (worst = 2).\"` for `underpowered_bin`); `NA` for `ok` cells.",
    "Variant whose US detection date anchors `hellinger_Tpred`. `D614G` is the canonical April-2020 reference used for Alpha and the early VOIs (Epsilon, Zeta, Eta, Iota); other variants chain through the lineage (Beta/Gamma/Theta/Kappa/Lambda → Alpha, Delta → Beta, Omicron → Delta).",
    "Number of class-1 (highest-entropy) Spike positions selected by the cell's window-level GMM; `NA` when the cell was rejected before the GMM step (`underpowered_bin`, `out_of_coverage`).",
    "Number of GMM components selected by BIC over `G` in 1, ..., 15; `NA` when the GMM did not run.",
    "`mclust` covariance model name selected by BIC (e.g., `V` = variable-variance univariate, `E` = equal-variance, `X` = single-component degenerate).",
    "Number of partitions the strategy produced for this cell (after window clipping). For an unclipped 20-month window: 19 for `sliding_2m`, 10 for `disjoint_2m`, 20 for `cumulative_1m`.",
    "Number of fitting frames computed for this metric. Each frame is a five-partition fitting sub-window centred on an interior partition.",
    "Number of clusters selected for this metric by mean silhouette over `k` in 2, ..., 6; `NA` when the metric was skipped (cell not `ok`, or class-1 too small).",
    "Mean silhouette width across all clustered sites at the chosen `k`. Ranges in -1 to 1; the Kaufman & Rousseeuw thresholds are 0.25 (reasonable), 0.50 (decisive), 0.70 (strong).",
    "Per-metric terminal status (`ok`, `skipped_too_few_sites`, `error`, ...). Distinct from `cell_status`: a cell can be `ok` while a particular metric (e.g., `hellinger_Tpred`) is skipped because its anchor falls outside the window.",
    "Index of the first frame centre at which all defining-SNP sites present in the frame's class-1 share a single, non-oversized cluster (Guard predicate `converged_snp`). `NA` if convergence never occurred within this metric's frames.",
    "Same as `convergence_first_snp` but for the variant's broader defining-mutation site set (`converged_mut`).",
    "Same as `convergence_first_snp` but for all well-clustered class-1 sites of the frame (`converged_class1`).",
    "Per-cell wall-clock seconds (the three metric rows for one cell repeat this value)."
  ),
  stringsAsFactors = FALSE
)
kbl(tab01_dict,
    caption = "Column dictionary for `tab_01_internal_quality.csv`.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## tab_02_cluster_membership.csv

**Granularity:** one row per `(cell, metric, site)`, restricted to
cells with `status == "ok"` and metrics whose `fda_*$status == "ok"`.
**Row count:** 5,058. **Purpose:** the bridge between the
clustering and the biology — for each clustered site, which cluster
it lives in and whether it is a defining-SNP or broader-mutation
site for the variant.

```{r tab02-dict}
tab02_dict <- data.frame(
  Column = c("dataset", "variant", "strategy", "metric",
             "cluster_id", "site",
             "is_snp", "is_mutation_site"),
  Type = c("chr", "chr", "chr", "chr",
           "int", "int",
           "lgl", "lgl"),
  Description = c(
    "Dataset name: `NCBI_US` or `GISAID_US`.",
    "WHO variant label.",
    "Windowing strategy (`sliding_2m`, `disjoint_2m`, or `cumulative_1m`).",
    "Curve type clustered. `entropy` = per-site Shannon-entropy curve over time (bits). `hellinger_T1` = per-site Hellinger trajectory from each subsequent partition's residue distribution to the window-start partition `T1`'s. `hellinger_Tpred` = same, but anchored at `T_pred` (the partition containing the predecessor variant's US detection date).",
    "Cluster ID assigned by `fdacluster::fdahclust` (1-based). The chosen `k` (number of clusters) is in `tab_01_internal_quality.csv$k_selected`; valid IDs run from 1 through that `k`.",
    "Spike position (1 to 1,273) belonging to the cluster.",
    "`TRUE` if `site` is in the variant's defining-SNP set (`sarscov2_variants$Defining_SNP_Sites[[variant]]`).",
    "`TRUE` if `site` is in the variant's broader defining-mutation set (`sarscov2_variants$Mutation_Sites[[variant]]`)."
  ),
  stringsAsFactors = FALSE
)
kbl(tab02_dict,
    caption = "Column dictionary for `tab_02_cluster_membership.csv`.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## tab_03_stability.csv

**Granularity:** one row per `(cell, metric, cluster_id)`, restricted
to cells with `status == "ok"` and metrics whose bootstrap procedure
ran. **Row count:** 469. **Purpose:** per-cluster Hennig (2007)
bootstrap Jaccard stability. The `is_stable` flag uses
`STABLE_JACCARD_THRESHOLD = 0.75`; rerunning aggregation with a
different threshold (e.g., 0.60) would re-classify clusters.

```{r tab03-dict}
tab03_dict <- data.frame(
  Column = c("dataset", "variant", "strategy", "metric",
             "cluster_id", "mean_jaccard", "is_stable",
             "n_bootstrap", "n_failed_bootstrap"),
  Type = c("chr", "chr", "chr", "chr",
           "int", "num", "lgl",
           "int", "int"),
  Description = c(
    "Dataset name: `NCBI_US` or `GISAID_US`.",
    "WHO variant label.",
    "Windowing strategy (`sliding_2m`, `disjoint_2m`, or `cumulative_1m`).",
    "Curve type clustered (see `tab_01$metric` for definitions).",
    "Cluster ID (1-based, identifies the same cluster as in `tab_02_cluster_membership.csv`).",
    "Hennig (2007) bootstrap Jaccard stability: the mean over `B = 100` site-resampling bootstrap rounds of the maximum Jaccard similarity between this original cluster's site set and any cluster of the resampled fit. Ranges in 0 to 1; closer to 1 = more stable.",
    "`TRUE` iff `mean_jaccard >= STABLE_JACCARD_THRESHOLD = 0.75` (Hennig's \"stable\" tier). The 0.60-0.75 range is \"patterned\"; below 0.60 is unstable.",
    "Number of bootstrap rounds attempted (`N_BOOTSTRAP = 100L`).",
    "Number of bootstrap rounds in which the recompute failed (e.g., the resampled site set was too small for `fdahclust` or yielded a degenerate fit)."
  ),
  stringsAsFactors = FALSE
)
kbl(tab03_dict,
    caption = "Column dictionary for `tab_03_stability.csv`.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## tab_04_cross_strategy_ari.csv

**Granularity:** one row per `(dataset, variant, metric,
strategy_pair)`, restricted to `(dataset, variant)` groups where
both strategies of the pair returned `status == "ok"` and the
common class-1 site set has at least 3 elements. **Row count:** 141.
**Purpose:** Adjusted Rand Index between the two cluster
assignments restricted to their common class-1 sites, quantifying
methodological concordance (Q2). The three pairs are
`(sliding_2m, disjoint_2m)`, `(sliding_2m, cumulative_1m)`, and
`(disjoint_2m, cumulative_1m)`.

```{r tab04-dict}
tab04_dict <- data.frame(
  Column = c("dataset", "variant", "metric",
             "strategy_a", "strategy_b",
             "n_sites_common", "k_a", "k_b", "ARI"),
  Type = c("chr", "chr", "chr",
           "chr", "chr",
           "int", "int", "int", "num"),
  Description = c(
    "Dataset name: `NCBI_US` or `GISAID_US`.",
    "WHO variant label.",
    "Curve type clustered (see `tab_01$metric` for definitions).",
    "Name of the first strategy in the pair (`sliding_2m`, `disjoint_2m`, or `cumulative_1m`).",
    "Name of the second strategy in the pair.",
    "Number of class-1 sites present in both strategies (size of the intersection of the two class-1 site sets). Rows with fewer than 3 common sites are dropped because ARI on so few items is uninformative.",
    "Number of clusters selected for `strategy_a` (matches `tab_01_internal_quality.csv$k_selected` for that `(dataset, variant, strategy_a, metric)` row).",
    "Number of clusters selected for `strategy_b`.",
    "Adjusted Rand Index [`mclust::adjustedRandIndex`] between the two cluster assignments restricted to the `n_sites_common` common sites. `1` = perfect agreement; `0` = chance agreement; negative = worse than chance."
  ),
  stringsAsFactors = FALSE
)
kbl(tab04_dict,
    caption = "Column dictionary for `tab_04_cross_strategy_ari.csv`.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## tab_05_snp_enrichment.csv

**Granularity:** one row per `(cell, metric)`. **Row count:** 216.
**Purpose:** Fisher exact test of whether the variant's defining-
SNP sites cluster together more than would be expected by chance
(Q4). The Bonferroni correction uses
`N_TOTAL_ENRICHMENT_TESTS = 220` (the design ceiling).

```{r tab05-dict}
tab05_dict <- data.frame(
  Column = c("dataset", "variant", "strategy", "metric",
             "cell_status", "enrichment_status",
             "n_snp_in_class1", "n_class1_total", "n_snp_total_truth",
             "fisher_p", "fisher_p_bonferroni",
             "max_overlap_cluster", "max_overlap_OR"),
  Type = c("chr", "chr", "chr", "chr",
           "chr", "chr",
           "int", "int", "int",
           "num", "num",
           "int", "num"),
  Description = c(
    "Dataset name: `NCBI_US` or `GISAID_US`.",
    "WHO variant label.",
    "Windowing strategy (`sliding_2m`, `disjoint_2m`, or `cumulative_1m`).",
    "Curve type clustered (see `tab_01$metric` for definitions).",
    "Cell-level terminal status (same domain as `tab_01$cell_status`).",
    "Per-metric enrichment status: `ok` (test ran), `no_snp_in_class1` (no defining-SNP site landed in the class-1 set), `fisher_failed` (degenerate contingency table), `skipped` (cell not `ok` or metric not `ok`), ...",
    "Count of defining-SNP sites that appear in this cell's class-1 site set (subset of the variant's `Defining_SNP_Sites`).",
    "Total number of sites in the cell's class-1 site set (matches `tab_01$n_class1_sites`).",
    "Total number of defining-SNP sites for the variant in the ground-truth catalogue (`length(sarscov2_variants$Defining_SNP_Sites[[variant]])`).",
    "One-tailed Fisher exact `p`-value for the most-enriched cluster, testing the null that defining-SNP sites are distributed across clusters in proportion to cluster size.",
    "Bonferroni-adjusted `p`: `min(1, fisher_p * N_TOTAL_ENRICHMENT_TESTS)` with `N_TOTAL_ENRICHMENT_TESTS = 220` (the design ceiling on the number of enrichment tests across the study).",
    "Cluster ID with the largest defining-SNP overlap; `NA` if no overlap.",
    "Odds ratio for the most-enriched cluster's 2x2 contingency table (`Inf` when one off-diagonal cell is zero, i.e., perfect or near-perfect overlap)."
  ),
  stringsAsFactors = FALSE
)
kbl(tab05_dict,
    caption = "Column dictionary for `tab_05_snp_enrichment.csv`.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## tab_06_per_frame.csv

**Granularity:** one row per `(cell, metric, frame)`, restricted to
cells with `status == "ok"`. **Row count:** 1,764. **Purpose:** the
per-frame diagnostics that drive the animations — the table to
consult when interpreting whether a specific frame in a GIF crossed
the convergence threshold or was rendered null by one of the three
quality guards.

```{r tab06-dict}
tab06_dict <- data.frame(
  Column = c("dataset", "variant", "strategy", "metric",
             "frame_idx", "centre_bin",
             "fitting_start_bin", "fitting_end_bin",
             "gmm_status", "gmm_G", "gmm_n_sequences",
             "fit_status", "n_class1_sites", "n_clusters",
             "silhouette_mean", "structurally_null",
             "n_snp_in_class1", "n_mut_in_class1",
             "converged_snp", "converged_mut", "converged_class1",
             "qualifying_cluster_snp", "qualifying_cluster_mut",
             "qualifying_cluster_class1",
             "anchor_idx", "n_oversized_clusters"),
  Type = c("chr", "chr", "chr", "chr",
           "int", "int", "int", "int",
           "chr", "int", "int",
           "chr", "int", "int",
           "num", "lgl",
           "int", "int",
           "lgl", "lgl", "lgl",
           "int", "int", "int",
           "int", "int"),
  Description = c(
    "Dataset name: `NCBI_US` or `GISAID_US`.",
    "WHO variant label.",
    "Windowing strategy (`sliding_2m`, `disjoint_2m`, or `cumulative_1m`).",
    "Curve type clustered (see `tab_01$metric` for definitions).",
    "Sequential frame index (1-based) within this `(cell, metric)`. Frames are emitted for every interior partition with at least `FITTING_WINDOW_HALFWIDTH = 2` partitions on either side.",
    "Index of the partition (bin) at the centre of this frame's fitting sub-window. Used as the frame's identifier in the GIF and as the time-axis position of the convergence indices in `tab_01`.",
    "Index of the first partition in the frame's 5-partition fitting sub-window (`centre_bin - 2`).",
    "Index of the last partition in the frame's 5-partition fitting sub-window (`centre_bin + 2`).",
    "Status of the per-frame GMM refit: `ok`, `G_eq_1` (Guard A fails: GMM selects a single component, so no class-1 set exists), `failed` (numerical failure in `mclust::Mclust`), or `skipped`.",
    "Number of GMM components selected by BIC over `G` in 1, ..., 15 for the frame's pooled-sequence entropy.",
    "Number of sequences pooled across the frame's 5 partitions (the basis for the per-frame entropy and GMM).",
    "Status of the per-frame `fdacluster::fdahclust` refit (`ok`, `skipped_too_few_sites`, `error`, ...).",
    "Number of class-1 sites in this frame's per-frame GMM (not the cell-level class-1 set; per-frame GMMs are independent of the cell-level fit).",
    "Number of clusters selected for this frame's `fdahclust` fit by silhouette over `k` in 2, ..., 6.",
    "Mean silhouette width at the frame's chosen `k`.",
    "Guard B flag. `TRUE` iff `silhouette_mean < MIN_MEAN_SILHOUETTE = 0.25` (Kaufman & Rousseeuw 1990 \"reasonable structure\" floor). Frames with `structurally_null = TRUE` cannot satisfy any convergence predicate regardless of cluster pattern.",
    "Number of defining-SNP sites that are in this frame's class-1 site set.",
    "Number of defining-mutation sites that are in this frame's class-1 site set.",
    "`TRUE` iff all SNP sites in this frame's class-1 share a single cluster whose size does not exceed `MAX_CLUSTER_SIZE = 33` (i.e., not an \"oversized\" cluster).",
    "`TRUE` iff the same holds for the variant's broader defining-mutation sites.",
    "`TRUE` iff all class-1 sites with positive per-site silhouette (Guard C: well-clustered sites) share a single non-oversized cluster.",
    "Cluster ID into which the SNP sites converged for this frame; `NA` if `converged_snp = FALSE`.",
    "Cluster ID into which the mutation sites converged; `NA` if `converged_mut = FALSE`.",
    "Cluster ID into which the class-1 sites converged; `NA` if `converged_class1 = FALSE`.",
    "Hellinger anchor partition index used to compute this frame's curves (`bin_idx` of `T1` or `T_pred`). `NA` for the `entropy` metric, which has no anchor.",
    "Number of clusters in this frame whose size exceeds `MAX_CLUSTER_SIZE = 33` (Omicron's defining-mutation count, the largest among the 12 WHO variants). Oversized clusters are rendered light grey in the GIF panels and are excluded from convergence predicates."
  ),
  stringsAsFactors = FALSE
)
kbl(tab06_dict,
    caption = "Column dictionary for `tab_06_per_frame.csv`.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

# Limitations

1. **17 underpowered cells.** Cells in which any bin contained fewer
   than `MIN_SEQUENCES_PER_BIN = 30L` sequences are excluded from
   clustering. This concentrates on the four low-prevalence VOIs
   (Epsilon, Eta, Iota, Zeta), where a single early-pandemic bin
   regularly contained 2–8 sequences despite the variant's overall
   window count exceeding 20,000. One Alpha × cumulative_1m ×
   GISAID US cell also crossed the threshold (one of 20 bins had
   6 sequences). Lowering `MIN_SEQUENCES_PER_BIN` would recover
   these cells at the cost of less reliable per-bin frequency
   estimates; a sensitivity sweep across thresholds is a natural
   follow-up.
2. **3 out-of-coverage cells.** Omicron × NCBI US is structurally
   excluded across all three strategies because NCBI US coverage
   ends in September 2021 and Omicron was first detected in the US
   in December 2021. Mitigated by the parallel run on GISAID US,
   which covers Omicron in all three strategies.
3. **US-only, Spike-only.** Other geographies and other SARS-CoV-2
   proteins are not analysed. Generalisation outside Spike is not
   tested.
4. **Window-length fixed at $\pm 10$ months.** The choice
   $K = 5$ bins at $b_{\max} = 2$ months is informed by the earlier
   "5 points before, 5 after" convention but is not itself a
   parameter sweep. A sensitivity sweep on window length is a
   natural follow-up.
5. **Three-strategy panel non-exhaustive.** Other discretisations
   exist (3-month disjoint, 1-month sliding) and were considered but
   excluded to keep the cell count tractable.
6. **No explicit "true" cluster partition.** Functional clustering
   has no ground truth in the strict sense of change-point detection
   (where the truth set is the WHO catalogue's detection dates).
   The SNP enrichment test (Q4) is a biological proxy — a
   *plausibility* check, not a *validation* of cluster correctness.
   Captions and discussion phrase this distinction explicitly.
7. **Hellinger reference is dual-anchored, not swept.** The study
   anchors the Hellinger trajectories at the window-start partition
   $T_1$ *and* at the partition containing the predecessor
   variant's US detection ($T_{\text{pred}}$). The map of
   predecessors is curated (Alpha → D614G; Beta / Gamma / Theta /
   Kappa / Lambda → Alpha; Delta → Beta; Omicron → Delta; Epsilon /
   Zeta / Eta / Iota → D614G). Alternative ancestry mappings (e.g.,
   by Pango parent rather than by WHO label) are not swept and would
   be a natural follow-up.
8. **Class-1 site reduction is a sequential bottleneck.** The full-
   window cell-level fits depend on the GMM having identified the
   correct high-entropy component (class 1). A small high-entropy
   class 1 may exclude positions whose entropy is just below the
   class boundary, and a misidentified class 1 propagates to all
   downstream metrics. The per-frame design partly mitigates this
   by re-fitting the GMM in every fitting sub-window, so a class-1
   misidentification at a single moment in time does not corrupt
   the neighbouring frames. A formal sensitivity analysis that
   takes the union of class-1 and class-2 sites is a supplementary
   check reserved for follow-up work.
9. **`fdahclust` warping disabled.** Curve alignment (DTW-style
   warping) is not used. The justification is methodological: the
   x-axis is bin index, *which at the population scale used here
   is calendar time*. Warping the calendar would correspond to
   claiming that the biological clock of the virus ran faster or
   slower in some bins than in others — a claim with no clear
   evidentiary basis at the level of national surveillance pools
   used in this study.
10. **Fisher exact assumes site independence under linkage
    disequilibrium.** The Fisher exact SNP enrichment test treats
    cluster assignment of each site as independent of every other
    site, which is violated by linkage disequilibrium across Spike
    positions co-segregating in real lineages. The reported
    $p$-values should therefore be read as descriptive concentration
    statistics rather than as strict null-hypothesis tests; a
    permutation-style sensitivity analysis that resamples cluster
    labels while preserving the SNP-site count is planned as a
    supplementary check.
11. **Bin-width robustness untested.** Bin widths are fixed by
    strategy. A sensitivity sweep on bin width within a strategy is
    a natural follow-up (mirroring the planned bin-width sweep on
    `Changepoint_Detection_Study`).
12. **GIF frame rate fixed at 1 fps.** Visual presentation of
    animations is a deliverable, not a tuning target; the frame
    rate was chosen to be slow enough that the eye can track
    cluster re-organisation across frames.
13. **N_WORKERS sensitivity to memory pressure.** The first-pass
    run at `N_WORKERS = 12` on the 125 GB shared server lost five
    cells to the kernel OOM-killer (Omicron × cumulative_1m,
    Alpha × cumulative_1m, Beta / Kappa / Lambda × sliding_2m).
    These were recovered cleanly at `N_WORKERS = 5`. The Hellinger
    pairwise computation for the heaviest cells (longest cumulative
    strategies on GISAID US) is the dominant allocation; users on
    smaller hosts should use a conservative `N_WORKERS` setting.

# ADEMP summary

```{r ademp-table}
ademp <- data.frame(
  Component = c("Aim",
                "Data-generating mechanism",
                "Estimand",
                "Methods",
                "Performance measures"),
  Description = c(
    "Cluster per-site Shannon entropy curves and pairwise Hellinger trajectories (dual-anchored at the window start T1 and at the partition containing the predecessor variant's US detection Tpred) from real SARS-CoV-2 Spike surveillance data, around the US first-detection dates of twelve WHO-labelled variants, under three temporal-partitioning strategies, and quantify the internal quality, resampling stability, cross-strategy concordance, and biological plausibility of the resulting clusters at both the cell level (full-window fit) and per frame (five-partition sliding fits).",
    "Per-variant 20-month windows of two real surveillance datasets (NCBI US and GISAID US), partitioned by three strategies, with GMM-based class-1 site selection per window (cell-level) and per five-partition fitting slice (per-frame), and no oracle leakage across cells.",
    "Per-(variant, dataset, strategy, metric) cluster partition of class-1 Spike sites at the cell level and at every interior fitting frame, characterised by the chosen number of clusters, silhouette width, per-cluster bootstrap Jaccard stability, cross-strategy Adjusted Rand Index, Fisher exact SNP enrichment p-value, and the first-frame convergence indices for the three convergence predicates (SNP / Mutation / class-1).",
    "fdacluster::fdahclust under warping_class = 'none', L2 metric, centroid_type = 'mean', linkage = 'complete', k chosen by silhouette over k in {2, ..., 6}, log1p input transformation; mclust::Mclust over G in {1, ..., 15} for class-1 site selection at both cell and frame levels.",
    "Average silhouette width (internal quality), bootstrap Jaccard stability with B = 100 (robustness), cross-strategy ARI (methodological concordance), Fisher exact SNP enrichment p-value with Bonferroni correction across N_TOTAL_ENRICHMENT_TESTS = 220 (biological plausibility), and the first-frame convergence indices per metric (longitudinal interpretability)."
  ),
  check.names = FALSE
)
kbl(ademp,
    caption = "ADEMP framing [@Morris2019]. Aim, Data-generating mechanism, Estimand, Methods, Performance measures.") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

# Invocation

```bash
# Working directory
cd analysis/FDA_Analysis

# Optional: pre-flight parse check (no model evaluation).
Rscript -e 'for (f in c("setup.R","helpers_fda.R","helpers_frames.R",
                         "helpers_panels.R","run_one_cell.R","fda_analysis.R",
                         "plot_results.R","animate_results.R",
                         "aggregate_tables.R","render_existing.R"))
              parse(file=f); cat("parse OK\n")'

# Phase 1: cell production (orchestrator). Recommended on a 64+ GB host:
TS=$(date +%Y%m%d_%H%M%S)
LOG="outputs/logs/orchestrator_${TS}.log"
N_WORKERS=8 nohup Rscript fda_analysis.R > "$LOG" 2>&1 &
echo $! > outputs/logs/orchestrator.pid
disown

# Phase 2-5 (after Phase 1 completes):
Rscript aggregate_tables.R   2>&1 | tee outputs/logs/aggregate_${TS}.log
Rscript plot_results.R       2>&1 | tee outputs/logs/plot_${TS}.log

# render_existing.R writes per-cell cluster_membership_*.csv files
# (parallelisable with xargs -P; see analysis/FDA_Analysis/render_existing.R).
Rscript -e '
  cells <- list.files("outputs/cells", pattern = "\\.rds$", full.names = TRUE)
  rows <- lapply(cells, function(p) {
    r <- readRDS(p)
    if (!is.null(r$status) && r$status == "ok")
      sprintf("%s %s %s", r$dataset, r$variant, r$strategy)
    else NULL
  })
  writeLines(unlist(rows), "ok_cells.txt")'
cat ok_cells.txt | xargs -P 8 -L 1 bash -c '
  SMOKE_DATASET=$0 SMOKE_VARIANT=$1 SMOKE_STRATEGY=$2 \
    Rscript render_existing.R \
    > "outputs/logs/render_${0}__${1}__${2}.log" 2>&1
  echo "[$(date +%H:%M:%S)] $0 $1 $2 exit=$?"'

nohup Rscript animate_results.R > outputs/logs/animate_${TS}.log 2>&1 &
```

Single-cell debug:

```r
setwd("analysis/FDA_Analysis")
source("setup.R")
source("helpers_fda.R")
source("helpers_frames.R")
source("run_one_cell.R")

config <- build_config()
truth  <- load_truth_catalogue()
fm     <- load_feature_matrix(config$DATASETS$GISAID_US$feature_rds,
                                expected_n_sites = config$N_SITES)
result <- run_one_cell(
  dataset_name  = "GISAID_US",
  variant_name  = "Delta",
  strategy_name = "disjoint_2m",
  feature_matrix = fm,
  truth_df       = truth,
  config         = config
)
str(result, max.level = 2L)
```

# Project structure

```
analysis/FDA_Analysis/
|-- README.Rmd                          (this file)
|-- README.md                           (knitted GitHub output)
|-- README.html                         (knitted HTML output)
|-- references_fda.bib                  (BibTeX source)
|-- captions_and_discussion.txt         (figure captions + main-text discussion)
|
|-- setup.R                             (config, helpers, paths, seeds, predecessor map)
|-- helpers_fda.R                       (per-cell window slicing, GMM, entropy, Hellinger, fdahclust)
|-- helpers_frames.R                    (per-frame GMM, fdahclust, silhouette, convergence)
|-- helpers_panels.R                    (dual-panel rendering for static plots and GIF frames)
|-- run_one_cell.R                      (per-cell driver, including per-frame loop)
|-- fda_analysis.R                      (orchestrator)
|
|-- plot_results.R                      (per-frame PNGs + per-cell summary PNGs)
|-- animate_results.R                   (per-cell × per-metric GIF animations)
|-- aggregate_tables.R                  (6 supplementary CSV tables)
|-- render_existing.R                   (per-cell cluster_membership CSV side-output)
|-- smoke_test.R                        (single-cell smoke test)
|
`-- outputs/
    |-- cells/cell_<dataset>__<variant>__<strategy>.rds   (72 per-cell result files)
    |-- logs/
    |   |-- orchestrator_<timestamp>.log
    |   |-- aggregate_<timestamp>.log
    |   |-- plot_<timestamp>.log
    |   |-- animate_<timestamp>.log
    |   |-- render_*.log                  (per-cell render_existing.R logs)
    |   `-- error_log.txt                 (per-cell errors, normally empty)
    |
    |-- plots/<dataset>/<variant>/<strategy>/
    |   |-- frame_entropy_<centre:03d>.png            (one per frame)
    |   |-- frame_hellinger_T1_<centre:03d>.png       (one per frame)
    |   |-- frame_hellinger_Tpred_<centre:03d>.png    (one per frame)
    |   |-- summary_entropy.png                       (full-window cell fit)
    |   |-- summary_hellinger_T1.png
    |   |-- summary_hellinger_Tpred.png
    |   |-- frames.manifest.csv                       (frame inventory per cell)
    |   |-- cluster_membership_entropy.csv            (rendered cluster attribution per site)
    |   |-- cluster_membership_hellinger_T1.csv
    |   `-- cluster_membership_hellinger_Tpred.csv
    |
    |-- gifs/<dataset>/<variant>/<strategy>/
    |   |-- anim_entropy.gif
    |   |-- anim_hellinger_T1.gif
    |   `-- anim_hellinger_Tpred.gif
    |
    `-- tables/
        |-- tab_01_internal_quality.csv     (216 rows; per cell × metric)
        |-- tab_02_cluster_membership.csv   (5,058 rows; per cell × metric × site)
        |-- tab_03_stability.csv            (469 rows; per cell × metric × cluster)
        |-- tab_04_cross_strategy_ari.csv   (141 rows; per dataset × variant × metric × strategy pair)
        |-- tab_05_snp_enrichment.csv       (216 rows; per cell × metric)
        `-- tab_06_per_frame.csv            (1,764 rows; per cell × metric × frame)
```

Each per-cell RDS holds the cell identifiers, window metadata, GMM
fit summary, the entropy and Hellinger matrices on the class-1 site
set (both raw and log1p-transformed, both anchors), the three
`fdacluster::fdahclust` fits with their chosen number of clusters
and silhouette values, the three per-cluster bootstrap Jaccard
stability vectors, the three per-cluster SNP overlap counts and
Fisher test statistics, the per-frame records for all three metrics,
the three convergence summaries, run metadata (status, wall time,
seeds), and R / `ViralEntropR` version strings. The six
supplementary CSVs are computed by `aggregate_tables.R` from these
per-cell files.

# Session information

```{r session, eval=TRUE, include=TRUE}
sessionInfo()
```

# References
