---
title: "Entropy Clustering, Hellinger Distance, and Change Point Analysis for Emerging Viral Variant Detection: A Simulation Study"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    theme: flatly
    highlight: tango
bibliography: references_detection.bib
nocite: '@*'
vignette: >
  %\VignetteIndexEntry{Entropy Clustering, Hellinger Distance, and Change Point Analysis for Emerging Viral Variant Detection: A Simulation Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

# Introduction

This vignette provides a comprehensive, reproducible, step-by-step walkthrough
of the `ViralEntropR` pipeline:

- Simulate a realistic viral evolutionary timeline with four competing variants
  (2020–2021).
- Partition the data into non-overlapping 2-month time windows.
- Calculate Hellinger distance matrices to quantify distributional shifts.
- Detect evolutionary change points using three complementary ECP approaches.

For each partition, we calculate site-wise Shannon entropies, cluster these
entropy profiles using a Gaussian Mixture Model (GMM), and identify which sites
fall into the highest-entropy class.  We then compute Hellinger distances
between partitions using the first partition as the reference.  If there are 12
partitions there are 11 comparisons; if a variant emerges in partition 3, the
correct change point is index 2 (difference between partition 3 and partition 1).

# Setup

```{r}
library(ViralEntropR)
library(dplyr)
library(lubridate)
library(ecp)
library(knitr)
library(kableExtra)
```

# Reference Sequence

```{r}
ref_sequence <- "MKTIIALSYIFCLVFADYKDDDDK"
sprintf("There are %d sites in the reference sequence", nchar(ref_sequence))
```

# Simulation via `simulate_variant_evolution()` 

| **Parameter**               | **Description**                                                                  | **Default** |
|-----------------------------|----------------------------------------------------------------------------------|-------------|
| `ref_sequences`             | data.frame which must include a `Date` column; OR a single sequence string.      | —           |
| `n_ref_months`              | number of months for the reference phase                                         | 3           |
| `start_date`                | simulation start (character or Date)                                             | —           |
| `end_date`                  | simulation end (character or Date)                                               | —           |
| `variants_config`           | integer vector, e.g., `c(2,6)`: Variant 1 gets 2 mutations; Variant 2 gets 6.   | —           |
| `variant_intervals`         | gap (in months) between the emergence of consecutive variants                    | —           |
| `n_new_mutations`           | de-novo sequences produced at a variant's first appearance                       | 1           |
| `mutation_rate`             | monthly growth multiplier per variant (scalar or per-variant vector)             | 2           |
| `mutation_rate_variability` | fractional spread ± around `mutation_rate`; 0 = deterministic growth            | 0.25        |
| `deleterious_rate`          | probability that any new mutation is deleterious                                 | 0           |
| `n_deleterious_limit`       | maximum sequences allowed to carry a deleterious mutation                        | 1           |
| `n_sequences_total`         | total sequences generated over the mutation phase (spread evenly per month)      | 140         |
| `ref_variability`           | if `TRUE`, introduce variability at sites 1 and 2 of the reference pool          | FALSE       |
| `n_ref_sequences`           | when `ref_sequences` is a single string, number of reference rows to create      | 100         |
| `prob_deletion_event`       | per-month probability of injecting a block of sequences with a random deletion   | 0           |
| `n_rows_to_delete`          | sequences affected when `prob_deletion_event` fires                              | 0           |
| `seed`                      | optional random seed for reproducibility                                         | NULL        |

## Simulate Data

In this simulation, `simulate_variant_evolution()` models the stepwise
emergence and competition of four viral variants over a two-year period from
`2020-01-01` through `2021-12-01`.  We begin with a 4-month reference phase
(`n_ref_months = 4`) in which 200 total reference sequences are sampled
(50 per month, `n_ref_sequences = 200`), and variability is introduced at
sites 1 and 2 (`ref_variability = TRUE`).  The mutation phase then generates
1,000 sequences (`n_sequences_total = 1000`) at 50 per month.  Four variants
emerge at months 5, 9, 14 and 19, carrying 2, 3, 4 and 2 mutations
respectively (`variants_config = c(2,3,4,2)`), with 2 de-novo sequences at
each emergence (`n_new_mutations = 2`) and growth multipliers of 1.25 per
variant (`mutation_rate`).  Although `deleterious_rate = 0` for
variant-originated deleterious mutations, each month has a 20% chance
(`prob_deletion_event = 0.2`) of injecting a block of 3 sequences with a
random deletion at one site (`n_rows_to_delete = 3`).  Finally, `seed = 2025`
ensures reproducibility.  The returned list includes `Simulation_Output` (with
per-sequence `Delet` flags), `Variant_Details`, `Simulation_Dates`, and
`Delet_Records` for downstream analysis.

Variants compete in a **pairwise** fashion, always pitting the newest active
variant against its immediate predecessor (or the original reference when only
one variant has emerged).  During months before any variant appears, all
sequences are drawn from the reference pool (the wild-type row plus up to two
reference-variability rows when `ref_variability = TRUE`).  Once Variant 1
emerges, each month's batch is composed of its de-novo copies plus sampled
reference sequences; when Variant 2 appears it competes directly with Variant
1, and so on — Variant 3 competes against Variant 2, Variant 4 against Variant
3.  The **fill pool** (sequences that pad the monthly quota after the two
competing variants are allocated) is composed of the reference rows plus any
older non-competing variants.  Sampling weights are assigned so that all
reference rows carry weight 1, while Variant *i* rows carry weight `2^(i+1)`,
ensuring that more recently emerged variants have higher selection probability
in subsequent months.  Growth is **stochastic**: each variant's monthly count
is drawn from
`Uniform(mult*(1 - variability), mult*(1 + variability))`
so that `mutation_rate_variability > 0` introduces realistic fluctuation.

```{r}
sim_result <- simulate_variant_evolution(
  ref_sequences             = ref_sequence,            # single sequence string
  n_ref_months              = 4,                       # 4-month reference phase
  start_date                = "2020-01-01",
  end_date                  = "2021-12-01",
  variants_config           = c(2, 3, 4, 2),           # 4 variants: 2,3,4,2 mutations
  variant_intervals         = c(4, 5, 5),              # V2 at month 9; V3 at 14; V4 at 19
  n_new_mutations           = 2,                       # 2 de-novo sequences at emergence
  mutation_rate             = c(1.25, 1.25, 1.25, 1.25), # per-variant growth multipliers
  mutation_rate_variability = 0,                       # deterministic growth (no spread)
  deleterious_rate          = 0,                       # no within-variant deleterious mutations
  n_deleterious_limit       = 0,
  n_sequences_total         = 1000,                    # 1000/(24-4) = 50 per period
  ref_variability           = TRUE,                    # variability at sites 1 and 2
  n_ref_sequences           = 200,                     # 200/4 = 50 reference rows per month
  prob_deletion_event       = 0.2,                     # 20% per-month deletion probability
  n_rows_to_delete          = 3,                       # 3 sequences affected if event fires
  seed                      = 2025
)
```

## Check the Pool

The pool returned is the fill pool from the last multi-variant competition
period.  Column 25 = Variant, column 26 = Phase, column 30 = sampling weight.

```{r}
sim_result$Pool[, c(25, 26, 30)]
```

## Emergence Periods and Sites

```{r}
print("Variant emergence period:")
sim_result$Variant_Details[[1]]$em
```

```{r}
print("Variant 1 sites:")
sim_result$Variant_Details[[1]]$pos %>% sort()
```

```{r}
# Month 9 translates into partition 5.
print("Variant 2 emergence period:")
sim_result$Variant_Details[[2]]$em
```

```{r}
print("Variant 2 sites:")
sim_result$Variant_Details[[2]]$pos %>% sort()
```

```{r}
# Month 14 translates into partition 7.
print("Variant 3 emergence period:")
sim_result$Variant_Details[[3]]$em
```

```{r}
print("Variant 3 sites:")
sim_result$Variant_Details[[3]]$pos %>% sort()
```

```{r}
# Month 19 translates into partition 10.
print("Variant 4 emergence period:")
sim_result$Variant_Details[[4]]$em
```

```{r}
print("Variant 4 sites:")
sim_result$Variant_Details[[4]]$pos %>% sort()
```

## Deleterious Periods and Sites

```{r}
sprintf("The number of deleterious periods is %d",
        length(sim_result$Delet_Records))

for (i in seq_along(sim_result$Delet_Records)) {
  cat(sprintf("Deleterious Period %d: month %d, site %d\n",
              i,
              sim_result$Delet_Records[[i]]$period,
              sim_result$Delet_Records[[i]]$site))
}
```

## Inspecting the Simulation Output

The sections below walk through selected time points to verify that the
simulation is behaving as expected. Each month's batch contains 50 sequences
(reference + variant). Key columns to watch are the **mutated sites** (which
will show a non-reference amino acid for variant sequences), **Variant**
(which strain each row belongs to), **Phase** (`R` = reference wild-type,
`RV` = reference with added site-1/2 variability, `M1`/`M2`... = mutation
phase), and **Delet** (flags sequences affected by a random deletion event).

## Reference Phase 1 (month 1)

During the reference phase only wild-type sequences are present. Phase `RV`
denotes reference sequences with low-level variability artificially introduced
at sites 1 and 2 (`ref_variability = TRUE`) to mimic the background noise
typical of real surveillance data.

```{r}
sim_result$Simulation_Output[1:10, ]
```

## Reference Phase 4 (month 4)

The final reference month before any variant emerges. All sequences should be
`Reference` strain, with Phase `R` or `RV`. This establishes the baseline
amino acid distribution against which downstream Hellinger distances will be
measured.

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 4, ][1:10, ]
```

## `Variant 1` Phase 1 (month 5): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

Variant 1 emerges with 2 de-novo sequences (`n_new_mutations = 2`). The
mutated sites carry a different amino acid to the reference. The remaining ~48
sequences are sampled from the reference pool. Notice that the variant
sequences are a very small minority at this point — this is the detection
challenge: identifying emergence signal against a large reference background.
`Reference` rows can be either `R` or `RV`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 5,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 1` Phase 2 (month 6): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

After one month of growth at rate 1.25, the number of Variant 1 sequences
increases to `r ceiling(2 * 1.25)`. The variant is still a small fraction of
the population, but its mutated sites are accumulating signal in the entropy
profile.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 6,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 1` Phase 4 (month 8): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

After three months of compounding growth (`ceiling(2 * 1.25^3)`), Variant 1
now contributes `r ceiling(ceiling(ceiling(2 * 1.25) * 1.25) * 1.25)` sequences.
Its mutated sites are now consistently different from the reference, and site
entropy at those positions is rising — setting the stage for detection in the
next partition.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 8,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 2` Phase 1 (month 9): sites `r sim_result$Variant_Details[[2]]$pos %>% sort()`

Variant 2 emerges with 2 de-novo sequences. Variants 1 and 2 are now in direct
pairwise competition. The remainder of the 50 sequences is sampled from the
fill pool (reference rows only at this stage, since there are no older
non-competing variants yet). Note that a random deletion event also fires this
month at site `r sim_result$Delet_Records[[1]]$site` — those affected rows
have `Delet = "Yes"` and carry a randomly chosen amino acid substitution at
that site.

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 9, ]
```

## `Variant 2` Phase 5 (month 13): sites `r sim_result$Variant_Details[[2]]$pos %>% sort()`

By month 13, both Variant 1 and Variant 2 have been growing for several months
and together dominate the monthly batch. The reference pool contributes only
the remaining fill sequences. The population is now clearly multi-strain, with
entropy elevated at the mutation sites of both variants.

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 13, ]
```

## `Variant 3` Phase 1 (month 14): sites `r sim_result$Variant_Details[[3]]$pos %>% sort()`

Variant 3 emerges and begins competing against Variant 2. Variant 1 is now
relegated to the fill pool with an exponentially increasing sampling weight
(`2^(1+1) = 4`), meaning it still appears but at lower frequency than the
two active competitors.

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 14, ]
```

# Sites Detection and Change Point Detection

## Encoding

```{r}
toy_example <- sim_result$Simulation_Output
n_col       <- nchar(ref_sequence)

colnames(toy_example)[seq_len(n_col)] <- as.character(seq_len(n_col))

# Encode character sequences to integers (A=1 … V=20, ambiguous 21–25, unknown=0)
char_mat   <- as.matrix(toy_example[, seq_len(n_col)])
AL_Cat_mix <- as.data.frame(encode_aa_sequence(char_mat))
AL_Cat_mix[] <- lapply(AL_Cat_mix, as.integer)

# Standardise Date column to the first day of each month
AL_Cat_mix$Date <- as.Date(format(as.Date(toy_example$Date), "%Y-%m-01"))
```

### View Post-Processed Matrix

```{r}
AL_Cat_mix[1:10, ]
```

## Partitioning

24 total months with a 4-month reference phase and 2-month non-overlapping
windows give 12 partitions.

> **How partitioning works:** Each partition contains all sequences observed
> within its time window. For each partition, `partition_time_windows()`
> computes the Shannon entropy at every site and then calls
> `cluster_sites_by_entropy()` to group sites into entropy classes via a
> Gaussian Mixture Model (GMM). The resulting `DataFrame` has three columns:
>
> - **`sites`** -- the site index (position in the sequence)
> - **`entropies`** -- the Shannon entropy at that site for this partition
> - **`class`** -- the raw GMM cluster label assigned by Mclust
>
> **Important:** At this stage `class` carries raw Mclust labels ordered by
> ascending component mean — class 1 corresponds to the lowest-entropy group
> and the highest class integer corresponds to the highest-entropy group.
> Function `relabel_entropy_classes()` inverts this convention so that
> class 1 consistently denotes the highest-entropy group, making
> cross-partition comparison of class labels unambiguous. Sites with zero
> entropy (invariant across all sequences in the window) are excluded from
> GMM clustering and will not appear in the DataFrame.
>
> **What to watch for:** As variants emerge and grow, their mutation sites
> should accumulate increasing entropy across successive partitions, eventually
> clustering into the highest-entropy group. The sentinel class `999` indicates
> all remaining sites have identical entropy (one undifferentiated group).

```{r}
part_data <- partition_time_windows(
  data          = AL_Cat_mix,
  n_sites       = n_col,
  window_length = 2,
  window_type   = 3,       # non-overlapping / jumping
  start_date    = "2020-01-01",
  end_date      = "2022-01-01"
)
```

### Dates Included in the First Partition

```{r}
part_data$Partitions[[1]]$Date %>% unique() %>% sort()
```

### Dates Included in the Last Partition

```{r}
part_data$Partitions[[part_data$N_partitions]]$Date %>% unique() %>% sort()
```

### Sites and Classes in Partition 1 (Jan--Feb 2020, reference only)

During the reference phase all sites are invariant or near-invariant -- no
variant signal is present. Most sites are excluded (zero entropy) or form a
single low-entropy group.

```{r}
part_data$Clusters[[1]]$DataFrame
```

### Sites and Classes in Partition 2 (Mar--Apr 2020, reference only)

```{r}
part_data$Clusters[[2]]$DataFrame
```

### Sites and Classes in Partition 3 (May--Jun 2020) vs. Variant 1 sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

Variant 1 emerged in month 5 (May 2020). After two months of growth its
mutation sites should begin to show elevated entropy relative to other sites.
Watch for the Variant 1 mutation sites appearing with higher class labels.

```{r}
part_data$Clusters[[3]]$DataFrame
```

### Sites and Classes in Partition 4 (Jul--Aug 2020)

```{r}
part_data$Clusters[[4]]$DataFrame
```

### Sites and Classes in Partition 5 (Sep--Oct 2020) vs. Variant 1 `r sim_result$Variant_Details[[1]]$pos %>% sort()` and Variant 2 `r sim_result$Variant_Details[[2]]$pos %>% sort()`

Variant 2 emerged in month 9 (Sep 2020). Both Variant 1 and Variant 2 mutation
sites should now be detectable as high-entropy clusters.

```{r}
part_data$Clusters[[5]]$DataFrame
```

### Sites and Classes in Partition 6 (Nov--Dec 2020) vs. Variant 1 `r sim_result$Variant_Details[[1]]$pos %>% sort()` and Variant 2 `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
part_data$Clusters[[6]]$DataFrame
```

### Sites and Classes in Partition 7 (Jan--Feb 2021) vs. Variants 1--3

Variant 3 emerged in month 14 (Jan 2021). Its mutation sites
(`r sim_result$Variant_Details[[3]]$pos %>% sort()`) should begin appearing
as a new elevated-entropy cluster alongside the existing variant sites.

```{r}
part_data$Clusters[[7]]$DataFrame
```

### Sites and Classes in Partition 8 (Mar--Apr 2021) vs. Variants 1--3

```{r}
part_data$Clusters[[8]]$DataFrame
```

### Sites and Classes in Partition 9 (May--Jun 2021) vs. Variants 1--3

```{r}
part_data$Clusters[[9]]$DataFrame
```

### Sites and Classes in Partition 10 (Jul--Aug 2021) vs. Variants 1--4

Variant 4 emerged in month 19 (Jul 2021). Its mutation sites
(`r sim_result$Variant_Details[[4]]$pos %>% sort()`) add to the high-entropy
pool. With four active variants, the entropy landscape is now complex and
multi-modal.

```{r}
part_data$Clusters[[10]]$DataFrame
```

### Sites and Classes in Partition 11 (Sep--Oct 2021) vs. Variants 1--4

```{r}
part_data$Clusters[[11]]$DataFrame
```

### Sites and Classes in Partition 12 (Nov--Dec 2021) vs. Variants 1--4

By the final partition all four variants have been circulating for several
months. The highest-entropy cluster should contain a union of mutation sites
from all four variants.

```{r}
part_data$Clusters[[12]]$DataFrame
```

## Hellinger Distance (HD)

The Hellinger distance quantifies the shift in amino acid distribution at each
site between a reference time point (partition T1) and each subsequent
partition. A distance of 0 means the distribution is identical to the
reference; a distance approaching 1 (when normalized) means the distribution
has changed maximally. Sites where a new variant has become prevalent will show
large, sustained Hellinger distances after the variant's emergence partition.

This is the key signal used by the change point detector: a sudden, sustained
increase in Hellinger distance at a set of sites is the signature of a new
variant entering the population.

### Calculate HD

```{r}
all_sites <- seq_len(n_col)

hellinger_mat <- calculate_hellinger_matrix(
  partitions = part_data$Partitions,
  sites      = all_sites,
  aa_levels  = 25L   # 25-character alphabet: 20 standard AAs + B, Z, X, *, -
)
```

### Examine HD Structure: `r n_col` sites x `r ncol(hellinger_mat)` partition differences

The matrix has one row per site and one column per partition comparison (T2
vs T1, T3 vs T1, ..., T12 vs T1). Rows corresponding to Variant 1 mutation
sites (`r sim_result$Variant_Details[[1]]$pos %>% sort()`) should show a
step-change in Hellinger distance starting around column T3 (the partition
covering Variant 1's first full window of growth).

```{r}
dim(hellinger_mat)
hellinger_mat
```

### Transpose for Change Point Detection

The change point algorithms expect a **time x features** matrix -- one row per
time point, one column per site. We transpose the Hellinger matrix to obtain
this format. Each row is now a multivariate observation: the full vector of
Hellinger distances across all `r n_col` sites at that partition comparison.
A change point corresponds to a partition where this vector suddenly shifts.

```{r}
dat_mat_t <- t(hellinger_mat)
dat_mat_t
```

## Change Point Detection

We apply three complementary Energy Change Point (ECP) methods to `dat_mat_t`.
All three operate on the same transposed Hellinger matrix but differ in their
algorithmic approach. A **change point** here corresponds to a partition index
at which the multivariate Hellinger distance vector shifts significantly --
i.e., a partition where new mutation sites suddenly become highly variable,
indicating a new variant has entered the population.

**Ground truth for comparison:** The four variants emerged at months 5, 9, 14
and 19, which map to partition indices 3, 5, 7 and 10 respectively (using
2-month non-overlapping windows). The correct change points in the Hellinger
series are therefore at indices **2, 4, 6 and 9** (one step behind, since
Hellinger at T_k measures the shift *relative to T1*, so the change appears
one index before the partition where the variant is first detectable).

### Method 1: Dynamic Programming via `ks.cp3o` applied to all availabe data

`ks.cp3o` uses dynamic programming to find the globally optimal segmentation
up to `K` change points. It is deterministic and exact but requires specifying
`K` (maximum number of change points) in advance. `minsize = 2` ensures each
segment contains at least 2 time points. This method is best used when the
approximate number of variants is known.

| **Parameter** | **Description**                                                              | **Default** |
|---------------|------------------------------------------------------------------------------|-------------|
| `Z`           | A `T x d` matrix: `T` time points, `d` features (sites).                     | --          |
| `K`           | Maximum number of change points to detect.                                   | 1           |
| `minsize`     | Minimum segment length (number of time points per segment).                  | 30          |
| `verbose`     | Print status updates during optimisation.                                    | FALSE       |

```{r}
cp_m1 <- ks.cp3o(
  Z       = dat_mat_t,
  K       = 11,        # upper bound: at most 11 CPs in 11 partition differences
  minsize = 2,         # each segment must span at least 2 partitions
  verbose = FALSE
)

print("Optimal change point locations (cp_m1$estimates):")
cp_m1$estimates

print("Full solution matrix across k = 1 to K change points (cp_m1$cpLoc):")
cp_m1$cpLoc
```

### Method 2: Expanding Window via `detect_changepoints_ecp`

This method applies `ks.cp3o` repeatedly with an expanding window: starting
from the first 3 partition differences and adding one time point at a time.
It simulates a real-time surveillance scenario where data accumulates
sequentially and you want to detect a change point as early as possible, without
knowing how many variants will ultimately emerge. `dynamic_k = TRUE` sets `K`
adaptively to `nrow(window) - 2` at each step.

```{r}
lower     <- 1
upper     <- 3
timesteps <- nrow(dat_mat_t) - upper

ECP_res <- detect_changepoints_ecp(
  data_matrix    = dat_mat_t,
  min_window     = lower,
  max_window     = upper,
  n_timesteps    = timesteps,
  rolling_window = FALSE,
  dynamic_k      = TRUE,
  minsize        = 2,
  verbose        = FALSE
)

# Union of all change points detected across all window sizes.
# Indicates which true change points the algorithm recovers
# at any stage of data accumulation (overall sensitivity).
ecp_indices <- ECP_res$ECP_est_list %>%
  unlist() %>%
  .[. > 1] %>%
  unique() %>%
  sort()

print("Change points detected at any window size (union):")
ecp_indices

# Early detection timeline: for each expanding window size, which true
# change points are detectable with the data available up to that point?
# Ground truth: variants emerged at partition indices 3, 5, 7, 10,
# corresponding to Hellinger change point indices 2, 4, 6, 9.
true_cps <- c(2, 4, 6, 9)

detection_timeline <- data.frame(
  window_size  = seq(upper, nrow(dat_mat_t)),
  detected_cps = sapply(ECP_res$ECP_est_list, function(cps) {
    detected <- intersect(true_cps, cps[cps > 1])
    if (length(detected) == 0L) "none" else paste(sort(detected), collapse = ", ")
  })
)

print("Early detection timeline — true change points recovered per window size:")
detection_timeline
```

### Method 3: Agglomerative Hierarchical ECP via `e.agglo`

`e.agglo` applies agglomerative hierarchical clustering directly to the 
multivariate time series, merging adjacent segments iteratively until a 
global goodness-of-fit criterion is maximised. Unlike `ks.cp3o`, it does 
not require the number of change points `K` to be specified in advance, 
making it well suited to surveillance settings where the number of emerging 
variants is unknown. Setting `penalty = function(cps) 0` imposes no 
regularisation penalty on model complexity, allowing the segmentation to 
be driven entirely by the energy distance signal in the data.

The `cluster` output assigns each time point to a segment, providing an
intuitive segmentation of the surveillance timeline that can be directly
compared to known variant emergence periods.

| **Parameter** | **Description**                                                              | **Default**  |
|---------------|------------------------------------------------------------------------------|--------------|
| `X`           | A `T x d` matrix: `T` time points, `d` features (sites).                     | --           |
| `member`      | Initial segment membership vector (one integer per time point).              | `1:nrow(X)`  |
| `alpha`       | Moment index for the energy distance metric (1 = first moment).              | 1            |
| `penalty`     | Penalty function applied to the goodness-of-fit statistic.                   | --           |

```{r}
cp_m3 <- e.agglo(
  X       = dat_mat_t,
  member  = seq_len(nrow(dat_mat_t)),  # naive: treat each time point as its own segment
  alpha   = 1,
  penalty = function(cps) { 0 }        # no penalty: let the data drive the segmentation
)

print("Change points estimated by method 3 (naive, no penalty):")
cp_m3$estimates

print("Segment membership per time point:")
cp_m3$cluster
```

### Comparing Methods Against Ground Truth

Overall, no single method recovers all four true change points cleanly,
which is expected given the simulation design: stochastic deletion events
and the exponential growth competition model create a complex,
heterogeneous signal. Method 3 is notable for detecting the latest true
change point (9) which the other methods miss, while Method 1 provides
the cleanest intermediate detection (4 and 6) and identifies latest change point 
(9) one period later (10). 

Method 2 detects indices 4 and 6 but only once a sufficient number of
time points have accumulated in the expanding window — a direct
consequence of sample size. This lag is not a failure of the algorithm but an 
inherent property of prospective surveillance: reliable change point detection
requires sufficient sample mass to warrant a shift in the multivariate 
distribution. A formal analysis of the relationship between sample size and
detection reliability, including scenario comparisons across varying sequence 
counts and variant growth rates, is provided in 
`analysis/Sample_Size_Simulation_Study/`.

# Conclusion

This vignette demonstrates the complete `ViralEntropR` detection pipeline
on a fully controlled, traceable example. By constructing a synthetic
evolutionary timeline with known variant emergence points, mutation sites,
and stochastic perturbations, we establish a clear ground truth against
which all pipeline components can be evaluated directly. The simulation
output confirms that entropy-based site selection captures genuine
evolutionary signal: mutation sites show measurably elevated entropy
from the partition following their variant's emergence, progressively
transitioning into higher GMM entropy classes as the variant grows in
prevalence. This transition is not instantaneous — newly emerged variants
contribute only a small number of sequences relative to the reference
background, and their mutation sites require one to several partitions
of growth before the entropy shift becomes large enough to enter the
highest-entropy class. This detection lag is an inherent property of
any population-level surveillance system and reflects the time required
for a novel variant to accumulate sufficient frequency to perturb the
site-level amino acid distribution. Once the shift is established, it
propagates as a measurable, sustained increase in Hellinger distance that
is recoverable by all three change point algorithms, with each method
offering a different trade-off between sensitivity and specificity as
shown above.

The `simulate_variant_evolution()` function is designed to be a general-purpose
benchmarking tool for bioinformatics researchers. The full parameter space, 
including  number of variants, mutation counts per variant, emergence timing, 
growth rates, stochastic deletion events, reference variability, and random 
seeds allows the systematic generation of arbitrarily large scenario libraries. 
Researchers can  use this infrastructure to stress-test detection algorithms 
under varying levels of signal strength, variant competition, background noise, 
and sampling density, and to compare the sensitivity and specificity of 
alternative change point methods before applying them to real surveillance data. 
The traceable ground  truth embedded in every simulation output makes formal 
evaluation of detection accuracy straightforward, as demonstrated in the method
comparison above.

# References