---
title: "NCBI SARS-CoV-2 Spike Protein Sequence Preprocessing: From Raw FASTA to an Analysis-Ready Integer-Encoded Matrix"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: flatly
    highlight: tango
bibliography: references_preprocessing.bib
nocite: '@*'
vignette: >
  %\VignetteIndexEntry{NCBI SARS-CoV-2 Spike Protein Sequence Preprocessing: From Raw FASTA to an Analysis-Ready Integer-Encoded Matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

# Introduction

This vignette demonstrates the complete `ViralEntropR` preprocessing pipeline
for real-world SARS-CoV-2 amino acid sequence data downloaded from the NCBI
SARS-CoV-2 Data Hub. Starting from a raw FASTA file, the pipeline produces a
clean, integer-encoded data frame with a `Date` column ready for
`partition_time_windows()` and all downstream entropy and Hellinger distance
analyses.

The pipeline consists of five steps:

1. **Download and load** the full FASTA dataset.
2. **Confirm** the canonical Spike protein length (1,273 aa).
3. **Extract and validate** collection dates and countries from FASTA headers.
4. **Remove** sequences with missing or unparseable metadata.
5. **Convert, filter, encode, and assemble** the final analysis-ready data frame.

> **Reproducibility note:** The full 137,132-sequence dataset (~181.5 MB) is
> archived on Zenodo (DOI: `10.5281/zenodo.19040165`).

# Setup

```{r libraries}
library(ViralEntropR)
library(Biostrings)
library(dplyr)
library(ggplot2)
```

# Step 1: Load the FASTA Dataset

## Download Data

The complete NCBI SARS-CoV-2 Spike protein dataset (137,132 sequences, ~181.5 MB) 
is archived on Zenodo under DOI [10.5281/zenodo.19040165](https://zenodo.org/records/19040165).
Download `sequences.fasta` manually from the Zenodo record and update the path 
in the code chunk below to match your local copy. 

```{r fasta-path-local, eval=TRUE, include=FALSE}
fasta_path <- file.path("data-raw", "sequences.fasta")
```

## Update Path

```{r fasta-path-fake, eval=FALSE}
fasta_path <- "path/to/sequences.fasta"   # update this
```

## Read FASTA file

```{r load-fasta-local}
fasta <- Biostrings::readAAStringSet(fasta_path)
sprintf("Total sequences loaded:  %d", length(fasta))
sprintf("Unique sequences:        %d", length(unique(fasta)))
```

Inspect a sample of sequence headers to understand the metadata layout before
extraction. NCBI Virus exports headers in a pipe-delimited format:

```
Accession | Country | Collection_Date |
```

```{r inspect-headers}
# Print 5 randomly sampled headers to see the field layout
set.seed(42)
cat(paste(sample(names(fasta), 5), collapse = "\n\n"))
```

## Curated metadata for 12 SARS-CoV-2 VOCs and VOIs

```{r variants}

# --- Overview of all 12 variants -------------------------------------------

data.frame(
  WHO_Label      = unlist(sarscov2_variants$WHO_Label),
  Pango_Lineage  = unlist(sarscov2_variants$Pango_Lineage),
  GISAID_Clade   = sarscov2_variants$GISAID_Clade,
  Nextstrain     = sarscov2_variants$Nextstrain_Clade,
  First_Detected = sarscov2_variants$Date_First_Detected,
  First_US       = sarscov2_variants$Date_First_Detected_US
)
```

```{r,eval=FALSE}
# To learn more
?sarscov2_variants
```

## Bundled FASTA sample

```{r sample}
# sarscov2_sample.fasta.gz contains 100 random sequences from the full NCBI dataset.
# Useful for testing pipeline code without downloading the full ~181.5 MB file.

path_sample  <- system.file("extdata", "sarscov2_sample.fasta.gz",
                     package = "ViralEntropR")
fasta_sample <- Biostrings::readAAStringSet(path_sample)

sprintf("Sample sequences: %d", length(fasta_sample))

# check NAMES structure
sample(fasta_sample@ranges@NAMES,1)

# Countries represented in the sample
sort(table(extract_fasta_countries(fasta_sample, position = 2)$countries),
     decreasing = TRUE)
```

```{r,eval=FALSE}
# To learn more
?sarscov2_sample
```

# Step 2: Canonical Spike Protein Length

All sequences in this dataset are 1,273 amino acids in length. No length
filtering is required because the NCBI download was configured to retrieve only
complete Spike protein sequences, which naturally match the canonical reference.

> **Why 1,273 amino acids?** The SARS-CoV-2 Spike glycoprotein reference
> sequence ([YP_009724390](https://www.ncbi.nlm.nih.gov/protein/1796318598))
> is 1,273 amino acids long. This length corresponds to the full Spike
> precursor protein, comprising the signal peptide, the S1 subunit (receptor
> binding domain, N-terminal domain, and furin cleavage site), and the S2
> subunit (fusion peptide, heptad repeats, transmembrane domain, and
> cytoplasmic tail). This fixed-length representation enforces positional 
> homology by construction, allowing site indices to be interpreted consistently 
> across sequences.

```{r confirm-length}
sprintf("All sequences are length 1,273: %s",
        all(width(fasta) == 1273L))
```

# Step 3: Extract and Validate Collection Dates

## Initial Date Extraction (yyyy-mm-dd)

We first attempt full-precision date extraction (`yyyy-mm-dd`) to identify
sequences that only carry year-month (`yyyy-mm`) or year-only (`yyyy`)
dates -- both common in NCBI submissions.

```{r dates-pass1}
# Pass 1: attempt full yyyy-mm-dd extraction
dates_ymd <- extract_fasta_dates(
  fasta,
  custom_pattern = "(?<=\\|)[0-9]{4}-(0?[1-9]|1[0-2])-(0?[1-9]|[12][0-9]|3[01]|00)"
)

sprintf("Sequences with full yyyy-mm-dd dates: %d",
        sum(!is.na(dates_ymd$raw_date_strings)))
sprintf("Sequences missing full date:          %d",
        sum(is.na(dates_ymd$raw_date_strings)))
```

```{r}
# Inspect headers of sequences that lack a full yyyy-mm-dd date
if (length(dates_ymd$missing_id) > 0) {
  cat("Sample headers with partial/missing dates:\n")
  cat(paste(head(names(fasta)[dates_ymd$missing_id], 5), collapse = "\n"))
}
```

## Second Pass: yyyy-mm Extraction

For monthly partitioning, `yyyy-mm` precision is sufficient and more inclusive
than requiring the full day. We re-extract dates in `yyyy-mm` format. Any
sequence that only has a year (`yyyy`) with no month will still fail and must
be removed.

```{r dates-pass2}
# Pass 2: extract yyyy-mm (covers both yyyy-mm-dd and yyyy-mm submissions)
dates_result <- extract_fasta_dates(
  fasta,
  custom_pattern = "(?<=\\|)[0-9]{4}-(0[1-9]|1[0-2])",
  date_format    = "%Y-%m"
)

sprintf("Sequences with parseable yyyy-mm dates: %d", 
        sum(!is.na(dates_result$raw_date_strings)))
sprintf("Sequences still missing a date:         %d",
        sum(is.na(dates_result$raw_date_strings)))
```

```{r}
# Sequences with year-only dates -- these cannot be placed in a monthly window
if (length(dates_result$missing_id) > 0) {
  cat("Headers with year-only dates (to be removed):\n")
  cat(paste(head(names(fasta)[dates_result$missing_id], 5), collapse = "\n"))
}
```

```{r}
# Remove sequences with year-only dates
if (length(dates_result$missing_id) > 0) {
  fasta        <- fasta[-dates_result$missing_id]
  dates_result <- extract_fasta_dates(
    fasta,
    custom_pattern = "(?<=\\|)[0-9]{4}-(0[1-9]|1[0-2])",
    date_format    = "%Y-%m"
  )
}

cat("Date extraction status after removal:", dates_result$message, "\n")
sprintf("Sequences remaining: %d", length(fasta))
```

## Date Distribution (Before Ambiguous Residue Filtering)

Visualising the date distribution helps verify that the data spans the expected
surveillance window and that no anomalous date clusters are present.

```{r date-plot-before, fig.width=10, fig.height=4}
dates_df_before <- data.frame(
  Date = as.Date(dates_result$corrected_dates)
) %>%
  mutate(YearMonth = format(Date, "%Y-%m"))

ggplot(dates_df_before, aes(x = YearMonth)) +
  geom_bar(stat = "count", fill = "steelblue") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    plot.title  = element_text(hjust = 0.5)
  ) +
  labs(
    x     = "Collection Month",
    y     = "Number of Sequences",
    title = "Sequence Count by Collection Month (before ambiguous residue filtering)"
  )
```

# Step 4: Extract and Validate Countries

```{r countries}
# Extract country field (position 2 = between first and second pipe)
countries_result <- extract_fasta_countries(fasta, position = 2)

sprintf("Sequences with parseable country: %d", 
        sum(!is.na(countries_result$countries)))
sprintf("Sequences with missing country:   %d",
        sum(is.na(countries_result$countries)))
```

```{r}
# Inspect headers with missing countries before removing
if (length(countries_result$missing_id) > 0) {
  cat("Sample headers with missing country field:\n")
  cat(paste(head(names(fasta)[countries_result$missing_id], 5), collapse = "\n"))
  
  # Remove sequences with missing countries
  fasta            <- fasta[-countries_result$missing_id]
  countries_result <- extract_fasta_countries(fasta, position = 2)
}

cat("Country extraction status after removal:", countries_result$message, "\n")
sprintf("Sequences remaining: %d", length(fasta))
```

```{r}
# Re-extract dates after removing sequences with missing countries
dates_result <- extract_fasta_dates(
  fasta,
  custom_pattern = "(?<=\\|)[0-9]{4}-(0[1-9]|1[0-2])",
  date_format    = "%Y-%m"
)

# Confirm no remaining missing values in either field
sprintf("Missing dates remaining:     %d", sum(is.na(dates_result$corrected_dates)))
sprintf("Missing countries remaining: %d", sum(is.na(countries_result$countries)))
```

```{r}
# Country distribution -- sequences per country of collection
countries_result$countries %>%
  table() %>%
  sort(decreasing = TRUE) 
```

# Step 5: Convert, Filter, and Encode the Sequence Matrix

## Convert FASTA to Character Matrix

`fasta_to_char_matrix()` converts the `AAStringSet` object to a plain R
character matrix (rows = sequences, columns = sites). This fully vectorized
implementation is substantially faster than row-by-row `strsplit` approaches
on large datasets.

```{r fasta-to-matrix}
char_mat <- fasta_to_char_matrix(fasta)

sprintf("Character matrix dimensions: %d sequences x %d sites",
        nrow(char_mat), ncol(char_mat))

# Inspect a small region
char_mat[1:5, 1:10]
```

## Filter Ambiguous Residues

Ambiguous amino acid codes (B, X, Z) and unrecognised characters (encoded as 0)
can distort entropy and Hellinger distance calculations by inflating apparent
diversity at a site. `filter_ambiguous_sequences()` removes any sequence
containing at least one such residue.

`option = 2` specifies a character matrix input (as returned by
`fasta_to_char_matrix()`).

```{r filter-ambiguous}
filtered <- filter_ambiguous_sequences(char_mat, option = 2)

cat(filtered$OriginalDim,     "\n")
cat(filtered$NewDim,          "\n")
cat(filtered$NumberAmbiguous, "\n")
cat(filtered$RangeAmbiguous,  "\n")
```

```{r}
# Extract the clean matrix and align metadata to retained rows
clean_char_mat  <- filtered$FilteredMatrix
deleted_rows    <- filtered$DeletedSeqId

# Remove deleted rows from date and country vectors
corrected_dates <- dates_result$corrected_dates[-deleted_rows]
countries       <- countries_result$countries[-deleted_rows]

sprintf("Clean matrix: %d sequences x %d sites",
        nrow(clean_char_mat), ncol(clean_char_mat))
```

## Encode Amino Acids to Integers

`encode_aa_sequence()` maps each character to its integer code using the
standard 25-character `ViralEntropR` alphabet. After ambiguous residue
filtering only codes 1--20 (the 20 standard amino acids) should remain.

| Codes | Characters | Meaning |
|---|---|---|
| 1--20 | A R N D C Q E G H I L K M F P S T W Y V | Standard amino acids |
| 21 | B | Asp/Asn ambiguity |
| 22 | Z | Glu/Gln ambiguity |
| 23 | X | Any amino acid (unknown) |
| 24 | * | Stop codon |
| 25 | - | Gap |
| 0  | (other) | Unrecognised character |

```{r encode}
int_mat <- encode_aa_sequence(clean_char_mat)

# Verify only codes 1-20 remain
code_range <- range(int_mat)
sprintf("Encoded value range: %d to %d  (expected: 1 to 20 after filtering)",
        code_range[1], code_range[2])
```

# Step 6: Assemble the Final Data Frame

We combine the integer matrix with the aligned date and country metadata into
a single data frame, standardise dates to the first of each month for monthly
partitioning, and sort by date. This object is passed directly to
`partition_time_windows()`.

```{r assemble}
n_sites <- ncol(int_mat)

# Build the data frame
AL_df           <- as.data.frame(int_mat)
colnames(AL_df) <- as.character(seq_len(n_sites))
AL_df[]         <- lapply(AL_df, as.integer)

# Standardise to first of each month for consistent monthly partitioning
AL_df$Date    <- as.Date(format(corrected_dates, "%Y-%m-01"))
AL_df$Country <- countries

# Sort by date -- required by partition_time_windows()
AL_df         <- AL_df[order(AL_df$Date), ]
rownames(AL_df) <- NULL

sprintf("Final data frame: %d sequences x %d site columns + Date + Country",
        nrow(AL_df), n_sites)
```

```{r}
# Preview first few rows -- site columns and metadata
AL_df[1:10, c(1:5, n_sites + 1, n_sites + 2)]
```

```{r}
# Country distribution in the clean final dataset
AL_df$Country %>% table() %>% sort(decreasing = TRUE)
```

## Date Distribution (After Filtering)

Compare the final date distribution against the earlier plot to confirm that
the filtering steps have not introduced systematic temporal bias.

```{r date-plot-after, fig.width=10, fig.height=4}
ggplot(AL_df, aes(x = format(Date, "%Y-%m"))) +
  geom_bar(stat = "count", fill = "steelblue") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    plot.title  = element_text(hjust = 0.5)
  ) +
  labs(
    x     = "Collection Month",
    y     = "Number of Sequences",
    title = "Sequence Count by Collection Month (clean dataset, ready for analysis)"
  )
```

```{r}
sprintf("Date range: %s to %s",
        format(min(AL_df$Date), "%Y-%m"),
        format(max(AL_df$Date), "%Y-%m"))

sprintf("Total sequences in final dataset: %d", nrow(AL_df))
```

## Filter `US` Sequences and Save

```{r save-post-processed-file, eval = FALSE}
AL_df_US <- AL_df[!is.na(AL_df$Country) & AL_df$Country == "USA", ]
sprintf("Sequences after country filter (%s): %d", "USA", nrow(AL_df_US))

saveRDS(AL_df_US, file.path("data-raw", "NCBI_US_unaligned_feature_matrix_1273aa.rds"))
```

# Downstream Analysis

The data frame `AL_df` is now ready for the full `ViralEntropR` pipeline. The
site columns contain integer-encoded amino acids (1--20), `Date` contains
first-of-month dates for monthly partitioning, and `Country` is available for
geographic stratification.

```{r partition-example}
# Example: partition into non-overlapping 2-month windows
part_data <- partition_time_windows(
  data          = AL_df,
  n_sites       = n_sites,
  window_length = 2,
  window_type   = 3,      # non-overlapping / jumping windows
  verbose       = FALSE
)

sprintf("Number of 2-month partitions created: %d", part_data$N_partitions)
cat("\nPartition date labels:\n")
cat(paste(part_data$Dates_Labels, collapse = "\n"))
```

```{r}
# Entropy-based site clustering for the first partition
cat("\nSite clustering for partition 1 (reference period):\n")
part_data$Clusters[[1]]$DataFrame
```

As the output shows, entropy-based analysis of early pandemic sequences
already flags several sites that subsequently proved critical across multiple
VOCs and VOIs:

- **Site 95 (T95I)** -- located in the N-terminal domain (NTD) of the Spike
  protein. The T95I substitution alters the topology of the NTD neutralization
  "supersite", reducing antibody recognition, and is associated with increased
  viral load. It is a defining mutation of the Iota (B.1.526) variant and also
  appears in Delta and Kappa [@Magazine2022].

- **Site 655 (H655Y)** -- located immediately upstream of the furin cleavage
  site at the S1/S2 boundary. The H655Y substitution enhances spike protein
  cleavage efficiency and viral replication, and has been described as a
  "gateway mutation" that promotes fitness in emerging lineages. It is a
  defining feature of Gamma (P.1) and Omicron [@Escalera2022].

- **Site 677 (Q677H/P)** -- situated only three residues upstream of the
  polybasic furin cleavage site (RRAR). Mutations at this position are thought
  to promote conformational changes that enhance accessibility of the S1/S2
  junction to proteolytic cleavage, increasing viral entry efficiency. Q677H
  independently evolved in at least six distinct US lineages in late 2020,
  a striking example of convergent evolution under positive selection [@Zeng2021].

The early detection of these sites by entropy clustering -- before the variants
carrying them were formally designated -- illustrates the prospective
surveillance potential of the `ViralEntropR` pipeline.

```{r}
# Entropy-based site clustering for the last partition
cat("\nSite clustering for partition 10 (final partition):\n")
part_data$Clusters[[10]]$DataFrame
```

From here, pass `part_data$Partitions` to `calculate_hellinger_matrix()` to
compute Hellinger distances across time windows, then transpose and apply
`detect_changepoints_ecp()` or `e.agglo()` for change point detection. See the
companion vignette *"Entropy Clustering, Hellinger Distance, and Change Point Analysis for Emerging Viral Variant Detection: A Simulation Study"* for a complete worked example.

# Pipeline Summary

```
Download sequences.fasta manually from Zenodo (https://zenodo.org/records/19040165)
  |
  v
readAAStringSet()                 # Step 1: Load FASTA into R
  |
  v
Confirm width() == 1273L          # Step 2: All sequences match canonical Spike length
                                  # (YP_009724390 -- no filtering required)
  |
  v
extract_fasta_dates()             # Step 3: Pass 1 -- attempt yyyy-mm-dd
  -> re-extract with yyyy-mm      #          Pass 2 -- broader pattern, remove year-only
  |
  v
extract_fasta_countries()         # Step 4: Extract country field from header
  -> remove missing               #          Iterative: remove, re-extract, confirm
  |
  v
fasta_to_char_matrix()            # Step 5: AAStringSet -> character matrix (rows x sites)
  |
  v
filter_ambiguous_sequences()      #          Remove sequences with B, X, Z, unknown
  -> align date/country vectors   #          Drop same rows from metadata
  |
  v
encode_aa_sequence()              #          Character -> integer (25-AA alphabet)
  |
  v
Assemble data frame               #          Add Date (yyyy-mm-01) + Country, sort
  |
  v
partition_time_windows()          # Downstream: partition into time windows
  |                               #             -> entropy clustering per partition
  v
calculate_hellinger_matrix()      #             Hellinger distances (sites x partitions)
  |
  v
detect_changepoints_ecp()         #             Change point detection
```

# References

