---
title: "Gaussian Mixture Model Site Selection on SARS-CoV-2 Spike Protein Shannon Entropy Distributions: 
Robustness to Monotone Transformations"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    includes:
      in_header: meta_gmm.html
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.height = 6,
  fig.width  = 12
)
```

# Overview

Gaussian Mixture Model (GMM) clustering is applied to per-site Shannon entropy
vectors derived from two SARS-CoV-2 spike protein datasets --- GISAID and NCBI
--- to identify sites exhibiting elevated evolutionary variability[@Fraley2002; 
@Cover2006; @Shannon1948].The primary goal of this analysis is not site detection 
per se, but rather to assess whether applying monotone transformations to the 
entropy distribution prior to GMM fitting provides any additional information or
improvement in site selection, and whether the added layer of methodological 
complexity is warranted.

A natural question is whether the pronounced right-skewness of the entropy
distribution, characteristic of genomic surveillance data where the majority of
sites are highly conserved, adversely affects GMM fitting and subsequent site
selection. To investigate this, six transformations are applied to the entropy
vectors and the resulting site selections are compared with those obtained on the
original scale. Five of the six transformations are monotone increasing ---
$\log(1+x)$, $\log(1+\log(1+x))$, $\sqrt{x}$, Box-Cox ($\lambda = -0.05$),
and logarithmic $\ln(x)$ --- and are therefore expected to preserve rank-based site
selection. The reciprocal $\frac{1}{1+x}$ is monotone decreasing and is included
as a deliberate contrast: by inverting the entropy ranking it provides a natural
negative control, confirming that site selection depends critically on the
preservation of rank ordering. While monotone increasing transformations preserve
rank ordering, Gaussian mixture models depend on the metric structure of the data
--- including inter-point distances and component variances --- and are therefore
not guaranteed to yield identical cluster assignments under such
transformations[@McLachlan2000; @Hennig2015]. Despite this, empirical
comparisons indicate that, although monotone increasing transformations alter the
geometry of the data, they do not materially affect the identification of
high-entropy sites in the SARS-CoV-2 spike protein datasets considered here,
and the additional methodological complexity they introduce is not justified.

**Document structure and scope of analysis**
The complete entropy distributions (all 1273 sites) are visualised initially to
illustrate the characteristic right-skewed profile of genomic sequence
variability, where most positions exhibit near-zero entropy and a small subset
displays elevated variability[@Korber2020]. All subsequent analysis ---
including GMM fitting, transformation experiments, and comparison tables --- is
conducted on filtered entropy vectors with zero-entropy sites removed. These
positions carry no variability signal and therefore do not contribute to
clustering. Additionally, singleton-entropy sites (observed in the NCBI dataset)
are excluded, as they are likely dominated by rare events or sequencing
artefacts rather than sustained evolutionary variation.

For each entropy vector, six sub-vectors are analysed: the complete filtered
vector and five trimmed subsets --- Q1+, median+, mean+, Q3+, and the 90th
percentile+. This constitutes a sensitivity analysis of GMM-based site selection
to the choice of pre-filtering threshold.

**Distributions visualization**
Throughout this document, input entropy distributions are rendered in
<span style="color:#4682B4;">**blue (`#4682B4`)**</span> and GMM-selected
site distributions in <span style="color:#2CA25F;">**green (`#2CA25F`)**</span>.

---

# Libraries and Helper Functions

```{r libraries}
library(ViralEntropR)
library(kableExtra)
library(dplyr) 
library(mclust)
library(ggplot2)
library(patchwork)
library(scales)
library(here)

source("gmm_site_selection.R")
source("generate_comparison_table.R")
```

---

# Output Directory

```{r output-dir}
TABLE_DIR <- here("analysis", "GMM_transformation_study", "tables")
dir.create(TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
table_path <- function(filename) file.path(TABLE_DIR, filename)
```

---

# Reference Sites

VOC-defining mutation sites are extracted from the `sarscov2_variants` package
dataset.  Two reference sets are used: NCBI comparisons use Alpha, Beta, Gamma,
and Delta (the VOCs present in the October 2021 NCBI dataset); GISAID
comparisons additionally include Omicron, which is represented in the GISAID
dataset through April 2024.

```{r important-sites}
data("sarscov2_variants", package = "ViralEntropR")

extract_voc_sites <- function(labels) {
  sort(unique(unlist(lapply(labels, function(label) {
    idx <- which(sarscov2_variants$WHO_Label == label)
    if (length(idx) == 0L) {
      warning(sprintf("WHO_Label '%s' not found in sarscov2_variants.", label))
      return(integer(0L))
    }
    sarscov2_variants$Mutation_Sites[[idx]]
  }))))
}

voc_labels_ncbi  <- c("Alpha", "Beta", "Gamma", "Delta")
voc_labels_gisaid <- c("Alpha", "Beta", "Gamma", "Delta", "Omicron")

important_sites_ncbi  <- extract_voc_sites(voc_labels_ncbi)
important_sites_gisaid <- extract_voc_sites(voc_labels_gisaid)

sprintf("NCBI  reference sites (Alpha/Beta/Gamma/Delta): %d",
        length(important_sites_ncbi))
print(important_sites_ncbi)

sprintf("GISAID reference sites (Alpha/Beta/Gamma/Delta/Omicron): %d",
        length(important_sites_gisaid))
print(important_sites_gisaid)
```

---

# Custom Transformation Objects

All transformations are monotone on the entropy support, ensuring rank ordering
is preserved. Cluster membership may still change when the number of components
G selected by BIC differs between scales; the comparison tables quantify this
empirically.

```{r transformations}
loglog1p1p_trans <- scales::trans_new(
  name      = "log1p(log1p)",
  transform = function(x) log(log(x + 1) + 1),
  inverse   = function(x) exp(exp(x) - 1) - 1,
  breaks    = scales::log_breaks()
)

reciprocal_trans <- scales::trans_new(
  name      = "reciprocal(1+x)",
  transform = function(x) 1 / (x + 1),
  inverse   = function(x) 1 / x - 1,
  breaks    = scales::pretty_breaks()
)

boxcox_trans <- function(lambda = -0.05) {
  scales::trans_new(
    name      = sprintf("Box-Cox(\u03bb=%.2f)", lambda),
    transform = function(x) if (lambda == 0) log(x) else (x^lambda - 1)/lambda,
    inverse   = function(x) if (lambda == 0) exp(x) else (lambda*x + 1)^(1/lambda),
    breaks    = scales::pretty_breaks()
  )
}
```

---

# Data

## Loading

`GISAID_US_aligned_feature_matrix_1273aa` and `Data_Matrix_Filtered_NCBI_Data_US`
are fully post-processed data matrices containing United States sequences only.
The preprocessing steps for the NCBI dataset are documented in the 
`preprocessing_pipeline` package vignette 
(`vignette("preprocessing_pipeline", package = "ViralEntropR")`); the GISAID
pipeline is recorded in `analysis/GISAID_data_preprocessing/GISAID_Preprocessing_Pipeline.Rmd`.

Each matrix has rows = sequences, columns = integer-encoded amino acid sites
1–1273 plus `Date` and `Country` metadata columns.

```{r load-data}
AL_df_GISAID <- readRDS(here("analysis","GISAID_data_preprocessing", 
                             "GISAID_US_aligned_feature_matrix_1273aa.rds"))
AL_df_NCBI   <- readRDS(here("data-raw", "Data_Matrix_Filtered_NCBI_Data_US"))

nr_gisaid <- nrow(AL_df_GISAID)
nr_ncbi   <- nrow(AL_df_NCBI)

nc_gisaid <- ncol(AL_df_GISAID)-2 # Date and Country
nc_ncbi   <- ncol(AL_df_NCBI)-2   # Date and Country

sprintf("GISAID: %d sequences x %d sites", nr_gisaid,nc_gisaid)
sprintf("NCBI:   %d sequences x %d sites", nr_ncbi,nc_ncbi)
```

## Entropy Vectors

```{r compute-entropy}
Entrop_Vec_GISAID <- apply(AL_df_GISAID[, 1:1273], 2, calculate_entropy)
Entrop_Vec_NCBI   <- apply(AL_df_NCBI[,   1:1273], 2, calculate_entropy)

sprintf("GISAID entropy range: [%.6f, %.6f]",
        min(Entrop_Vec_GISAID), max(Entrop_Vec_GISAID))
sprintf("NCBI   entropy range: [%.6f, %.6f]",
        min(Entrop_Vec_NCBI), max(Entrop_Vec_NCBI))
```

## Zero and Singleton Removal

```{r preprocessing}
sngl_gisaid <- -(
  (1/nr_gisaid) * log2(1/nr_gisaid) +
  ((nr_gisaid-1)/nr_gisaid) * log2((nr_gisaid-1)/nr_gisaid)
)
sngl_ncbi <- -(
  (1/nr_ncbi) * log2(1/nr_ncbi) +
  ((nr_ncbi-1)/nr_ncbi) * log2((nr_ncbi-1)/nr_ncbi)
)

sprintf("GISAID — zero-entropy sites:      %d", sum(Entrop_Vec_GISAID == 0))
sprintf("NCBI   — zero-entropy sites:      %d", sum(Entrop_Vec_NCBI == 0))
sprintf("NCBI   — singleton-entropy sites: %d",
        sum(Entrop_Vec_NCBI == round(sngl_ncbi, 6L)))
sprintf("GISAID   — singleton-entropy sites: %d",
        sum(Entrop_Vec_GISAID == round(sngl_gisaid, 6L)))

Entrop_Vec_GISAID_no0S <- Entrop_Vec_GISAID[Entrop_Vec_GISAID != 0]

Entrop_Vec_NCBI_no0S <- Entrop_Vec_NCBI[
  Entrop_Vec_NCBI != 0 &
  Entrop_Vec_NCBI != round(sngl_ncbi, 6L)
]

sprintf("GISAID — sites retained after zero removal:             %d",
        length(Entrop_Vec_GISAID_no0S))
sprintf("NCBI   — sites retained after zero + singleton removal: %d",
        length(Entrop_Vec_NCBI_no0S))
```
---

# Complete Data — Distribution Overview

The entropy distributions of the complete (unfiltered) vectors are shown here
for both datasets to illustrate the characteristic right-skewed profile.  No
GMM is fitted to the complete data; all subsequent analysis uses the
zero/singleton-removed vectors.

```{r dist-complete-gisaid}
plot_entropy_distributions(Entrop_Vec_GISAID, prob = 0.9,
                           dataset_label = "GISAID \u2014 complete (all 1273 sites)",
                           bins = 10)
```

```{r dist-complete-ncbi}
plot_entropy_distributions(Entrop_Vec_NCBI, prob = 0.9,
                           dataset_label = "NCBI \u2014 complete (all 1273 sites)",
                           bins = 10)
```

---

# Untransformed Analysis (Zeros/Singletons Removed)

GMM is fitted on the zero/singleton-removed entropy vectors.  The component
with the highest mean entropy is selected as the high-variability cluster.

## GISAID

```{r gmm-gisaid-no0s}
GISAID_gmm_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S,
  prob             = 0.9,
  transfr          = NULL,
  hist             = TRUE,
  calc_gmm_orig    = TRUE,
  calc_gmm_transfr = FALSE,
  dataset_label    = "GISAID \u2014 zeros removed",
  bins             = 20
)
```

```{r table-gisaid-untransf}
tbl <- save_orig_comparison_table(
  GISAID_gmm_no0S,
  table_caption   = "Site selection \u2014 GISAID, zeros removed, untransformed",
  save_path       = table_path("NonTransformed_GISAID_no0S.html"),
  important_sites = important_sites_gisaid
)
tbl
```

## NCBI

```{r gmm-ncbi-no0s}
NCBI_gmm_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S,
  prob             = 0.9,
  transfr          = NULL,
  hist             = TRUE,
  calc_gmm_orig    = TRUE,
  calc_gmm_transfr = FALSE,
  dataset_label    = "NCBI \u2014 zeros and singletons removed",
  bins             = 20
)
```

```{r table-ncbi-untransf}
tbl <- save_orig_comparison_table(
  NCBI_gmm_no0S,
  table_caption   = "Site selection \u2014 NCBI, zeros and singletons removed, untransformed",
  save_path       = table_path("NonTransformed_NCBI_no0S.html"),
  important_sites = important_sites_ncbi
)
tbl
```

---

# Transformation Experiments

All transformations are applied exclusively to the zero/singleton-removed
vectors.  For each transformation, the GMM is fitted on the transformed scale;
selected sites are reported back on the original entropy scale for
interpretability.

## Box-Cox Lambda Exploration

$$
x^{(\lambda)} = \begin{cases} \dfrac{x^{\lambda} - 1}{\lambda} & \text{if } \lambda \neq 0 \\[6pt] \ln(x) & \text{if } \lambda = 0 \end{cases}
$$

where $x > 0$ is the per-site Shannon entropy and $\lambda$ is the
transformation parameter. For $\lambda$ close to zero the Box-Cox transformation
approximates the natural logarithm, providing near-logarithmic compression of
the right tail while remaining defined at $x = 0$ in the limit as
$\lambda \to 0$.

Nine equally spaced values of λ across [−2, 2] are applied to the
zero/singleton-removed vectors to identify which value most effectively
yield the most symmetric shape of the distribution.  

```{r boxcox-lambda-exploration}
boxcox_transform <- function(x, lambda) {
  if (lambda == 0) log(x) else (x^lambda - 1) / lambda
}
lambdas <- seq(-2, 2, length.out = 9L)

plot_boxcox_grid <- function(ev, dataset_label) {
  plots <- lapply(seq_along(lambdas), function(i) {
    lam  <- lambdas[i]
    data <- boxcox_transform(ev, lam)
    ggplot2::ggplot(data.frame(x = data), ggplot2::aes(x = x)) +
      ggplot2::geom_histogram(fill = "steelblue", colour = "white",
                              alpha = 0.85, bins = 20L) +
      ggplot2::labs(title = sprintf("\u03bb = %.2f", lam),
                    x = "Transformed entropy", y = "Count") +
      ggplot2::theme_bw(base_size = 9) +
      ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold",
                                                               hjust = 0.5),
                     panel.grid.minor = ggplot2::element_blank())
  })
  print(
    patchwork::wrap_plots(plots, ncol = 3L) +
      patchwork::plot_annotation(
        title = sprintf("Box-Cox transformation \u2014 %s", dataset_label),
        theme = ggplot2::theme(plot.title = ggplot2::element_text(
          face = "bold", hjust = 0.5, size = 12)))
  )
}

plot_boxcox_grid(Entrop_Vec_GISAID_no0S, "GISAID (zeros removed)")
plot_boxcox_grid(Entrop_Vec_NCBI_no0S,   "NCBI (zeros and singletons removed)")
```

## Log1p

```{r transf-log1p}
GISAID_log1p_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S, transfr = scales::log1p_trans(),
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "GISAID \u2014 zeros removed", bins = 20
)

NCBI_log1p_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S, transfr = scales::log1p_trans(),
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "NCBI \u2014 zeros and singletons removed", bins = 20
)

save_transformed_comparison_table(
  GISAID_log1p_no0S,
  "Site selection \u2014 GISAID, zeros removed, Log1p",
  table_path("Log1p_GISAID_no0S.html"), important_sites_gisaid
)

save_transformed_comparison_table(
  NCBI_log1p_no0S,
  "Site selection \u2014 NCBI, zeros and singletons removed, Log1p",
  table_path("Log1p_NCBI_no0S.html"), important_sites_ncbi
)
```

## Log1p∘Log1p (Double Compression)

```{r transf-loglog}
GISAID_loglog_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S, transfr = loglog1p1p_trans,
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "GISAID \u2014 zeros removed", bins = 20
)

NCBI_loglog_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S, transfr = loglog1p1p_trans,
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "NCBI \u2014 zeros and singletons removed", bins = 20
)

save_transformed_comparison_table(
  GISAID_loglog_no0S,
  "Site selection \u2014 GISAID, zeros removed, Log1p\u2218Log1p",
  table_path("LogLog_GISAID_no0S.html"), important_sites_gisaid
)

save_transformed_comparison_table(
  NCBI_loglog_no0S,
  "Site selection \u2014 NCBI, zeros and singletons removed, Log1p\u2218Log1p",
  table_path("LogLog_NCBI_no0S.html"), important_sites_ncbi
)
```

## Square Root

```{r transf-sqrt}
GISAID_sqrt_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S, transfr = scales::sqrt_trans(),
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "GISAID \u2014 zeros removed", bins = 20
)

NCBI_sqrt_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S, transfr = scales::sqrt_trans(),
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "NCBI \u2014 zeros and singletons removed", bins = 20
)

save_transformed_comparison_table(
  GISAID_sqrt_no0S,
  "Site selection \u2014 GISAID, zeros removed, Square root",
  table_path("Sqrt_GISAID_no0S.html"), important_sites_gisaid
)

save_transformed_comparison_table(
  NCBI_sqrt_no0S,
  "Site selection \u2014 NCBI, zeros and singletons removed, Square root",
  table_path("Sqrt_NCBI_no0S.html"), important_sites_ncbi
)
```

## Reciprocal

```{r transf-reciprocal}
GISAID_recip_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S, transfr = reciprocal_trans,
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "GISAID \u2014 zeros removed", bins = 20
)

NCBI_recip_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S, transfr = reciprocal_trans,
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "NCBI \u2014 zeros and singletons removed", bins = 20
)

save_transformed_comparison_table(
  GISAID_recip_no0S,
  "Site selection \u2014 GISAID, zeros removed, Reciprocal",
  table_path("Reciprocal_GISAID_no0S.html"), important_sites_gisaid
)

save_transformed_comparison_table(
  NCBI_recip_no0S,
  "Site selection \u2014 NCBI, zeros and singletons removed, Reciprocal",
  table_path("Reciprocal_NCBI_no0S.html"), important_sites_ncbi
)
```

## Box-Cox (λ = −0.05)

```{r transf-boxcox}
bc_trans <- boxcox_trans(lambda = -0.05)

GISAID_bc_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S, transfr = bc_trans,
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "GISAID \u2014 zeros removed", bins = 20
)

NCBI_bc_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S, transfr = bc_trans,
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "NCBI \u2014 zeros and singletons removed", bins = 20
)

save_transformed_comparison_table(
  GISAID_bc_no0S,
  "Site selection \u2014 GISAID, zeros removed, Box-Cox (\u03bb = \u22120.05)",
  table_path("BoxCox_GISAID_no0S.html"), important_sites_gisaid
)

save_transformed_comparison_table(
  NCBI_bc_no0S,
  "Site selection \u2014 NCBI, zeros and singletons removed, Box-Cox (\u03bb = \u22120.05)",
  table_path("BoxCox_NCBI_no0S.html"), important_sites_ncbi
)
```

## Log (natural log)

```{r transf-log}
GISAID_log_no0S <- gmm_site_selection(
  Entrop_Vec_GISAID_no0S, transfr = scales::log_trans(),
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "GISAID \u2014 zeros removed", bins = 20
)

NCBI_log_no0S <- gmm_site_selection(
  Entrop_Vec_NCBI_no0S, transfr = scales::log_trans(),
  prob = 0.9, hist = TRUE, calc_gmm_orig = FALSE, calc_gmm_transfr = TRUE,
  dataset_label = "NCBI \u2014 zeros and singletons removed", bins = 20
)

save_transformed_comparison_table(
  GISAID_log_no0S,
  "Site selection \u2014 GISAID, zeros removed, Log",
  table_path("Log_GISAID_no0S.html"), important_sites_gisaid
)

save_transformed_comparison_table(
  NCBI_log_no0S,
  "Site selection \u2014 NCBI, zeros and singletons removed, Log",
  table_path("Log_NCBI_no0S.html"), important_sites_ncbi
)
```

---

---

# Summary: % Overlap Across All Transformations

The tables below consolidate the percentage of VOC reference sites recovered
by each GMM subset across all transformations.  Cell colour intensity is
proportional to overlap — deeper green indicates higher recovery.  The maximum
value in each transformation column is shown in bold.

## GISAID summary table

```{r summary-gisaid}
gisaid_results <- list(
  "Untransformed"    = GISAID_gmm_no0S,
  "Log1p"            = GISAID_log1p_no0S,
  "Log1p\u2218Log1p" = GISAID_loglog_no0S,
  "Square root"      = GISAID_sqrt_no0S,
  "Reciprocal"       = GISAID_recip_no0S,
  "Box-Cox"          = GISAID_bc_no0S,
  "Log"              = GISAID_log_no0S
)

tbl_summary_gisaid <- generate_summary_table(
  results_list    = gisaid_results,
  important_sites = important_sites_gisaid,
  table_caption   = "Summary: % overlap with GISAID VOC reference sites (Alpha / Beta / Gamma / Delta / Omicron) by GMM subset and transformation",
  save_path       = table_path("Summary_GISAID.html")
)
tbl_summary_gisaid
```

## NCBI summary table

```{r summary-ncbi}
ncbi_results <- list(
  "Untransformed"    = NCBI_gmm_no0S,
  "Log1p"            = NCBI_log1p_no0S,
  "Log1p\u2218Log1p" = NCBI_loglog_no0S,
  "Square root"      = NCBI_sqrt_no0S,
  "Reciprocal"       = NCBI_recip_no0S,
  "Box-Cox"          = NCBI_bc_no0S,
  "Log"              = NCBI_log_no0S
)

tbl_summary_ncbi <- generate_summary_table(
  results_list    = ncbi_results,
  important_sites = important_sites_ncbi,
  table_caption   = "Summary: % overlap with NCBI VOC reference sites (Alpha / Beta / Gamma / Delta) by GMM subset and transformation",
  save_path       = table_path("Summary_NCBI.html")
)
tbl_summary_ncbi
```

# Session Information

```{r session-info}
sessionInfo()
```

# References