---
title: "Unsupervised Recovery of SARS-CoV-2 Variant Structure via Entropy-Driven Site Selection and PAM Clustering: Precision, Recall, and F1 Evaluation Across Wild-Type and Delta-Dominated Surveillance Periods"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: flatly
    highlight: tango
bibliography: references_accuracy.bib
nocite: '@*'
vignette: >
  %\VignetteIndexEntry{Unsupervised Recovery of SARS-CoV-2 Variant Structure via Entropy-Driven Site Selection and PAM Clustering: Precision, Recall, and F1 Evaluation Across Wild-Type and Delta-Dominated Surveillance Periods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 9,
  fig.height = 6
)
```

# Introduction

Entropy-based site selection is only useful if the sites it identifies carry
genuine biological signal. This vignette provides a rigorous empirical test:
we take real SARS-CoV-2 Spike protein sequences from two well-separated
surveillance windows --- **May--June 2020**, when the ancestral wild-type was
circulating in the United States, and **July--August 2021**, when the Delta
variant (*B.1.617.2*) was dominant --- and ask whether the sites selected by
`cluster_sites_by_entropy()` are sufficient to recover the correct temporal
structure when the two groups are clustered together.

Several terms are used throughout this vignette and are defined here for
clarity.

## Definitions

### Variants of Concern/Interest/Under Monitoring

During the COVID-19 pandemic, the World Health Organization (WHO) established
a variant classification framework to prioritise public health responses based 
on the epidemiological and clinical significance of emerging SARS-CoV-2 lineages.
Three principal designations were used in practice.

A \textbf{Variant of Concern (VOC)} is a lineage for which there is evidence of 
one or more of the following: increased transmissibility or a detrimental change 
in COVID-19 epidemiology; increased virulence or change in clinical disease 
presentation; or decreased effectiveness of public health measures, diagnostics,
vaccines, or therapeutics. Five lineages were designated as VOCs during the 
pandemic: Alpha (B.1.1.7), Beta (B.1.351), Gamma (P.1), Delta (B.1.617.2), 
and Omicron (B.1.1.529).

A \textbf{Variant of Interest (VOI)} is a lineage with genetic changes that are
predicted or known to affect virus characteristics such as transmissibility, 
disease severity, immune escape, or diagnostic and therapeutic performance, and
which has been identified to cause significant community transmission or 
multiple COVID-19 clusters in multiple countries. VOIs were monitored closely
but had not yet accumulated sufficient evidence for VOC designation. Examples,
at various points during the pandemic, included Lambda (C.37), Mu (B.1.621), 
Kappa (B.1.617.1), Iota (B.1.526), and Epsilon (B.1.427/B.1.429).

A \textbf{Variant Under Monitoring (VUM)} is a lineage carrying genetic changes
suspected to affect virus characteristics but for which evidence of phenotypic 
or epidemiological impact is limited, warranting enhanced surveillance without 
immediate escalation to VOI or VOC status. Variant classifications were dynamic
and subject to revision as new evidence became available.

### Single Nucleotide Polymorphism (SNPs)

A **single nucleotide polymorphism (SNP)** is formally defined as a single
nucleotide substitution in the genome that is present at an appreciable
frequency in a population, often operationally taken as at least 1%. In the 
context of SARS-CoV-2 genomic surveillance, however, the term “SNP” has 
frequently been used more loosely to refer to lineage-associated mutations, 
including those described at the amino acid level. In this work, we retain 
the term “SNP” for consistency with the surveillance literature, but use it 
interchangeably with the more precise notion of a defining mutation at a site, 
referring to amino acid substitutions in the Spike protein that characterize 
specific viral lineages. 

### Gaussian mixture models (GMM)

A **Gaussian mixture model (GMM)** is a probabilistic model that represents a 
given distribution as a weighted sum of Gaussian components; here it is applied 
to the vector of per-site Shannon entropies to identify sites with high
variability --- those sites form the discriminating feature set passed to the 
clustering step.

### Partitioning Around Medoids (PAM)

A partitioning around medoids (PAM) algorithm is a deterministic clustering
method that partitions observations into a fixed number of clusters by selecting
representative data points, called medoids, and assigning each observation to
the nearest medoid according to a specified dissimilarity measure. In this work,
PAM is applied to the reduced sequence representation derived from high-entropy 
sites, using the Gower distance to quantify pairwise dissimilarity between
sequences. Unlike centroid-based methods such as k-means, PAM is robust to noise
and accommodates mixed or non-Euclidean data, making it well suited to 
categorical sequence data.

## Accuracy Study Logic

The pipeline mirrors the approach used in the main detection study, but applied
retrospectively to a known ground truth. Because we know which sequences belong
to which period, we can evaluate clustering accuracy formally using **precision**,
**recall** and **F1 score**, and cross-reference the discriminating sites against 
the curated Delta SNP catalogue in `sarscov2_variants`. The dominance of the 
Delta variant in the United States during July--August 2021 is well established
from national surveillance data [@CDC2022].

The analysis proceeds in nine steps:

1. **Load and preprocess** the full NCBI Spike protein dataset, following the
   identical pipeline described in the *Preprocessing* vignette.
2. **Partition** US sequences into two 2-month time windows.
3. **Select sites** via GMM entropy clustering on the combined data, achieving
   dimensionality reduction from 1,273 sites to a compact discriminating
   subset.
4. **Cluster** the combined sequences using PAM with Gower distance, with the
   optimal number of clusters determined by Silhouette analysis.
5. **Visualise** the clustering in 2D via t-SNE, highlighting wild-type
   sequences (black crosses) and medoid sequences (transparent red diamonds).
6. **Evaluate** accuracy via precision, recall and F1 score across k = 2, 3, 4, 
   and 9.
7. **Characterise** each cluster via modal amino acid profiles, pairwise contrast 
   analysis identifying sites that differentiate the wild-type cluster from all 
   others, and annotating defining mutations. 
8. **Examine borderline sequences** at k = 4 by identifying sequences assigned to 
   wild-type cluster but falling in the visual boundary regions of the t-SNE 
   layout and comparing their mutational profiles to cluster medoids.
9. **Apply the pipeline at scale** to all ~109,536 available US sequences.

> **Positive class definition:** Period 1 (wild-type, May--June 2020) sequences
> correctly assigned to the cluster with the highest Period 1 purity.
> Precision, recall and F1 are computed for this positive class.

# Libraries

```{r libraries}
library(ViralEntropR)
library(dplyr)
library(ggplot2)
library(scales)       # pretty_breaks for silhouette plot
library(cluster)      # pam(), daisy()
library(Rtsne)        # t-SNE visualisation
library(future.apply) # parallel PAM silhouette loop
library(kableExtra)   # compact HTML tables
```

# Global Parameters

All user-facing parameters for steps 1---8 are gathered here. Adjust `COUNTRY`,
the two time windows, and `K_HIGHLIGHT` before re-rendering; no other chunk 
needs to change.

```{r parameters}
# ── User-modifiable parameters ───────────────────────────────────────────────

# Data
COUNTRY         <- "USA"          # filter to this country (NULL = all countries)
UNIQUE_SEQS     <- TRUE           # TRUE = deduplicate sequences before PAM
                                  # FALSE = keep all sequences (including duplicates)
N_SITES         <- 1273L          # number of Spike protein sites

# Time windows
PERIOD1_START   <- "2020-05-01"   # Period 1: wild-type phase
PERIOD1_END     <- "2020-07-01"
PERIOD2_START   <- "2021-07-01"   # Period 2: Delta-dominant phase
PERIOD2_END     <- "2021-09-01"

# PAM silhouette sweep
K_MAX           <- 40L                # maximum number of PAM clusters to evaluate
K_HIGHLIGHT     <- c(2L, 3L, 4L, 9L)  # PAM k values to plot in detail

# t-SNE parameters
TSNE_PERPLEXITY <- 10L
TSNE_SEED       <- 42L

# Parallelism (future.apply)
N_WORKERS       <- max(1L, parallel::detectCores() - 1L)
```

# Load Preprocessed Data

The preprocessing steps applied here follow the same pipeline described in 
detail in the *NCBI SARS-CoV-2 Spike Protein Sequence Preprocessing: From Raw FASTA to an Analysis-Ready Integer-Encoded Matrix* vignette. Readers are referred to that 
vignette for a full discussion of each step, including date extraction, 
ambiguous residue filtering, integer encoding, and data frame assembly.

The raw NCBI FASTA file is loaded and processed in a single pipeline. Dates are
extracted from sequence headers using a two-pass regex strategy: a full `yyyy-mm-dd` 
pass first, followed by a `yyyy-mm` pass to recover sequences with month-level
precision only. Sequences with unresolvable dates or missing country annotations
are removed. The remaining sequences are converted to a character matrix, filtered 
for ambiguous amino acid codes (B, J, X, Z), integer-encoded using the 25-character
`ViralEntropR` alphabet, and assembled into a data frame with integer site
columns `1:N_SITES` plus `Date` and `Country` columns. Rows are sorted by date
and the index is reset.

```{r load-data}
# Replace 'fasta_path' with the file path to your Zenodo download
fasta_path <- file.path("data-raw", "sequences.fasta")              # replace(!)
fasta      <- Biostrings::readAAStringSet(fasta_path)

# --- Dates ------------------------------------------------------------------
# Pass 1: identify sequences with year-only dates
dates_ymd <- extract_fasta_dates(
  fasta,
  custom_pattern = "(?<=\\|)[0-9]{4}-(0?[1-9]|1[0-2])-(0?[1-9]|[12][0-9]|3[01]|00)"
)

# Pass 2: extract yyyy-mm, remove year-only
dates_result <- extract_fasta_dates(
  fasta,
  custom_pattern = "(?<=\\|)[0-9]{4}-(0[1-9]|1[0-2])",
  date_format    = "%Y-%m"
)

if (!is.na(dates_result$missing_id[1]) && length(dates_result$missing_id) > 0) {
  fasta        <- fasta[-dates_result$missing_id]
  dates_result <- extract_fasta_dates(
    fasta,
    custom_pattern = "(?<=\\|)[0-9]{4}-(0[1-9]|1[0-2])",
    date_format    = "%Y-%m"
  )
}

# --- Countries --------------------------------------------------------------
countries_result <- extract_fasta_countries(fasta, position = 2)

if (!is.na(countries_result$missing_id[1]) && length(countries_result$missing_id) > 0) {
  fasta            <- fasta[-countries_result$missing_id]
  countries_result <- extract_fasta_countries(fasta, position = 2)
  dates_result     <- extract_fasta_dates(
    fasta,
    custom_pattern = "(?<=\\|)[0-9]{4}-(0[1-9]|1[0-2])",
    date_format    = "%Y-%m"
  )
}

# --- Convert, filter, encode ------------------------------------------------
char_mat       <- fasta_to_char_matrix(fasta)
filtered       <- filter_ambiguous_sequences(char_mat, option = 2)
clean_char_mat <- filtered$FilteredMatrix
deleted_rows   <- filtered$DeletedSeqId

corrected_dates <- dates_result$corrected_dates[-deleted_rows]
countries       <- countries_result$countries[-deleted_rows]

int_mat <- encode_aa_sequence(clean_char_mat)

# --- Assemble data frame ----------------------------------------------------
n_sites         <- ncol(int_mat)
AL_df           <- as.data.frame(int_mat)
colnames(AL_df) <- as.character(seq_len(n_sites))
AL_df[]         <- lapply(AL_df, as.integer)
AL_df$Date      <- as.Date(format(corrected_dates, "%Y-%m-01"))
AL_df$Country   <- countries
AL_df           <- AL_df[order(AL_df$Date), ]
rownames(AL_df) <- NULL

sprintf("Final data frame: %d sequences x %d sites", nrow(AL_df), n_sites)

```

## Country Filter

Restricting to a single country (`US`) reduces confounding from inter-country
sampling biases and surveillance heterogeneity. All downstream analyses use
`AL_df` after this filter.

```{r filter-country}
# Filter by country if specified
if (!is.null(COUNTRY)) {
  AL_df <- AL_df[!is.na(AL_df$Country) & AL_df$Country == COUNTRY, ]
  sprintf("Sequences after country filter (%s): %d", COUNTRY, nrow(AL_df))
}
```

# Step 1: Data Partitioning

## Partition into Two Time Windows (Wild-Type vs Delta)

We extract two non-overlapping 2-month windows: Period 1 covers the reference
wild-type phase and Period 2 covers the Delta-dominant phase.

```{r partition}
part1 <- partition_time_windows(
  data          = AL_df,
  n_sites       = N_SITES,
  window_length = 2L,
  window_type   = 2L,           # single 2-month window (start_date to end_date)
  start_date    = PERIOD1_START,
  end_date      = PERIOD1_END,
  verbose       = FALSE
)

part2 <- partition_time_windows(
  data          = AL_df,
  n_sites       = N_SITES,
  window_length = 2L,
  window_type   = 2L,
  start_date    = PERIOD2_START,
  end_date      = PERIOD2_END,
  verbose       = FALSE
)

df_p1 <- part1$Partitions[[1]]
df_p2 <- part2$Partitions[[1]]

sprintf("Period 1 (%s to %s): %d sequences",
        PERIOD1_START, PERIOD1_END, nrow(df_p1))
sprintf("Period 2 (%s to %s): %d sequences",
        PERIOD2_START, PERIOD2_END, nrow(df_p2))
```

## Per-Period Entropy Clustering

```{r per-period-clusters}
# Helper: extract highest-entropy sites from a cluster DataFrame
top_sites <- function(clust_df) {
  sort(clust_df[clust_df$class == max(clust_df$class), ]$sites)
}

# Period 1 top sites
sites_p1 <- top_sites(part1$Clusters[[1]]$DataFrame)

cat("Period 1 highest-entropy sites:\n")
print(sites_p1)

# Period 2 top sites
sites_p2 <- top_sites(part2$Clusters[[1]]$DataFrame)

cat("\nPeriod 2 highest-entropy sites:\n")
print(sites_p2)
```

> **Biological note — site 681 (furin cleavage site).** Site 681 is of
> exceptional epidemiological significance and its detection in the Period 1
> (wild-type) entropy profile is noteworthy. Position 681 sits immediately
> upstream of the RRAR polybasic furin cleavage site at the S1/S2 boundary,
> a structural feature that is unique to SARS-CoV-2 among closely related
> sarbecoviruses and was absent in SARS-CoV-1. The furin cleavage site enables
> the Spike protein to be pre-cleaved by furin-family proteases within the
> secretory pathway before virion release, substantially increasing the
> efficiency of cell entry and facilitating airway tropism. This feature is
> widely regarded as a key determinant of SARS-CoV-2's high transmissibility
> in humans [@Peacock2021].
> 
> Substitutions at position 681 subsequently became defining mutations of two
> of the most consequential VOCs: **P681H** (proline to histidine) in Alpha
> (B.1.1.7), and **P681R** (proline to arginine) in Delta (B.1.617.2). Both
> substitutions enhance furin cleavage efficiency, P681R more so than P681H, 
> by altering the charge and conformation of the cleavage loop, resulting
> in faster S1/S2 processing, more efficient membrane fusion, and increased
> competitive fitness. The entropy signal at site 681 in the wild-type period
> therefore reflects early, low-frequency variation at a position that was
> already under strong positive selection pressure [@Johnson2021].

# Step 2: Combine Both Partitions

## Analyze the Selected Sites

We combine both periods and run GMM entropy clustering on the union to identify
sites that discriminate the two time points. This achieves dimensionality
reduction from `r N_SITES` sites to a smaller informative subset.

```{r combined-gmm}
# Combine periods -- rows are sequences, columns are sites
df_combined           <- rbind(df_p1, df_p2)
rownames(df_combined) <- seq_len(nrow(df_combined))

# Period 1 size -- used throughout for labelling
n_p1 <- nrow(df_p1)
n_p2 <- nrow(df_p2)
n_total <- nrow(df_combined)

cat(sprintf("Combined: %d sequences (%d Period 1 + %d Period 2)\n",
            n_total, n_p1, n_p2))

# Entropy on site columns only
entrp_combined <- apply(df_combined[, seq_len(N_SITES), drop = FALSE],
                        2, calculate_entropy)

clust_combined <- cluster_sites_by_entropy(entrp_combined, nr = n_total)
clust_combined_df <- clust_combined$DataFrame

selected_sites <- sort(
  clust_combined_df[clust_combined_df$class == max(clust_combined_df$class), ]$sites
)

cat(sprintf("\nGMM selected %d sites from combined data:\n",
            length(selected_sites)))
print(selected_sites)
```

## Cross-Reference with Known Variant Sites

```{r cross-reference}
# ── Helper: pull SNP catalogue for a named variant ──────────────────────────
get_variant_snps <- function(label) {
  idx   <- which(unlist(sarscov2_variants$WHO_Label) == label)
  sites <- sarscov2_variants$Defining_SNP_Sites[[idx]]
  snps  <- sarscov2_variants$Defining_SNPs[[idx]]
  list(label = label, sites = sites, snps = snps)
}

# Retrieve catalogues for the four VOCs of interest
var_delta <- get_variant_snps("Delta")
var_alpha <- get_variant_snps("Alpha")
var_beta  <- get_variant_snps("Beta")
var_gamma <- get_variant_snps("Gamma")

# Convenience aliases (used throughout downstream chunks)
delta_sites <- var_delta$sites;  delta_snps <- var_delta$snps
alpha_sites <- var_alpha$sites;  alpha_snps <- var_alpha$snps
beta_sites  <- var_beta$sites;   beta_snps  <- var_beta$snps
gamma_sites <- var_gamma$sites;  gamma_snps <- var_gamma$snps

# ── Print SNP tables ─────────────────────────────────────────────────────────
for (v in list(var_delta, var_alpha, var_beta, var_gamma)) {
  cat(sprintf("\n%s defining SNP sites:\n", v$label))
  print(data.frame(SNP = v$snps, Site = v$sites))
}

# ── Overlap with GMM-selected sites ─────────────────────────────────────────
overlap_list <- lapply(list(var_delta, var_alpha, var_beta, var_gamma),
                       function(v) intersect(selected_sites, v$sites))
names(overlap_list) <- c("Delta", "Alpha", "Beta", "Gamma")

# Keep Delta overlap as scalar for backward compatibility
overlap <- overlap_list[["Delta"]]

# voc_list defined here so it is available to all downstream chunks
# (delta-flag, pam-and-plots, mode-profile, contrast-gmm)
voc_list <- list(Delta = var_delta, Alpha = var_alpha,
                 Beta  = var_beta,  Gamma = var_gamma)

cat("\nGMM-selected sites overlapping with VOC defining SNP sites:\n")
for (nm in names(overlap_list)) {
  ov <- overlap_list[[nm]]
  cat(sprintf("  %-6s: %s\n",
              nm,
              if (length(ov) > 0) paste(ov, collapse = ", ") else "none"))
}

# ── Note on shared mutations across VOCs ─────────────────────────────────────
# Identify sites present in the SNP catalogues of more than one VOC
all_voc_sites <- lapply(voc_list, function(v) v$sites)
site_freq     <- table(unlist(all_voc_sites))
shared_sites  <- as.integer(names(site_freq[site_freq > 1L]))

cat("\nSites shared by more than one VOC catalogue:\n")
for (s in sort(shared_sites)) {
  vocs_with_site <- names(voc_list)[
    vapply(voc_list, function(v) s %in% v$sites, logical(1L))
  ]
  # Collect the actual SNP names at this site for each VOC
  snp_labels <- vapply(vocs_with_site, function(vn) {
    v   <- voc_list[[vn]]
    idx <- which(v$sites == s)
    paste(v$snps[idx], collapse = "/")
  }, character(1L))
  cat(sprintf("  Site %-5d: %s\n", s,
              paste(sprintf("%s (%s)", vocs_with_site, snp_labels),
                    collapse = ", ")))
}
```

> **Note on the combined dataset.** The union of the two time windows spans
> the wild-type phase through the peak of Delta dominance, and the
> GMM-selected sites capture mutation signatures from *all* major VOCs
> circulating during this period. Several sites carry mutations that recur
> independently across multiple variants, a hallmark of convergent evolution
> under strong positive selection. For example, site 501 (N501Y) is shared
> by Alpha, Beta and Gamma --- all three independently acquired the same
> asparagine-to-tyrosine substitution because it enhances ACE2 receptor
> binding affinity. Similarly, site 484 (E484K in Beta and Gamma, E484Q in
> Delta/Kappa) is a key immune evasion site; the different amino acid outcomes
> (K vs Q) at the same position across variants illustrate that while the
> *site* is under selection, the *specific substitution* depends on the
> genetic background. Site 614 (D614G) is present across virtually all VOCs
> and was the first globally dominant mutation, increasing Spike
> conformational stability and transmissibility [@Korber2020].

# Step 3: PAM Clustering with Gower Distance

Partitioning Around Medoids (PAM) is a robust clustering algorithm that
minimises the sum of pairwise dissimilarities between each sequence and the
medoid of its assigned cluster [@Kaufman1990]. Unlike K-Means, PAM medoids are
actual observed sequences rather than computed centroids, making them directly
interpretable as representative viral haplotypes.

## Prepare Feature Matrix

```{r prepare-pam}
# Subset to selected sites and convert to factors (required for Gower distance)
AL_Cat_sub <- df_combined[, as.character(selected_sites), drop = FALSE]
AL_Cat_sub[] <- lapply(AL_Cat_sub, as.integer)

if (UNIQUE_SEQS) {
  # Deduplicate -- record original row indices for labelling
  AL_Cat_fac    <- AL_Cat_sub %>% mutate_if(is.integer, as.factor)
  unique_idx    <- which(!duplicated(AL_Cat_fac))
  AL_Cat_fac    <- AL_Cat_fac[unique_idx, , drop = FALSE]
  orig_period   <- ifelse(unique_idx <= n_p1, 1L, 2L)
  cat(sprintf("Unique sequences: %d (from %d total)\n",
              nrow(AL_Cat_fac), n_total))
} else {
  AL_Cat_fac  <- AL_Cat_sub %>% mutate_if(is.integer, as.factor)
  orig_period <- ifelse(seq_len(n_total) <= n_p1, 1L, 2L)
  cat(sprintf("All sequences retained: %d\n", nrow(AL_Cat_fac)))
}

n_p1_eff <- sum(orig_period == 1L)
n_p2_eff <- sum(orig_period == 2L)
cat(sprintf("Period 1 sequences in PAM input: %d\n", n_p1_eff))
cat(sprintf("Period 2 sequences in PAM input: %d\n", n_p2_eff))
```

## Gower Distance Matrix

Gower distance between two sequences $\mathbf{x}$ and $\mathbf{y}$ is defined
as $d_G(\mathbf{x}, \mathbf{y}) = \frac{1}{p} \sum_{j=1}^{p} \delta_j \, s_j$,
where $p$ is the number of sites, $\delta_j \in \{0, 1\}$ indicates whether
site $j$ is comparable for the given pair, and $s_j$ is a site-specific
dissimilarity score that equals $0$ when $x_j = y_j$ and $1$ otherwise for
categorical variables such as amino acid codes [@Gower1971].

```{r gower, results='hide', message=FALSE}
gower_dist <- daisy(AL_Cat_fac, metric = "gower")
```

## Silhouette Analysis

The silhouette width for sequence $i$ is defined as
$s(i) = \frac{b(i) - a(i)}{\max\{a(i),\, b(i)\}}$, where $a(i)$ is the mean
Gower distance from $i$ to all other sequences in its own cluster and $b(i)$
is the mean Gower distance from $i$ to all sequences in the nearest
neighbouring cluster; $s(i) \in [-1, 1]$, with values close to $1$ indicating
a well-separated assignment and values close to $0$ or negative indicating
overlap or potential misclassification [@Rousseeuw1987].

We sweep k from 2 to `K_MAX` in parallel to find the optimal number of
clusters.

```{r silhouette, cache=TRUE}
future::plan(future::multisession, workers = N_WORKERS)

sil_width <- c(NA, future_sapply(2:K_MAX, function(k) {
  pam(gower_dist, diss = TRUE, k = k)$silinfo$avg.width
}))

future::plan(future::sequential)

# Silhouette plot -- grid lines, x-ticks every 5
# Drop k=1 (undefined, NA) before plotting to avoid missing-value warnings.
# na.rm = TRUE on geom calls provides an additional safety net.
sil_df <- data.frame(k = seq_len(K_MAX), width = sil_width)
sil_df <- sil_df[!is.na(sil_df$width), ]   # removes k=1 row

ggplot(sil_df, aes(x = k, y = width)) +
  geom_line(colour = "black", na.rm = TRUE) +
  geom_point(colour = "darkorange", shape = 1, size = 2, na.rm = TRUE) +
  geom_vline(xintercept = K_HIGHLIGHT, linetype = "dashed",
             colour = "steelblue", alpha = 0.6) +
  scale_x_continuous(limits       = c(2L, K_MAX),
                     breaks        = seq(5, K_MAX, by = 5),
                     minor_breaks  = seq(2, K_MAX, by = 1)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
  labs(title    = "Silhouette Analysis: Optimal Number of Clusters",
       subtitle  = sprintf("Blue dashed lines mark k = %s (highlighted solutions)",
                           paste(K_HIGHLIGHT, collapse = ", ")),
       x = "Number of clusters",
       y = "Average Silhouette Width") +
  theme_bw() +
  theme(panel.grid.major = element_line(colour = "grey85"),
        panel.grid.minor = element_line(colour = "grey93"),
        plot.title       = element_text(hjust = 0.5),
        plot.subtitle    = element_text(hjust = 0.5, size = 9),
        axis.title.x     = element_text(colour = "steelblue"),
        axis.title.y     = element_text(colour = "darkorange"))

optimal_k <- which.max(sil_width)
cat(sprintf("Optimal k by silhouette: %d  (width = %.4f)\n",
            optimal_k, max(sil_width, na.rm = TRUE)))
```

# Step 4: PAM Fitting and t-SNE Visualisation

## t-SNE Embedding

t-distributed Stochastic Neighbor Embedding (t-SNE) is a non-linear
dimensionality reduction method that maps high-dimensional pairwise distances
to a 2D layout by preserving local neighborhood structure [@vanderMaaten2008].
Distances in the t-SNE plot are not interpretable in absolute terms --- only
the relative proximity of points within and between clusters carries meaning
--- and the same Gower distance matrix is used as input for both PAM and t-SNE
to ensure consistency between the clustering and its visualization.

To reiterate, t-SNE is run once on the Gower distance matrix and the resulting 
2D coordinates are reused for all k values. This ensures that spatial positions
are identical across plots so that differences in cluster boundaries are directly 
comparable.

Each plot uses three visual layers:

- **Coloured shapes** -- PAM cluster assignment.
- **Black cross overlay** -- Period 1 (wild-type) sequences.
- **Filled red diamond** -- medoid of each cluster: unlike K-Means centroids,
  PAM medoids are actual observed sequences drawn directly from the data,
  making them biologically interpretable centers of their cluster.

```{r tsne, cache=TRUE}
set.seed(TSNE_SEED)
tsne_obj <- Rtsne(gower_dist, is_distance = TRUE,
                  perplexity = TSNE_PERPLEXITY)

tsne_base <- tsne_obj$Y %>%
  as.data.frame() %>%
  setNames(c("X", "Y")) %>%
  mutate(period = orig_period)
```

## VOC SNP Matching Helpers

The following helpers are used throughout Steps 4--8 to count and name VOC
defining SNP matches in individual sequences. `count_snp_matches()` returns
an integer count; `get_matched_snps()` returns the names of matched SNPs.
These functions operate only on the `selected_sites` subset, so overlap counts
refer to defining SNPs that were also selected by the entropy GMM.

```{r delta-flag}
# ── Helper: count how many of a variant's SNPs are present in a sequence ────
# seq_int: named integer vector (site name -> encoded amino acid value)
# voc:     list with $sites and $snps (from get_variant_snps())
# Returns: integer count of matching SNP alleles
count_snp_matches <- function(seq_int, voc) {
  # Only consider SNP sites that are present in the selected_sites subset
  overlap_sites <- intersect(as.character(voc$sites), names(seq_int))
  if (length(overlap_sites) == 0L) return(0L)

  expected_codes <- vapply(as.integer(overlap_sites), function(s) {
    snp_idx <- which(voc$sites == s)
    aa_char <- substring(voc$snps[snp_idx], nchar(voc$snps[snp_idx]))
    as.integer(encode_aa_sequence(matrix(aa_char))[1L, 1L])
  }, integer(1L))
  names(expected_codes) <- overlap_sites

  sum(vapply(overlap_sites, function(s) {
    isTRUE(seq_int[[s]] == expected_codes[[s]])
  }, logical(1L)))
}

# ── Apply to all Period 1 sequences ─────────────────────────────────────────
# p1_rows_int: row indices into AL_Cat_int that correspond to Period 1
p1_rows_int <- which(orig_period == 1L)

# Build integer version of the deduplicated feature matrix for display/mode calc
AL_Cat_int <- AL_Cat_fac %>%
  mutate_if(is.factor, function(x) as.integer(as.character(x)))

# Convert to named integer matrix for fast row-wise access
AL_Cat_int_mat <- as.matrix(AL_Cat_int)
colnames(AL_Cat_int_mat) <- colnames(AL_Cat_int)

# voc_list <- list(Delta = var_delta, Alpha = var_alpha,
#                  Beta  = var_beta,  Gamma = var_gamma)

# match_counts: matrix [n_p1_int x 4] -- SNP match count per Period 1 sequence
match_counts <- do.call(cbind, lapply(voc_list, function(voc) {
  vapply(p1_rows_int, function(r) {
    count_snp_matches(AL_Cat_int_mat[r, ], voc)
  }, integer(1L))
}))
rownames(match_counts) <- p1_rows_int

# ── Helper: return names of matched SNPs (not just count) ───────────────────
get_matched_snps <- function(seq_int, voc) {
  overlap_sites <- intersect(as.character(voc$sites), names(seq_int))
  if (length(overlap_sites) == 0L) return(character(0L))
  vapply(overlap_sites, function(s) {
    snp_idx   <- which(voc$sites == as.integer(s))
    aa_char   <- substring(voc$snps[snp_idx], nchar(voc$snps[snp_idx]))
    exp_code  <- as.integer(encode_aa_sequence(matrix(aa_char))[1L, 1L])
    if (isTRUE(seq_int[[s]] == exp_code)) voc$snps[snp_idx] else NA_character_
  }, character(1L)) |> (\(x) x[!is.na(x)])()
}

# ── Summary table with SNP detail ───────────────────────────────────────────
cat("Period 1 sequences by number of VOC SNP matches\n")
cat("(only SNPs at GMM-selected sites are considered)\n\n")

for (voc_nm in names(voc_list)) {
  voc         <- voc_list[[voc_nm]]
  n_overlap   <- length(intersect(as.character(voc$sites),
                                  as.character(selected_sites)))
  counts      <- match_counts[, voc_nm]
  cat(sprintf("%s  (%d defining SNPs, %d overlap with selected sites):\n",
              voc_nm, length(voc$sites), n_overlap))
  if (n_overlap == 0L) {
    cat("  No overlap with GMM-selected sites -- skip.\n\n")
    next
  }
  for (m in seq(n_overlap, 0L)) {
    seq_idx <- which(counts == m)
    if (length(seq_idx) == 0L && m < n_overlap) next
    cat(sprintf("  %d/%d SNP match(es): %d sequence(s)",
                m, n_overlap, length(seq_idx)))
    if (m > 0L && length(seq_idx) > 0L) {
      # Collect matched SNP names for each sequence at this tier
      snp_sets <- lapply(p1_rows_int[seq_idx], function(r) {
        get_matched_snps(AL_Cat_int_mat[r, ], voc)
      })
      # Unique SNP patterns
      snp_patterns <- sort(unique(vapply(snp_sets,
                                         paste, character(1L), collapse = "+")))
      cat(sprintf("  [SNP pattern(s): %s]", paste(snp_patterns, collapse = " | ")))
    }
    cat("\n")
  }
  cat("\n")
}

```

## t-SNE Plot Function

A reusable helper that renders the t-SNE embedding coloured by PAM cluster
assignment. Three visual layers are overlaid: cluster-coloured points,
black crosses for Period 1 sequences, and red diamonds for cluster medoids.

```{r plot-function}
# Reusable plot function: takes PAM clustering vector and medoid row indices,
# produces a t-SNE plot with three visual layers:
#   1. Coloured shapes by cluster
#   2. Black cross overlay for Period 1 (wild-type) sequences
#   3. Filled red diamond for each cluster medoid
plot_tsne <- function(pam_clust, k, medoid_rows,
                      tsne_df    = tsne_base,
                      period_vec = orig_period,
                      xlim_fixed = NULL,
                      ylim_fixed = NULL) {

  df <- tsne_df %>%
    mutate(cluster   = factor(pam_clust),
           is_p1     = (period_vec == 1L),
           is_medoid = row_number() %in% medoid_rows)

  p <- ggplot(df, aes(x = X, y = Y, colour = cluster, shape = cluster)) +
    geom_point(size = 2, alpha = 0.7) +
    scale_shape_manual(values = seq(0, k)) +
    # Black cross overlay for all Period 1 sequences
    geom_point(data         = subset(df, is_p1),
               aes(x = X, y = Y),
               colour       = "black", shape = 4, size = 2,
               inherit.aes  = FALSE) +
    # Semi-transparent red diamond: alpha reveals the medoid is a real data point.
    geom_point(data         = subset(df, is_medoid),
               aes(x = X, y = Y),
               colour       = "red", shape = 18, size = 5,
               alpha        = 0.45,
               inherit.aes  = FALSE) +
    labs(title    = sprintf("PAM k = %d  |  t-SNE visualisation", k),
         subtitle = "Crosses = Period 1 (wild-type);  Red diamonds = cluster medoids",
         x = "t-SNE dimension 1",
         y = "t-SNE dimension 2") +
    theme_bw() +
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, size = 9))

  # Apply fixed axis ranges if supplied -- ensures identical plot area across
  # all k values regardless of legend size differences
  if (!is.null(xlim_fixed) && !is.null(ylim_fixed))
    p <- p + coord_cartesian(xlim = xlim_fixed, ylim = ylim_fixed)

  print(p)
  invisible(p)
}
```

## PAM and Medoid Analysis

```{r pam-and-plots, results="asis"}
# Helper: render kableExtra table in results='asis' chunks inside loops
kt <- function(tbl) cat(as.character(tbl), "\n")
# Fit PAM for all highlighted k values
pam_fits <- lapply(K_HIGHLIGHT, function(k) {
  pam(gower_dist, diss = TRUE, k = k)
})
names(pam_fits) <- paste0("k", K_HIGHLIGHT)

# Precompute fixed axis ranges from the common t-SNE coordinates so that all
# plots have an identical data area regardless of legend size differences.
x_pad    <- diff(range(tsne_base$X)) * 0.05
y_pad    <- diff(range(tsne_base$Y)) * 0.05
xlim_fix <- range(tsne_base$X) + c(-x_pad, x_pad)
ylim_fix <- range(tsne_base$Y) + c(-y_pad, y_pad)

for (k in K_HIGHLIGHT) {
  fit         <- pam_fits[[paste0("k", k)]]
  medoid_rows <- fit$id.med

  # ── Subsection heading ──────────────────────────────────────────────────
  cat(sprintf("\n## PAM k = %d\n\n### t-SNE Plot\n\n", k))

  # ── Plot ─────────────────────────────────────────────────────────────────
  plot_tsne(fit$clustering, k, medoid_rows,
            xlim_fixed = xlim_fix, ylim_fixed = ylim_fix)

  # ── Medoid sequences kable ───────────────────────────────────────────────
  cat(sprintf("\n### Medoid Sequences (k = %d)\n\n", k))
  medoid_aa <- decode_aa_sequence(
    as.matrix(AL_Cat_int[medoid_rows, , drop = FALSE])
  )
  medoid_df           <- as.data.frame(medoid_aa)
  colnames(medoid_df) <- as.character(selected_sites)
  rownames(medoid_df) <- paste0("cl_", seq_len(k))

  # Red cell_spec wherever cl_i (i > 1) differs from cl_1 at that site.
  # Direct [row, col_name] access -- avoids any as.character(df[row,]) pitfalls.
  if (k > 1L) {
    for (ri in 2L:k) {
      for (nm in colnames(medoid_df)) {
        v1 <- medoid_df[1L, nm]; vi <- medoid_df[ri, nm]
        if (vi != v1)
          medoid_df[ri, nm] <- cell_spec(vi, format = "html",
                                         color = "red", bold = TRUE)
      }
    }
  }

  kt(
    kbl(medoid_df, format = "html", escape = FALSE,
        caption = sprintf("Medoid sequences (k = %d) | red = differs from cl_1", k)) |>
      kable_styling(bootstrap_options = c("condensed", "hover"),
                    font_size = 11, full_width = FALSE) |>
      row_spec(1L, background = "#d6eaf8")
  )

  # ── VOC SNP content kable ────────────────────────────────────────────────
  cat(sprintf("\n### VOC SNP Content of Medoids (k = %d)\n\n", k))

  voc_tbl <- do.call(rbind, lapply(seq_len(k), function(cl_i) {
    row_i <- medoid_rows[cl_i]
    vapply(names(voc_list), function(voc_nm) {
      voc   <- voc_list[[voc_nm]]
      n_ov  <- length(intersect(as.character(voc$sites),
                                as.character(selected_sites)))
      n_m   <- count_snp_matches(AL_Cat_int_mat[row_i, ], voc)
      sprintf("%d / %d", n_m, n_ov)
    }, character(1L))
  }))
  voc_tbl           <- as.data.frame(voc_tbl)
  rownames(voc_tbl) <- paste0("cl_", seq_len(k))

  kt(
    kbl(voc_tbl, format = "html",
        caption = sprintf("VOC SNP matches per medoid (matched / overlap sites), k = %d", k)) |>
      kable_styling(bootstrap_options = c("condensed", "hover"),
                    font_size = 11, full_width = FALSE)
  )
  cat("\n")
}
```

# Step 5: Modal Amino Acid Profile per Cluster

For each k value, we compute the most frequent (modal) amino acid at each
GMM-selected site within each cluster. This reveals the characteristic sequence
signature of each temporal group and allows direct comparison with the medoid.
The mode matrices computed here are also used by the contrast analysis in Step 7.

```{r mode-profile, results="asis"}
get_mode <- function(x) { ux <- unique(x); ux[which.max(tabulate(match(x, ux)))] }

# Store mode matrices for Step 7 contrast analysis.
mode_results <- list()

for (k in K_HIGHLIGHT) {
  fit         <- pam_fits[[paste0("k", k)]]
  clust_vec   <- fit$clustering
  medoid_rows <- fit$id.med
  k_levels    <- sort(unique(clust_vec))

  mode_mat <- do.call(rbind, lapply(k_levels, function(cl) {
    rows <- which(clust_vec == cl)
    vapply(colnames(AL_Cat_int), function(s) get_mode(AL_Cat_int[rows, s]),
           integer(1L))
  }))
  rownames(mode_mat) <- paste0("cl_", k_levels)
  colnames(mode_mat) <- colnames(AL_Cat_int)

  mode_aa           <- decode_aa_sequence(as.matrix(mode_mat))
  rownames(mode_aa) <- paste0("cl_", k_levels)
  colnames(mode_aa) <- as.character(selected_sites)

  medoid_int          <- as.matrix(AL_Cat_int[medoid_rows, , drop = FALSE])
  medoid_aa           <- decode_aa_sequence(medoid_int)
  rownames(medoid_aa) <- paste0("med_", k_levels)
  colnames(medoid_aa) <- as.character(selected_sites)

  mode_results[[paste0("k", k)]] <- list(
    mode_int  = mode_mat, k_levels = k_levels, clust_vec = clust_vec)

  same_flag <- all(mode_mat == medoid_int)
  obs_note  <- if (same_flag)
    sprintf("**Mode = Medoid for all clusters (k = %d): tight cohesion confirmed.**", k)
  else {
    diff_s <- which(apply(mode_mat != medoid_int, 2, any))
    sprintf("**Sites where mode differs from medoid (k = %d): %s**",
            k, paste(selected_sites[diff_s], collapse = ", "))
  }

  cat(sprintf("\n### k = %d\n\n%s\n\n", k, obs_note))

  # Build combined df with cell_spec red highlighting where mode != medoid
  combined_df <- as.data.frame(rbind(mode_aa, medoid_aa), stringsAsFactors = FALSE)
  n_cl <- length(k_levels)
  for (col_i in seq_along(selected_sites)) {
    col_nm <- as.character(selected_sites[col_i])
    for (ci in seq_along(k_levels)) {
      mv <- mode_aa[ci, col_i]; dv <- medoid_aa[ci, col_i]
      if (mv != dv) {
        combined_df[ci,        col_nm] <- cell_spec(mv, color="red", bold=TRUE)
        combined_df[n_cl + ci, col_nm] <- cell_spec(dv, color="red", bold=TRUE)
      }
    }
  }

  kt(
    kbl(combined_df, format="html", escape=FALSE,
        caption=sprintf("Mode (cl_* = blue) / Medoid (med_* = yellow) | red = differs  [k=%d]", k)) |>
      kable_styling(bootstrap_options=c("condensed","hover"), font_size=11, full_width=FALSE) |>
      row_spec(seq_len(n_cl),           background="#d6eaf8") |>
      row_spec(seq(n_cl+1L, n_cl*2L),   background="#fef9e7")
  )
  cat("\n")
}
```

> **Observation.** Across k = 2, 3 and 4, the modal amino acid at every
> selected site within each cluster coincides exactly with the PAM medoid
> sequence. This agreement confirms that each cluster is internally
> homogeneous: the most centrally located sequence in Gower space is also the
> most frequent sequence at every position, which is precisely what one expects
> from a viral population dominated by a single emergent haplotype. At k = 9
> the correspondence is no longer exact at every site, reflecting the finer
> subdivision of sequence space into smaller, less internally uniform clusters.
> The overall mutational profile of each medoid nonetheless remains intact,
> indicating that the core variant signal is preserved even as the clustering
> resolution increases.

# Step 6: Precision, Recall and F1 Score

For each k value tested (2, 3, 4 and 9), we compute classification metrics
treating Period 1 sequences as the positive class. Since PAM cluster labels are
arbitrary integers, we first identify the cluster with the highest Period 1
purity and designate that as the predicted positive label.

```{r metrics}
compute_metrics <- function(clust_vec, period_vec, n_p1_eff) {
  p1_idx <- which(period_vec == 1L)
  cl_purity <- tapply(clust_vec[p1_idx], clust_vec[p1_idx], length)
  positive_cluster <- as.integer(names(which.max(cl_purity)))
  pred_positive <- clust_vec == positive_cluster
  true_positive <- period_vec == 1L
  TP <- sum( pred_positive &  true_positive)
  FP <- sum( pred_positive & !true_positive)
  FN <- sum(!pred_positive &  true_positive)
  TN <- sum(!pred_positive & !true_positive)
  precision <- if ((TP + FP) > 0) TP / (TP + FP) else NA_real_
  recall    <- if ((TP + FN) > 0) TP / (TP + FN) else NA_real_
  f1        <- if (!is.na(precision) && !is.na(recall) && (precision + recall) > 0)
               2 * precision * recall / (precision + recall) else NA_real_
  list(k = length(unique(clust_vec)), positive_cluster = positive_cluster,
       TP = TP, FP = FP, FN = FN, TN = TN,
       precision = round(precision, 4), recall = round(recall, 4),
       f1 = round(f1, 4))
}

metrics_list <- lapply(K_HIGHLIGHT, function(k) {
  compute_metrics(pam_fits[[paste0("k", k)]]$clustering, orig_period, n_p1_eff)
})
metrics_df <- do.call(rbind, lapply(metrics_list, as.data.frame))
print(metrics_df)
```

> **On the interplay between statistical optimality and biological knowledge.**
> The silhouette analysis provides a useful unsupervised criterion for cluster
> compactness and separation. It correctly identifies k = 2 as the most
> statistically compact solution. However, k = 2 conflates all non-ancestral
> variants into a single cluster, obscuring the variant-level structure that
> is biologically meaningful. As k increases to 3 and 4, F1 scores improve
> and individual VOC-associated clusters become identifiable, which is confirmed
> visually through t-SNE, quantitatively through the medoid SNP content
> tables, and formally through precision and recall. Practitioners working
> with biological sequence data should treat silhouette analysis as a starting
> point, not a definitive criterion: statistical optimality and biological
> interpretability need not coincide. Selecting k requires integrating three 
> complementary sources of evidence: the silhouette curve, visual inspection of 
> the t-SNE layout, and domain knowledge. No single criterion is sufficient on 
> its own.

# Step 7: Site Differentiation — Wild-Type Cluster vs Later Clusters

Using the mode matrices computed in Step 5, we ask: at which Spike sites does
the modal amino acid in each non-wild-type cluster differ from that of the
wild-type cluster?

A site flagged across multiple k values and multiple non-wild-type clusters is
a robust, reproducible differentiator. The ranked frequency table below counts
how many of the `r sum(vapply(K_HIGHLIGHT, function(k) length(unique(pam_fits[[paste0("k",k)]]$clustering))-1L, integer(1L)))` total (k, cluster) comparisons flag each site,
and annotates each site with its VOC SNP membership where applicable [@Korber2020;@CDC2022].

## Annotation helper

```{r site-annotation-helper}
# annotate_site_full(s): compact VOC annotation for Spike site s.
# Uses ALL 12 variants in sarscov2_variants (not just the 4 in voc_list).
# Distinguishes:
#   "VOC-SNP"  = site is a defining SNP for that VOC (Defining_SNP_Sites)
#   "VOC"      = site is mutated in that VOC but is NOT a defining SNP
#   ""         = not found in any VOC catalogue
# Examples: 452 -> "Alpha/Delta/Epsilon/Iota/Kappa-SNP"
#            95 -> "Delta/Iota/Kappa/Omicron"
#           190 -> "Gamma-SNP; Iota"
annotate_site_full <- function(s) {
  labels    <- unlist(sarscov2_variants$WHO_Label)
  snp_sites <- sarscov2_variants$Defining_SNP_Sites
  mut_sites <- sarscov2_variants$Mutation_Sites
  snp_vocs  <- labels[vapply(snp_sites, function(ss) s %in% ss, logical(1L))]
  mut_only  <- setdiff(labels[vapply(mut_sites, function(ms) s %in% ms, logical(1L))],
                       snp_vocs)
  parts <- c(
    if (length(snp_vocs) > 0L) paste0(paste(snp_vocs, collapse = "/"), "-SNP"),
    if (length(mut_only)  > 0L) paste(mut_only, collapse = "/"))
  if (length(parts) == 0L) "" else paste(parts, collapse = "; ")
}
```

## Contrasts calculation

```{r contrast-gmm, results="asis"}
max_possible <- sum(vapply(K_HIGHLIGHT, function(k) {
  length(unique(pam_fits[[paste0("k", k)]]$clustering)) - 1L
}, integer(1L)))

site_diff_counts <- integer(length(selected_sites))
names(site_diff_counts) <- as.character(selected_sites)
contrast_rows <- list()

for (k in K_HIGHLIGHT) {
  mr        <- mode_results[[paste0("k", k)]]
  clust_vec <- mr$clust_vec
  k_levels  <- mr$k_levels
  mode_int  <- mr$mode_int

  p1_idx <- which(orig_period == 1L)
  purity <- tapply(clust_vec[p1_idx], clust_vec[p1_idx], length)
  wt_cl  <- as.integer(names(which.max(purity)))
  wt_row <- which(k_levels == wt_cl)

  for (cl in k_levels[k_levels != wt_cl]) {
    cl_row        <- which(k_levels == cl)
    diff_vec      <- mode_int[cl_row, ] != mode_int[wt_row, ]
    diff_sites_cl <- selected_sites[diff_vec]

    site_diff_counts[as.character(diff_sites_cl)] <-
      site_diff_counts[as.character(diff_sites_cl)] + 1L

    voc_str <- if (length(diff_sites_cl) > 0L)
      vapply(diff_sites_cl, annotate_site_full, character(1L))
    else character(0L)

    contrast_rows[[length(contrast_rows)+1L]] <- data.frame(
      k=k, wt=wt_cl, vs=cl, n=length(diff_sites_cl),
      sites = if (length(diff_sites_cl)>0L) paste(diff_sites_cl, collapse=", ") else "-",
      VOC   = if (length(diff_sites_cl)>0L)
        paste(sprintf("%s[%s]", diff_sites_cl, voc_str), collapse=", ") else "-",
      stringsAsFactors=FALSE)
  }
}

# Ranked frequency table
freq_df <- data.frame(
  Site = as.integer(names(site_diff_counts)),
  Freq = as.integer(site_diff_counts),
  Pct  = round(100 * site_diff_counts / max_possible, 1),
  stringsAsFactors = FALSE)
freq_df <- freq_df[freq_df$Freq > 0L, ]
if (nrow(freq_df) > 0L) {
  freq_df <- freq_df[order(-freq_df$Freq), ]
  freq_df$VOC <- vapply(freq_df$Site, annotate_site_full, character(1L))
  cat(sprintf("\n**Site differentiation across all %d (k, cluster) comparisons:**\n\n",
              max_possible))
  kt(
    kbl(freq_df, format="html",
        col.names=c("Site","Freq",sprintf("%% of %d comparisons", max_possible),"VOC"),
        caption="Sites ranked by differentiation frequency (wild-type vs other clusters)") |>
      kable_styling(bootstrap_options=c("condensed","hover","striped"),
                    font_size=11, full_width=FALSE) |>
      column_spec(3,
        color = ifelse(freq_df$Pct >= 50, "darkred", "black"),
        bold  = freq_df$Pct >= 50)
  )
  cat("\n")
} else {
  cat("\n*No differentiating sites found (all clusters share the same modal residues).*\n\n")
}

# Per-k breakdown
if (length(contrast_rows) > 0L) {
  contrast_df <- do.call(rbind, contrast_rows)
  cat("\n**Per-(k, cluster) breakdown:**\n\n")
  kt(
    kbl(contrast_df, format="html",
        caption="Differentiating sites: wild-type cluster vs each other cluster") |>
      kable_styling(bootstrap_options=c("condensed","hover"), font_size=11, full_width=FALSE)
  )
}
```

## Biological Interpretation of Key Differentiating Sites

> The tables above reveal sites that consistently differentiate the wild-type
> cluster from all others across multiple k values. Beyond the expected
> VOC-defining SNP sites --- D614G (site 614, present in virtually all VOCs
> [@Korber2020]), sites 452, 478, 501 and 681 --- the analysis flags two
> clusters of NTD residues whose significance is not immediately apparent from
> VOC catalogues alone.
>
> **Site 190 -- NTD allosteric control node.** Residue N188 and nearby NTD
> residues including position 190 show large state-dependent interaction energy
> changes with S2 residues that regulate RBD opening, indicating an allosteric
> control node modulating the closed-to-open transition of the Spike trimer
> [@Wrobel2020]. Substitutions here can alter the open/closed equilibrium,
> influencing ACE2 engagement and antibody accessibility without residing
> within the receptor-binding motif itself.
>
> **Sites 141--158 -- NTD antigenic supersite and immune-escape module.**
> Residues 140--150 form the N-terminal half of the NTD antigenic supersite
> loop, while residues 152, 153, 155, 157 and 158 form a single exposed loop
> functioning as an immune-escape module, heavily targeted by NTD-directed
> neutralising antibodies [@McCallum2021]. Deletions and substitutions in
> 141--158 arise repeatedly under immune pressure, remodel the supersite
> conformation, and confer resistance to NTD-directed antibodies while leaving
> ACE2 binding largely intact [@Liu2021NTD]. Their recurrent selection across
> Alpha, Beta, Gamma and Delta is a hallmark of convergent immune-escape
> evolution. Crucially, these sites were identified here by entropy alone ---
> no prior annotation was used.

# Step 8: Borderline Period 1 Sequences

PAM assigns each sequence to its nearest medoid by Gower distance, so no
wild-type-cluster sequence can be strictly closer to a foreign medoid in Gower
space --- that is the definition of cluster membership. However, the t-SNE
projection reveals that some cluster-1 sequences sit visually close to the
territory of neighbouring clusters in 2D. These sequences are not misclassified
in Gower space, but their position in the t-SNE layout suggests they occupy an
intermediate region of sequence space, making them biologically interesting
candidates for closer inspection. We identify them by hardcoded coordinate
regions derived from direct visual inspection of the k = 4 t-SNE plot.

```{r borderline, results="asis"}
# Borderline analysis for k=4 only. Boundary regions from visual inspection:
#   Region A: X > 15  AND  Y <= 0          (right border, near cluster 2)
#   Region B: X > -10 AND  X < 10  AND  Y < -10  (center bottom border, clusters 3/4)
# Applies to ALL cluster-1 sequences regardless of time period.

k         <- 4L
fit       <- pam_fits[[paste0("k", k)]]
clust_vec <- fit$clustering
med_rows  <- fit$id.med

wt_cl       <- 1L
foreign_cls <- setdiff(sort(unique(clust_vec)), wt_cl)
wt_rows     <- which(clust_vec == wt_cl)

cat(sprintf("\n### k = %d  (wild-type = cluster %d)\n\n", k, wt_cl))

# Apply coordinate filter
region_A <- tsne_base$X[wt_rows] > 15  & tsne_base$Y[wt_rows] <= 0
region_B <- tsne_base$X[wt_rows] > -10 & tsne_base$X[wt_rows] < 10 &
            tsne_base$Y[wt_rows] < -10
is_bl   <- region_A | region_B
bl_rows <- wt_rows[is_bl]

cat(sprintf("%d wt-cluster sequences fall in the boundary t-SNE region(s).\n\n",
            length(bl_rows)))

if (length(bl_rows) == 0L) {
  cat("*No sequences in the defined boundary regions.*\n\n")
} else {

  # t-SNE distances to own and nearest foreign medoid
  wt_med_x <- tsne_base$X[med_rows[wt_cl]]
  wt_med_y <- tsne_base$Y[med_rows[wt_cl]]
  d_own    <- sqrt((tsne_base$X[bl_rows] - wt_med_x)^2 +
                   (tsne_base$Y[bl_rows] - wt_med_y)^2)

  fgn_dist_mat <- vapply(foreign_cls, function(cl) {
    fx <- tsne_base$X[med_rows[cl]]; fy <- tsne_base$Y[med_rows[cl]]
    sqrt((tsne_base$X[bl_rows] - fx)^2 + (tsne_base$Y[bl_rows] - fy)^2)
  }, numeric(length(bl_rows)))

  d_fgn  <- if (length(foreign_cls) == 1L) as.numeric(fgn_dist_mat)
             else apply(fgn_dist_mat, 1L, min)
  margin <- d_own - d_fgn

  period_lbl <- ifelse(orig_period[bl_rows] == 1L, "P1", "P2")

  dist_tbl <- data.frame(
    global_idx = bl_rows,
    period     = period_lbl,
    tsne_X     = round(tsne_base$X[bl_rows], 1),
    tsne_Y     = round(tsne_base$Y[bl_rows], 1),
    d_own      = round(d_own,  2),
    d_fgn      = round(d_fgn, 2),
    margin     = round(margin, 2)
  )
  dist_tbl <- dist_tbl[order(dist_tbl$margin), ]

  kt(
    kbl(dist_tbl, format = "html",
        col.names = c("Row idx", "Period", "t-SNE X", "t-SNE Y",
                      "d(own med)", "d(nearest fgn med)", "Margin"),
        caption = sprintf(
          "Boundary-region wt-cluster sequences, sorted by margin [k=%d, wt=cl%d]",
          k, wt_cl)) |>
      kable_styling(bootstrap_options = c("condensed","hover"),
                    font_size = 11, full_width = FALSE) |>
      column_spec(7L,
                  color = ifelse(dist_tbl$margin < 0, "red", "black"),
                  bold  = dist_tbl$margin < 0)
  )
  cat("\n")

  # VOC SNP content table
  n_ref         <- 1L + length(foreign_cls)
  all_rows_bl   <- c(med_rows[wt_cl], med_rows[foreign_cls], dist_tbl$global_idx)
  row_labels_bl <- c(
    sprintf("wt_med (cl%d)", wt_cl),
    sprintf("fgn_med (cl%d)", foreign_cls),
    sprintf("border[row%d] %s", dist_tbl$global_idx, dist_tbl$period)
  )
  voc_bl <- do.call(rbind, lapply(all_rows_bl, function(r) {
    vapply(names(voc_list), function(voc_nm) {
      voc  <- voc_list[[voc_nm]]
      n_ov <- length(intersect(as.character(voc$sites),
                               as.character(selected_sites)))
      n_m  <- count_snp_matches(AL_Cat_int_mat[r, ], voc)
      sprintf("%d / %d", n_m, n_ov)
    }, character(1L))
  }))
  voc_bl           <- as.data.frame(voc_bl)
  rownames(voc_bl) <- row_labels_bl
  n_bl             <- nrow(voc_bl) - n_ref

  tbl_voc <- kbl(voc_bl, format = "html",
      caption = sprintf(
        "VOC SNP content: wt medoid | foreign medoids | borderline sequences [k=%d]",
        k)) |>
    kable_styling(bootstrap_options = c("condensed","hover"),
                  font_size = 11, full_width = FALSE) |>
    row_spec(1L,             background = "#d6eaf8") |>
    row_spec(seq(2L, n_ref), background = "#fef9e7")
  if (n_bl > 0L)
    tbl_voc <- row_spec(tbl_voc, seq(n_ref + 1L, nrow(voc_bl)),
                        background = "#fce4ec")
  kt(tbl_voc)
  cat("\n")

  # Sequence comparison table
  own_med_aa <- as.data.frame(decode_aa_sequence(
    as.matrix(AL_Cat_int[med_rows[wt_cl], , drop = FALSE])))
  foreign_aa <- do.call(rbind, lapply(foreign_cls, function(cl)
    as.data.frame(decode_aa_sequence(
      as.matrix(AL_Cat_int[med_rows[cl], , drop = FALSE])))))
  border_aa  <- as.data.frame(decode_aa_sequence(
    as.matrix(AL_Cat_int[bl_rows, , drop = FALSE])))

  compare_df <- rbind(own_med_aa, foreign_aa, border_aa)
  colnames(compare_df) <- as.character(selected_sites)
  rownames(compare_df) <- c(
    sprintf("wt_med (cl%d)", wt_cl),
    sprintf("fgn_med (cl%d)", foreign_cls),
    sprintf("border[row%d] %s margin=%.2f",
            dist_tbl$global_idx, dist_tbl$period, dist_tbl$margin)
  )

  out_df <- compare_df
  for (ri in 2L:nrow(compare_df)) {
    for (nm in colnames(compare_df)) {
      v1 <- compare_df[1L, nm]; vi <- compare_df[ri, nm]
      if (!is.na(vi) && vi != v1)
        out_df[ri, nm] <- cell_spec(vi, format = "html", color = "red", bold = TRUE)
    }
  }
  kt(
    kbl(out_df, format = "html", escape = FALSE,
        caption = sprintf(
          "wt medoid (blue) | foreign medoids (yellow) | borderline (pink) | red = differs from wt medoid  [k=%d]",
          k)) |>
      kable_styling(bootstrap_options = c("condensed","hover"),
                    font_size = 11, full_width = FALSE) |>
      row_spec(1L,             background = "#d6eaf8") |>
      row_spec(seq(2L, n_ref), background = "#fef9e7") |>
      row_spec(seq(n_ref + 1L, nrow(out_df)), background = "#fce4ec")
  )
  cat("\n")
}
```

# Step 9: Full-Dataset Clustering (All 109,536 Sequences)

The preceding analysis was conducted on a focused two-period subset to enable
supervised evaluation against a known ground truth. Here we apply the same
entropy-GMM-PAM pipeline to the **complete preprocessed dataset** (`AL_df`,
all available USA sequences across all time periods) to demonstrate that the
algorithm achieves robust cluster separation at scale, without restricting to
a specific surveillance window.

## Site Selection on Full Dataset

```{r full-gmm, cache=TRUE}
# Entropy on all N_SITES sites across all sequences in AL_df
entrp_full <- apply(AL_df[, seq_len(N_SITES), drop = FALSE], 2L, calculate_entropy)

clust_full    <- cluster_sites_by_entropy(entrp_full, nr = nrow(AL_df))
clust_full_rl <- relabel_entropy_classes(clust_full$DataFrame)

selected_sites_full <- sort(
  clust_full_rl[clust_full_rl$class == 1L, ]$sites
)

cat(sprintf("Full-dataset GMM selected %d sites from %d sequences:\n",
            length(selected_sites_full), nrow(AL_df)))
print(selected_sites_full)

# Overlap with VOC defining SNP sites -- uses same get_variant_snps() as Step 2.
# The difference vs. the two-period selected_sites is expected: entropy was
# computed on 109k sequences with different frequency distributions, so GMM
# selects a different (potentially overlapping) site subset.
# Delta two-period: 5 overlap sites (19,452,478,681,950).
# Delta full-dataset overlap depends on selected_sites_full computed above.
overlap_full <- lapply(voc_list, function(v)
  intersect(selected_sites_full, v$sites))
names(overlap_full) <- names(voc_list)

cat("\nGMM-selected sites overlapping with VOC defining SNP sites (full dataset):\n")
for (nm in names(overlap_full)) {
  ov <- overlap_full[[nm]]
  cat(sprintf("  %-6s: %s\n", nm,
              if (length(ov) > 0L) paste(ov, collapse = ", ") else "none"))
}
```

## Gower Distance

```{r full-gower, cache=TRUE}
# Subset to selected sites and convert to factors
AL_full_fac <- AL_df[, as.character(selected_sites_full), drop = FALSE]
AL_full_fac[] <- lapply(AL_full_fac, function(x) as.factor(as.integer(x)))

# Deduplicate before Gower: 109k sequences produce an infeasible (for laptop RAM)
# distance matrix (~48 GB). Unique sequences are sufficient for PAM/t-SNE and
# dramatically reduce computation while preserving the full sequence diversity.
full_unique_idx  <- which(!duplicated(AL_full_fac))
AL_full_fac_uniq <- AL_full_fac[full_unique_idx, , drop = FALSE]

cat(sprintf("Unique sequences for full-dataset clustering: %d (from %d total)\n",
            nrow(AL_full_fac_uniq), nrow(AL_full_fac)))

gower_dist_full <- daisy(AL_full_fac_uniq, metric = "gower")
```

## Silhouette Analysis

```{r full-silhouette, cache=TRUE}
future::plan(future::multisession, workers = N_WORKERS)

sil_width_full <- c(NA, future_sapply(2:K_MAX, function(k) {
  pam(gower_dist_full, diss = TRUE, k = k)$silinfo$avg.width
}))

future::plan(future::sequential)

sil_df_full <- data.frame(k = seq_len(K_MAX), width = sil_width_full)
sil_df_full <- sil_df_full[!is.na(sil_df_full$width), ]

ggplot(sil_df_full, aes(x = k, y = width)) +
  geom_line(colour = "black") +
  geom_point(colour = "darkorange", shape = 1, size = 2) +
  geom_vline(xintercept = c(6L, 7L, 9L), linetype = "dashed", colour = "steelblue", alpha = 0.6) +
  scale_x_continuous(limits = c(2L, K_MAX),
                     breaks = seq(5, K_MAX, by = 5),
                     minor_breaks = seq(2, K_MAX, by = 1)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6)) +
  labs(title    = "Silhouette Analysis: Full Dataset (All Sequences)",
       subtitle = "Blue dashed lines mark k = 6, 7, 9",
       x = "Number of clusters", y = "Average Silhouette Width") +
  theme_bw() +
  theme(panel.grid.major  = element_line(colour = "grey85"),
        panel.grid.minor  = element_line(colour = "grey93"),
        plot.title        = element_text(hjust = 0.5),
        plot.subtitle     = element_text(hjust = 0.5, size = 9),
        axis.title.x      = element_text(colour = "steelblue"),
        axis.title.y      = element_text(colour = "darkorange"))

optimal_k_full <- which.max(sil_width_full)
cat(sprintf("Optimal k by silhouette (full dataset): %d  (width = %.4f)\n",
            optimal_k_full, max(sil_width_full, na.rm = TRUE)))
```

## PAM Fitting

We run t-SNE once on the full Gower distance matrix and recolour by each PAM
solution (k = 2, 3 and 7). Fixed axis ranges ensure spatial positions are
identical across all plots.

```{r full-pam, cache=TRUE}
K_HIGHLIGHT_FULL <- c(6L, 7L, 9L)

pam_fits_full <- lapply(K_HIGHLIGHT_FULL, function(k)
  pam(gower_dist_full, diss = TRUE, k = k))
names(pam_fits_full) <- paste0("k", K_HIGHLIGHT_FULL)

for (k in K_HIGHLIGHT_FULL) {
  cat(sprintf("k=%d cluster sizes: ", k))
  cat(paste(table(pam_fits_full[[paste0("k",k)]]$clustering), collapse=" | "), "\n")
}
```

## t-SNE Visualisation

```{r full-tsne, cache=TRUE}
set.seed(TSNE_SEED)
tsne_full_obj <- Rtsne(gower_dist_full, is_distance = TRUE,
                       perplexity = TSNE_PERPLEXITY)

tsne_full_base <- as.data.frame(tsne_full_obj$Y) |>
  setNames(c("X", "Y"))
```

```{r full-int-mat}
# Integer matrix for medoid decoding and VOC matching (built once, reused)
AL_full_int     <- AL_full_fac_uniq |>
  mutate_if(is.factor, function(x) as.integer(as.character(x)))
AL_full_int_mat <- as.matrix(AL_full_int)

# VOC list for full dataset (Alpha, Beta, Delta, Gamma only)
voc_list_full <- list(
  Delta = get_variant_snps("Delta"),
  Alpha = get_variant_snps("Alpha"),
  Beta  = get_variant_snps("Beta"),
  Gamma = get_variant_snps("Gamma")
)

# Fixed axis ranges for identical plot area across all k values
x_pad_f    <- diff(range(tsne_full_base$X)) * 0.05
y_pad_f    <- diff(range(tsne_full_base$Y)) * 0.05
xlim_fix_f <- range(tsne_full_base$X) + c(-x_pad_f,  x_pad_f)
ylim_fix_f <- range(tsne_full_base$Y) + c(-y_pad_f,  y_pad_f)
```

```{r full-plots, results="asis"}
for (k in K_HIGHLIGHT_FULL) {
  fit_f      <- pam_fits_full[[paste0("k", k)]]
  med_f      <- fit_f$id.med
  cl_sizes_f <- as.integer(table(fit_f$clustering))

  # ── t-SNE plot ─────────────────────────────────────────────────────────────
  cat(sprintf("\n## Full Dataset PAM k = %d\n\n### t-SNE Plot\n\n", k))

  df_f <- tsne_full_base |>
    mutate(cluster   = factor(fit_f$clustering),
           is_medoid = row_number() %in% med_f)

  p_f <- ggplot(df_f, aes(x = X, y = Y, colour = cluster, shape = cluster)) +
    geom_point(size = 1.5, alpha = 0.5) +
    scale_shape_manual(values = seq(0, k)) +
    geom_point(data        = subset(df_f, is_medoid),
               aes(x = X, y = Y),
               colour      = "red", shape = 18, size = 5, alpha = 0.45,
               inherit.aes = FALSE) +
    coord_cartesian(xlim = xlim_fix_f, ylim = ylim_fix_f) +
    labs(title    = sprintf("PAM k = %d | Full Dataset t-SNE", k),
         subtitle = sprintf("Unique sequences: %d | Red diamonds = cluster medoids",
                            nrow(AL_full_fac_uniq)),
         x = "t-SNE dimension 1", y = "t-SNE dimension 2") +
    theme_bw() +
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, size = 9))
  print(p_f)

  # ── Medoid sequences table ──────────────────────────────────────────────────
  cat(sprintf("\n### Medoid Sequences (k = %d)\n\n", k))

  med_aa <- decode_aa_sequence(as.matrix(AL_full_int[med_f, , drop = FALSE]))
  med_df <- as.data.frame(med_aa)
  colnames(med_df) <- as.character(selected_sites_full)
  rownames(med_df) <- paste0("cl_", seq_len(k))
  med_df <- cbind(N = cl_sizes_f, med_df)

  if (k > 1L) {
    for (ri in 2L:k) {
      for (nm in as.character(selected_sites_full)) {
        v1 <- med_df[1L, nm]; vi <- med_df[ri, nm]
        if (!is.na(vi) && vi != v1)
          med_df[ri, nm] <- cell_spec(vi, format = "html", color = "red", bold = TRUE)
      }
    }
  }
  kt(
    kbl(med_df, format = "html", escape = FALSE,
        caption = sprintf(
          "Medoid sequences (k=%d, full dataset) | N = cluster size | red = differs from cl_1",
          k)) |>
      kable_styling(bootstrap_options = c("condensed","hover"),
                    font_size = 11, full_width = FALSE) |>
      row_spec(1L, background = "#d6eaf8")
  )

  # ── VOC SNP content table ───────────────────────────────────────────────────
  cat(sprintf("\n### VOC SNP Content (k = %d)\n\n", k))

  voc_tbl_f <- do.call(rbind, lapply(seq_len(k), function(cl_i) {
    row_i <- med_f[cl_i]
    vapply(names(voc_list_full), function(voc_nm) {
      voc  <- voc_list_full[[voc_nm]]
      n_ov <- length(intersect(as.character(voc$sites),
                               as.character(selected_sites_full)))
      n_m  <- count_snp_matches(AL_full_int_mat[row_i, ], voc)
      sprintf("%d / %d", n_m, n_ov)
    }, character(1L))
  }))
  voc_tbl_f           <- as.data.frame(voc_tbl_f)
  rownames(voc_tbl_f) <- paste0("cl_", seq_len(k))

  kt(
    kbl(voc_tbl_f, format = "html",
        caption = sprintf(
          "VOC SNP matches per medoid (matched / overlap sites), full dataset k=%d",
          k)) |>
      kable_styling(bootstrap_options = c("condensed","hover"),
                    font_size = 11, full_width = FALSE)
  )
  cat("\n")
}
```

# Summary (Clustering Accuracy)

```{r summary}
cat("== Analysis Summary ==\n\n")
cat(sprintf("Country filter:         %s\n",
            ifelse(is.null(COUNTRY), "none (all countries)", COUNTRY)))
cat(sprintf("Unique sequences:       %s\n", UNIQUE_SEQS))
cat(sprintf("Period 1 sequences:     %d  (%s to %s)\n",
            n_p1_eff, PERIOD1_START, PERIOD1_END))
cat(sprintf("Period 2 sequences:     %d  (%s to %s)\n",
            n_p2_eff, PERIOD2_START, PERIOD2_END))
cat(sprintf("GMM-selected sites:     %d  [%s]\n",
            length(selected_sites),
            paste(selected_sites, collapse = ", ")))
cat("\nVOC SNP overlap with GMM-selected sites:\n")
for (nm in names(overlap_list)) {
  ov <- overlap_list[[nm]]
  cat(sprintf("  %-6s: %s\n",
              nm,
              if (length(ov) > 0) paste(ov, collapse = ", ") else "none"))
}
cat(sprintf("\nOptimal k by silhouette:   %d\n", optimal_k))
cat(sprintf("Optimal k by F1 score:     %d\n\n",
            K_HIGHLIGHT[which.max(metrics_df$f1)]))
cat("Classification metrics (k = 2, 3, 4, 9):\n")
print(metrics_df[, c("k", "positive_cluster", "TP", "FP", "FN",
                      "precision", "recall", "f1")])
```

# Conclusions

## Clustering Accuracy Study (Steps 1--8)

`ViralEntropR`'s entropy-driven dimensionality reduction, combined with PAM
clustering on Gower distance, recovers the temporal structure of SARS-CoV-2
evolution without any supervised learning or prior knowledge of mutation
topology.

**Perfect recall across all solutions.** All PAM solutions (k = 2, 3 and 4)
achieve recall = 1.00: every wild-type sequence (Period 1, May--June 2020) is
assigned to the wild-type cluster. No ancestral sequence is assigned to the
Delta-dominant partition despite the method being entirely unsupervised.

**F1 score peaks at k = 9, consistent with known biology.** While k = 2
achieves the highest silhouette width, k = 9 yields the best classification
performance and is the biologically justified solution. During the combined
surveillance window, the United States was host not only to the ancestral
wild-type and the four designated VOCs (Alpha, Beta, Gamma and Delta) but also
to several co-circulating Variants of Interest --- Kappa, Iota, Epsilon and
Lambda --- each carrying distinct subsets of Spike mutations. The pairwise
contrast analysis in Step 7 recovers mutation signatures consistent with
several of these lineages, suggesting that the nine-cluster solution captures
genuine sub-variant structure.

**The method is purely information-theoretic.** No feature engineering beyond
Shannon entropy scoring is applied, and the spatial topology of sites along the
Spike protein is entirely ignored. Sites are selected because their amino acid
distribution changed between surveillance periods --- as Shannon's foundational
framework predicts, entropy naturally captures the informative positions
[@Shannon1948].

**Mode equals medoid: cluster cohesion confirmed.** For k = 2, 3 and 4, the 
modal amino acid at every selected site within each cluster matches the PAM
medoid sequence exactly, confirming that these clusters are not statistical 
artefacts but reflect genuine population bottlenecks imposed by emergent 
variants. At k = 9, a small number of sites show minor discrepancies
between the mode and the medoid, as expected when clusters become smaller and
the within-cluster frequency distribution at some positions is more diffuse.
Critically, these discrepancies do not alter the VOC SNP classification of any
medoid: the defining mutation signatures remain intact, and the affected sites
carry non-defining substitutions that do not change the lineage assignment.

**Unsupervised recovery of functionally annotated sites.** Steps 7 and 8 used
no biological annotation. Yet the analysis independently flagged the NTD
allosteric control node near residue 190 [@Wrobel2020] and the antigenic
supersite spanning positions 141--158 [@McCallum2021;@Liu2021NTD] --- both
implicated in immune evasion and RBD conformational regulation. This
convergence with independently established structural biology validates that
the entropy pipeline captures genuine evolutionary signal.

## Full-Dataset Clustering: Global Separation Patterns (Step 9)

The full-dataset analysis (Step 9) is conceptually distinct from the 
accuracy study. It operates on all ~109,536 available USA sequences across 
the full surveillance timeline, with GMM site selection 
(`selected_sites_full`) performed on the entire dataset. The selected sites 
therefore reflect population-wide entropy rather than the contrast between 
two specific windows, and the two site sets are expected to differ — this 
is by design. Prior to Gower distance computation and PAM clustering, 
sequences are deduplicated and only unique haplotypes are retained. This 
step is computationally necessary: the full 109,536-sequence distance matrix 
would require approximately 48 GB of RAM, which exceeds the capacity of a 
typical workstation. Deduplication reduces the working set to a tractable 
number of unique sequences while preserving the full haplotype diversity of 
the dataset, since PAM medoids and t-SNE coordinates are determined by 
sequence identity rather than sequence count.

Across all values of k tried, the t-SNE plots show visible separation among
clusters. At k = 9, the VOC SNP content of the medoid sequences confirms clear
separation of all four major VOCs circulating at the cut-off time of the
dataset: Alpha, Beta, Delta and Gamma each have a dedicated cluster whose
medoid carries the corresponding defining mutations. This demonstrates that the
entropy-GMM-PAM pipeline is capable of recovering the known variant structure of 
the pandemic without any labelled outcomes or restriction to time windows.

`ViralEntropR` provides a transparent, reproducible, and computationally
efficient framework for detecting and characterizing viral variant structure
in large sequence datasets, combining rigorous information theory with the
interpretability required for epidemiological application.The results confirm 
that Shannon entropy, computed without any prior biological annotation, is 
sufficient to select a compact site subset that recovers known variant structure
at both the two-period and full-dataset scales.

# References
