---
title: "Variant Simulation and Analysis: Step by Step"
author: "Package Maintainer"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Variant Simulation and Analysis: Step by Step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

# Introduction

This vignette provides a comprehensive, step-by-step walkthrough of the
`ViralEntroR` pipeline. We reproduce the exact benchmark scenario from the
study design:

- Simulate a realistic viral evolutionary timeline with four competing variants
  (2020–2021).
- Partition the data into non-overlapping 2-month time windows.
- Calculate Hellinger distance matrices to quantify distributional shifts.
- Detect evolutionary change points using three complementary ECP approaches.

# Setup

```{r}
library(ViralEntropR)
library(dplyr)
library(lubridate)
library(ecp)
library(knitr)
library(kableExtra)
```

# Reference Sequence

```{r}
ref_sequence <- "MKTIIALSYIFCLVFADYKDDDDK"
sprintf("There are %d sites in the reference sequence", nchar(ref_sequence))
```

# Simulation Function: `simulate_variant_evolution`

|       **Parameter**         |                                 **Description**                                  | **Default** |
|-----------------------------|----------------------------------------------------------------------------------|-------------|
| `ref_sequences`             | data.frame which must include a `Date` column; OR a single sequence string.      | —           |
| `n_ref_months`              | number of months for the reference phase                                         | 3           |
| `start_date`                | simulation start (character or Date)                                             | —           |
| `end_date`                  | simulation end (character or Date)                                               | —           |
| `variants_config`           | numeric vector, e.g., `c(2,6)`: Variant1 gets 2 mutations; Variant2 gets 6.     | —           |
| `variant_intervals`         | gap (in months) between the emergence of variants                                | —           |
| `n_new_mutations`           | number of de novo sequences produced at a variant's emergence                    | 1           |
| `mutation_rate`             | multiplier for production in subsequent months (may be specified as a vector)    | 2           |
| `mutation_rate_variability` | variability (±fraction) around `mutation_rate`                                   | 0.25        |
| `deleterious_rate`          | probability that any new mutation is deleterious                                 | 0           |
| `n_deleterious_limit`       | max sequences allowed to carry a deleterious mutation                            | 1           |
| `n_sequences_total`         | total sequences to generate over the mutation phase (spread evenly)              | 140         |
| `ref_variability`           | if `TRUE`, add variability to the reference sequences at sites 1 and 2          | FALSE       |
| `n_ref_sequences`           | if `ref_sequences` is a single string, number of rows to create                  | 100         |
| `prob_deletion_event`       | probability each period contains a block of deleterious mutations                | 0           |
| `n_rows_to_delete`          | number of sequences affected if `prob_deletion_event` fires                      | 0           |
| `seed`                      | optional random seed                                                             | NULL        |

## Simulate Data

In this simulation, `simulate_variant_evolution()` models the stepwise 
emergence and competition of four viral variants over a two-year period from 
`2020-01-01` through `2021-12-01`. We begin with a 4-month reference phase
(`n_ref_months = 4`) in which 200 total reference sequences are sampled 
(50 per month, `n_ref_sequences = 200`), with variability introduced at
sites 1 and 2 (`ref_variability = TRUE`). The mutation phase generates 1,000 
sequences (`n_sequences_total = 1000`) at 50 per month. Four variants emerge at
months 5, 9, 14, and 19, carrying 2, 3, 4, and 2 mutations respectively 
(`variants_config = c(2,3,4,2)`), with 2 de novo sequences at each emergence 
(`n_new_mutations = 2`) and growth multipliers of 1.25 each. Although 
`deleterious_rate = 0`, each month has a 20% chance (`prob_deletion_event = 0.2`)
of injecting a block of 3 deleterious sequences at a random site
(`n_rows_to_delete = 3`). `seed = 2025` ensures reproducibility.

Variants compete in a "pairwise" fashion, always pitting the newest active 
variant against the prior dominant strain (or the original reference). During 
months before any variant appears, all sequences are drawn from the reference 
pool. Once Variant 1 emerges, each month's batch is composed of its de novo 
copies plus sampled reference sequences; when Variant 2 appears it competes 
directly with Variant 1, and so on. Sampling weights ensure that more recently 
emerged variants have higher selection probability: reference rows carry weight 1,
while Variant _i_ rows carry weight `2^(i+1)`.

```{r}
sim_result <- simulate_variant_evolution(
  ref_sequences             = ref_sequence,
  n_ref_months              = 4,
  start_date                = "2020-01-01",
  end_date                  = "2021-12-01",
  variants_config           = c(2, 3, 4, 2),
  variant_intervals         = c(4, 5, 5),
  n_new_mutations           = 2,
  mutation_rate             = c(1.25, 1.25, 1.25, 1.25),
  mutation_rate_variability = 0,
  deleterious_rate          = 0,
  n_deleterious_limit       = 0,
  n_sequences_total         = 1000,
  ref_variability           = TRUE,
  n_ref_sequences           = 200,
  prob_deletion_event       = 0.2,
  n_rows_to_delete          = 3,
  seed                      = 2025
)
```

## Check the Pool

```{r}
sim_result$Pool[, c(25, 26, 30)]   # Variant, Phase, and weights
```

## Emergence Periods and Sites

```{r}
print("Variant 1 emergence period:")
sim_result$Variant_Details[[1]]$em
```

```{r}
print("Variant 1 sites:")
sim_result$Variant_Details[[1]]$pos %>% sort()
```

```{r}
# Month 9 translates into partition 5.
print("Variant 2 emergence period:")
sim_result$Variant_Details[[2]]$em
```

```{r}
print("Variant 2 sites:")
sim_result$Variant_Details[[2]]$pos %>% sort()
```

```{r}
# Month 14 translates into partition 7.
print("Variant 3 emergence period:")
sim_result$Variant_Details[[3]]$em
```

```{r}
print("Variant 3 sites:")
sim_result$Variant_Details[[3]]$pos %>% sort()
```

```{r}
# Month 19 translates into partition 10.
print("Variant 4 emergence period:")
sim_result$Variant_Details[[4]]$em
```

```{r}
print("Variant 4 sites:")
sim_result$Variant_Details[[4]]$pos %>% sort()
```

## Deleterious Periods and Sites

```{r}
sprintf("The number of deleterious periods is %d",
        length(sim_result$Delet_Records))

print("Deleterious Period 1:")
sim_result$Delet_Records[[1]]$period
```

## Reference Phase 1 (month 1)

Phase `RV` means reference sequences with added variability.

```{r}
sim_result$Simulation_Output[1:10, ]
```

## Reference Phase 4 (month 4)

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 4, ][1:10, ]
```

## `Variant 1` Phase 1 (month 5): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

`Reference` can be either `R` or `RV`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 5,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 1` Phase 2 (month 6): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

Number of Variant 1 sequences here is `r ceiling(2 * 1.25)`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 6,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 1` Phase 4 (month 8): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

Number of Variant 1 sequences here is `r ceiling(ceiling(ceiling(2 * 1.25) * 1.25) * 1.25)`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 8,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 2` Phase 1 (month 9): sites `r sim_result$Variant_Details[[2]]$pos %>% sort()`

Number of Variant 2 sequences here is 2. Number of Variant 1 sequences here is
`r ceiling(ceiling(ceiling(ceiling(2 * 1.25) * 1.25) * 1.25) * 1.25)` as they 
continue to multiply. Now Variants 1 and 2 are in competition. The rest of the 
sequences are sampled from the reference pool. There is also a deleterious 
mutation here at site `r sim_result$Delet_Records[[1]]$site`.

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 9, ]
```

## `Variant 2` Phase 5 (month 13): sites `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 13, ]
```

## `Variant 3` Phase 1 (month 14): sites `r sim_result$Variant_Details[[3]]$pos %>% sort()`

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 14, ]
```

# Sites Detection and Change Point Detection

## Encoding

```{r}
toy_example <- sim_result$Simulation_Output
n_col       <- nchar(ref_sequence)

colnames(toy_example)[seq_len(n_col)] <- as.character(seq_len(n_col))

# Encode character sequences to integers
char_mat    <- as.matrix(toy_example[, seq_len(n_col)])
AL_Cat_mix  <- as.data.frame(encode_aa_sequence(char_mat))
AL_Cat_mix[] <- lapply(AL_Cat_mix, as.integer)

# Standardised Date column (first day of the month)
AL_Cat_mix$Date <- as.Date(format(as.Date(toy_example$Date), "%Y-%m-01"))
```

### View Post-Processed Matrix

```{r}
AL_Cat_mix[1:10, ]
```

## Partitioning

Twelve months of mutation phase (24 months total, 4 reference) with 2-month
non-overlapping windows give 12 partitions.

```{r}
part_data <- partition_time_windows(
  data          = AL_Cat_mix,
  n_sites       = n_col,
  window_length = 2,
  window_type   = 3,
  start_date    = "2020-01-01",
  end_date      = "2022-01-01"
)
```

### Dates Included in the First Partition

```{r}
part_data$Partitions[[1]]$Date %>% unique() %>% sort()
```

### Dates Included in the Last Partition

```{r}
part_data$Partitions[[length(part_data$Partitions)]]$Date %>% unique() %>% sort()
```

### Sites and Classes Detected in Partition 1

```{r}
part_data$Clusters[[1]]$DataFrame
```

### Sites and Classes Detected in Partition 2

```{r}
part_data$Clusters[[2]]$DataFrame
```

### Sites and Classes Detected in Partition 3 vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`

```{r}
part_data$Clusters[[3]]$DataFrame
```

### Sites and Classes Detected in Partition 4

```{r}
part_data$Clusters[[4]]$DataFrame
```

### Sites and Classes Detected in Partition 5 vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()` and `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
part_data$Clusters[[5]]$DataFrame
```

### Sites and Classes Detected in Partition 6 vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()` and `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
part_data$Clusters[[6]]$DataFrame
```

### Sites and Classes Detected in Partition 7 vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()` and `r sim_result$Variant_Details[[3]]$pos %>% sort()`

```{r}
part_data$Clusters[[7]]$DataFrame
```

### Sites and Classes Detected in Partition 8

```{r}
part_data$Clusters[[8]]$DataFrame
```

### Sites and Classes Detected in Partition 9

```{r}
part_data$Clusters[[9]]$DataFrame
```

### Sites and Classes Detected in Partition 10 vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()`, `r sim_result$Variant_Details[[3]]$pos %>% sort()` and `r sim_result$Variant_Details[[4]]$pos %>% sort()`

```{r}
part_data$Clusters[[10]]$DataFrame
```

### Sites and Classes Detected in Partition 11

```{r}
part_data$Clusters[[11]]$DataFrame
```

### Sites and Classes Detected in Partition 12

```{r}
part_data$Clusters[[12]]$DataFrame
```

## Hellinger Distance (HD)

### Calculate HD

```{r}
all_sites <- seq_len(n_col)

hellinger_mat <- calculate_hellinger_matrix(
  partitions = part_data$Partitions,
  sites      = all_sites,
  aa_levels  = 20L   # 20 standard amino acids
)
```

### Examine HD Structure: `r n_col` sites × (N partitions − 1) comparisons

```{r}
# Rows = sites, Columns = partition differences (T2 vs T1, T3 vs T1, ...)
dim(hellinger_mat)
hellinger_mat
```

### Transpose for Change Point Detection

Now we have multivariate (24-dimensional) observations at each partition
difference — ready for ECP.

```{r}
dat_mat_t <- t(hellinger_mat)
dat_mat_t
```

## Change Point Detection

### Method 1: Dynamic Programming via `ks.cp3o`

| **Parameter** | **Description**                                              | **Default** |
|---------------|--------------------------------------------------------------|-------------|
| `Z`           | A `T × d` matrix of the time series.                        | —           |
| `K`           | Maximum number of change points.                             | 1           |
| `minsize`     | Minimum segment size.                                        | 30          |
| `verbose`     | Print status updates.                                        | FALSE       |

```{r}
cp_m1 <- ks.cp3o(
  Z       = dat_mat_t,
  K       = 11,
  minsize = 2,
  verbose = FALSE
)

print("Change points estimated by Method 1:")
cp_m1$estimates

print("Other change point locations:")
cp_m1$cpLoc
```

### Method 2: Rolling ECP via `detect_changepoints_ecp`

Starting with the first 3 partitions, then expanding one step at a time.

```{r}
lower     <- 1
upper     <- 3
timesteps <- nrow(dat_mat_t) - upper

ECP_res <- detect_changepoints_ecp(
  data_matrix = dat_mat_t,
  min_window  = lower,
  max_window  = upper,
  n_timesteps = timesteps,
  rolling_window = FALSE,
  dynamic_k   = TRUE,
  n_cores     = 1L,
  minsize     = 2,
  verbose     = FALSE
)

ecp_indices <- ECP_res$ECP_est_list %>%
  unlist() %>%
  .[. >= 1] %>%
  unique() %>%
  sort()

print("Change points estimated by Method 2:")
ecp_indices
```

### Method 3: Agglomerative ECP via `e.agglo` (preferred)

| **Parameter** | **Description**                                              | **Default** |
|---------------|--------------------------------------------------------------|-------------|
| `X`           | A `T × d` matrix of the time series.                        | —           |
| `member`      | Initial membership vector.                                   | `1:nrow(X)` |
| `alpha`       | Moment index for cluster distance.                           | 1           |
| `penalty`     | Goodness-of-fit penalty function.                            | —           |

```{r}
cp_m3 <- e.agglo(
  X       = dat_mat_t,
  member  = seq_len(nrow(dat_mat_t)),
  alpha   = 1,
  penalty = function(cps) { 0 }
)

print("Change points estimated by Method 3 (naive):")
cp_m3$estimates

print("Cluster membership per time point:")
cp_m3$cluster
```
