---
title: "Variant Simulation and Analysis: Step by Step"
author: "Package Maintainer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Variant Simulation and Analysis: Step by Step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

# Introduction

This vignette provides a comprehensive, step-by-step walkthrough of the
`ViralEntropR` pipeline. We reproduce the exact benchmark scenario from the
study design:

- Simulate a realistic viral evolutionary timeline with four competing variants
  (2020–2021).
- Partition the data into non-overlapping 2-month time windows.
- Calculate Hellinger distance matrices to quantify distributional shifts.
- Detect evolutionary change points using three complementary ECP approaches.

For each partition, we calculate site-wise Shannon entropies, cluster these
entropy profiles using a Gaussian Mixture Model (GMM), and identify which sites
fall into the highest-entropy class.  We then compute Hellinger distances
between partitions using the first partition as the reference.  If there are 12
partitions there are 11 comparisons; if a variant emerges in partition 3, the
correct change point is index 2 (difference between partition 3 and partition 1).

# Setup

```{r}
library(ViralEntropR)
library(dplyr)
library(lubridate)
library(ecp)
library(knitr)
library(kableExtra)
```

# Reference Sequence

```{r}
ref_sequence <- "MKTIIALSYIFCLVFADYKDDDDK"
sprintf("There are %d sites in the reference sequence", nchar(ref_sequence))
```

# Simulation Function: `simulate_variant_evolution`

| **Parameter**               | **Description**                                                                  | **Default** |
|-----------------------------|----------------------------------------------------------------------------------|-------------|
| `ref_sequences`             | data.frame which must include a `Date` column; OR a single sequence string.      | —           |
| `n_ref_months`              | number of months for the reference phase                                         | 3           |
| `start_date`                | simulation start (character or Date)                                             | —           |
| `end_date`                  | simulation end (character or Date)                                               | —           |
| `variants_config`           | integer vector, e.g., `c(2,6)`: Variant 1 gets 2 mutations; Variant 2 gets 6.   | —           |
| `variant_intervals`         | gap (in months) between the emergence of consecutive variants                    | —           |
| `n_new_mutations`           | de-novo sequences produced at a variant's first appearance                       | 1           |
| `mutation_rate`             | monthly growth multiplier per variant (scalar or per-variant vector)             | 2           |
| `mutation_rate_variability` | fractional spread ± around `mutation_rate`; 0 = deterministic growth            | 0.25        |
| `deleterious_rate`          | probability that any new mutation is deleterious                                 | 0           |
| `n_deleterious_limit`       | maximum sequences allowed to carry a deleterious mutation                        | 1           |
| `n_sequences_total`         | total sequences generated over the mutation phase (spread evenly per month)      | 140         |
| `ref_variability`           | if `TRUE`, introduce variability at sites 1 and 2 of the reference pool          | FALSE       |
| `n_ref_sequences`           | when `ref_sequences` is a single string, number of reference rows to create      | 100         |
| `prob_deletion_event`       | per-month probability of injecting a block of sequences with a random deletion   | 0           |
| `n_rows_to_delete`          | sequences affected when `prob_deletion_event` fires                              | 0           |
| `seed`                      | optional random seed for reproducibility                                         | NULL        |

## Simulate Data

In this simulation, `simulate_variant_evolution()` models the stepwise
emergence and competition of four viral variants over a two-year period from
`2020-01-01` through `2021-12-01`.  We begin with a 4-month reference phase
(`n_ref_months = 4`) in which 200 total reference sequences are sampled
(50 per month, `n_ref_sequences = 200`), and variability is introduced at
sites 1 and 2 (`ref_variability = TRUE`).  The mutation phase then generates
1,000 sequences (`n_sequences_total = 1000`) at 50 per month.  Four variants
emerge at months 5, 9, 14 and 19, carrying 2, 3, 4 and 2 mutations
respectively (`variants_config = c(2,3,4,2)`), with 2 de-novo sequences at
each emergence (`n_new_mutations = 2`) and growth multipliers of 1.25 per
variant (`mutation_rate`).  Although `deleterious_rate = 0` for
variant-originated deleterious mutations, each month has a 20% chance
(`prob_deletion_event = 0.2`) of injecting a block of 3 sequences with a
random deletion at one site (`n_rows_to_delete = 3`).  Finally, `seed = 2025`
ensures reproducibility.  The returned list includes `Simulation_Output` (with
per-sequence `Delet` flags), `Variant_Details`, `Simulation_Dates`, and
`Delet_Records` for downstream analysis.

Variants compete in a **pairwise** fashion, always pitting the newest active
variant against its immediate predecessor (or the original reference when only
one variant has emerged).  During months before any variant appears, all
sequences are drawn from the reference pool (the wild-type row plus up to two
reference-variability rows when `ref_variability = TRUE`).  Once Variant 1
emerges, each month's batch is composed of its de-novo copies plus sampled
reference sequences; when Variant 2 appears it competes directly with Variant
1, and so on — Variant 3 competes against Variant 2, Variant 4 against Variant
3.  The **fill pool** (sequences that pad the monthly quota after the two
competing variants are allocated) is composed of the reference rows plus any
older non-competing variants.  Sampling weights are assigned so that all
reference rows carry weight 1, while Variant *i* rows carry weight `2^(i+1)`,
ensuring that more recently emerged variants have higher selection probability
in subsequent months.  Growth is **stochastic**: each variant's monthly count
is drawn from
`Uniform(mult*(1 - variability), mult*(1 + variability))`
so that `mutation_rate_variability > 0` introduces realistic fluctuation.

```{r}
sim_result <- simulate_variant_evolution(
  ref_sequences             = ref_sequence,            # single sequence string
  n_ref_months              = 4,                       # 4-month reference phase
  start_date                = "2020-01-01",
  end_date                  = "2021-12-01",
  variants_config           = c(2, 3, 4, 2),           # 4 variants: 2,3,4,2 mutations
  variant_intervals         = c(4, 5, 5),              # V2 at month 9; V3 at 14; V4 at 19
  n_new_mutations           = 2,                       # 2 de-novo sequences at emergence
  mutation_rate             = c(1.25, 1.25, 1.25, 1.25), # per-variant growth multipliers
  mutation_rate_variability = 0,                       # deterministic growth (no spread)
  deleterious_rate          = 0,                       # no within-variant deleterious mutations
  n_deleterious_limit       = 0,
  n_sequences_total         = 1000,                    # 1000/(24-4) = 50 per period
  ref_variability           = TRUE,                    # variability at sites 1 and 2
  n_ref_sequences           = 200,                     # 200/4 = 50 reference rows per month
  prob_deletion_event       = 0.2,                     # 20% per-month deletion probability
  n_rows_to_delete          = 3,                       # 3 sequences affected if event fires
  seed                      = 2025
)
```

## Check the Pool

The pool returned is the fill pool from the last multi-variant competition
period.  Column 25 = Variant, column 26 = Phase, column 30 = sampling weight.

```{r}
sim_result$Pool[, c(25, 26, 30)]
```

## Emergence Periods and Sites

```{r}
print("Variant emergence period:")
sim_result$Variant_Details[[1]]$em
```

```{r}
print("Variant 1 sites:")
sim_result$Variant_Details[[1]]$pos %>% sort()
```

```{r}
# Month 9 translates into partition 5.
print("Variant 2 emergence period:")
sim_result$Variant_Details[[2]]$em
```

```{r}
print("Variant 2 sites:")
sim_result$Variant_Details[[2]]$pos %>% sort()
```

```{r}
# Month 14 translates into partition 7.
print("Variant 3 emergence period:")
sim_result$Variant_Details[[3]]$em
```

```{r}
print("Variant 3 sites:")
sim_result$Variant_Details[[3]]$pos %>% sort()
```

```{r}
# Month 19 translates into partition 10.
print("Variant 4 emergence period:")
sim_result$Variant_Details[[4]]$em
```

```{r}
print("Variant 4 sites:")
sim_result$Variant_Details[[4]]$pos %>% sort()
```

## Deleterious Periods and Sites

```{r}
sprintf("The number of deleterious periods is %d",
        length(sim_result$Delet_Records))

for (i in seq_along(sim_result$Delet_Records)) {
  cat(sprintf("Deleterious Period %d: month %d, site %d\n",
              i,
              sim_result$Delet_Records[[i]]$period,
              sim_result$Delet_Records[[i]]$site))
}
```

## Reference Phase 1 (month 1)

Phase `RV` denotes reference sequences with added variability at sites 1 or 2.

```{r}
sim_result$Simulation_Output[1:10, ]
```

## Reference Phase 4 (month 4)

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 4, ][1:10, ]
```

## `Variant 1` Phase 1 (month 5): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

`Reference` can be either `R` or `RV`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 5,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 1` Phase 2 (month 6): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

Number of Variant 1 sequences here is `r ceiling(2 * 1.25)`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 6,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 1` Phase 4 (month 8): sites `r sim_result$Variant_Details[[1]]$pos %>% sort()`

Number of Variant 1 sequences here is
`r ceiling(ceiling(ceiling(2 * 1.25) * 1.25) * 1.25)`.

```{r}
sim_result$Simulation_Output[
  sim_result$Simulation_Output$Period == 8,
  c(sim_result$Variant_Details[[1]]$pos %>% sort(), 25:29)
][1:10, ]
```

## `Variant 2` Phase 1 (month 9): sites `r sim_result$Variant_Details[[2]]$pos %>% sort()`

Number of Variant 2 sequences here is 2.  Number of Variant 1 sequences here is
`r ceiling(ceiling(ceiling(ceiling(2 * 1.25) * 1.25) * 1.25) * 1.25)` as they
continue to multiply.  Variants 1 and 2 are now in competition.  The remainder
of the 50 sequences is sampled from the fill pool (reference rows only, since
there are no older non-competing variants yet).  There is also a deleterious
mutation here at site `r sim_result$Delet_Records[[1]]$site`.

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 9, ]
```

## `Variant 2` Phase 5 (month 13): sites `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 13, ]
```

## `Variant 3` Phase 1 (month 14): sites `r sim_result$Variant_Details[[3]]$pos %>% sort()`

```{r}
sim_result$Simulation_Output[sim_result$Simulation_Output$Period == 14, ]
```

# Sites Detection and Change Point Detection

## Encoding

```{r}
toy_example <- sim_result$Simulation_Output
n_col       <- nchar(ref_sequence)

colnames(toy_example)[seq_len(n_col)] <- as.character(seq_len(n_col))

# Encode character sequences to integers (A=1 … V=20, ambiguous 21–25, unknown=0)
char_mat   <- as.matrix(toy_example[, seq_len(n_col)])
AL_Cat_mix <- as.data.frame(encode_aa_sequence(char_mat))
AL_Cat_mix[] <- lapply(AL_Cat_mix, as.integer)

# Standardise Date column to the first day of each month
AL_Cat_mix$Date <- as.Date(format(as.Date(toy_example$Date), "%Y-%m-01"))
```

### View Post-Processed Matrix

```{r}
AL_Cat_mix[1:10, ]
```

## Partitioning

24 total months with a 4-month reference phase and 2-month non-overlapping
windows give 12 partitions.

```{r}
part_data <- partition_time_windows(
  data          = AL_Cat_mix,
  n_sites       = n_col,
  window_length = 2,
  window_type   = 3,       # non-overlapping / jumping
  start_date    = "2020-01-01",
  end_date      = "2022-01-01"
)
```

### Dates Included in the First Partition

```{r}
part_data$Partitions[[1]]$Date %>% unique() %>% sort()
```

### Dates Included in the Last Partition

```{r}
part_data$Partitions[[part_data$N_partitions]]$Date %>% unique() %>% sort()
```

### Sites and Classes Detected in the First Partition

```{r}
part_data$Clusters[[1]]$DataFrame
```

### Sites and Classes Detected in the Second Partition

```{r}
part_data$Clusters[[2]]$DataFrame
```

### Sites and Classes Detected in the Third Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`

```{r}
part_data$Clusters[[3]]$DataFrame
```

### Sites and Classes Detected in the 4th Partition

```{r}
part_data$Clusters[[4]]$DataFrame
```

### Sites and Classes Detected in the 5th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()` and `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
part_data$Clusters[[5]]$DataFrame
```

### Sites and Classes Detected in the 6th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()` and `r sim_result$Variant_Details[[2]]$pos %>% sort()`

```{r}
part_data$Clusters[[6]]$DataFrame
```

### Sites and Classes Detected in the 7th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()` and `r sim_result$Variant_Details[[3]]$pos %>% sort()`

```{r}
part_data$Clusters[[7]]$DataFrame
```

### Sites and Classes Detected in the 8th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()` and `r sim_result$Variant_Details[[3]]$pos %>% sort()`

```{r}
part_data$Clusters[[8]]$DataFrame
```

### Sites and Classes Detected in the 9th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()` and `r sim_result$Variant_Details[[3]]$pos %>% sort()`

```{r}
part_data$Clusters[[9]]$DataFrame
```

### Sites and Classes Detected in the 10th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()`, `r sim_result$Variant_Details[[3]]$pos %>% sort()` and `r sim_result$Variant_Details[[4]]$pos %>% sort()`

```{r}
part_data$Clusters[[10]]$DataFrame
```

### Sites and Classes Detected in the 11th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()`, `r sim_result$Variant_Details[[3]]$pos %>% sort()` and `r sim_result$Variant_Details[[4]]$pos %>% sort()`

```{r}
part_data$Clusters[[11]]$DataFrame
```

### Sites and Classes Detected in the 12th Partition vs. `r sim_result$Variant_Details[[1]]$pos %>% sort()`, `r sim_result$Variant_Details[[2]]$pos %>% sort()`, `r sim_result$Variant_Details[[3]]$pos %>% sort()` and `r sim_result$Variant_Details[[4]]$pos %>% sort()`

```{r}
part_data$Clusters[[12]]$DataFrame
```

## Hellinger Distance (HD)

### Calculate HD

```{r}
all_sites <- seq_len(n_col)

hellinger_mat <- calculate_hellinger_matrix(
  partitions = part_data$Partitions,
  sites      = all_sites,
  aa_levels  = 25L   # verify
)
```

### Examine HD Structure: `r n_col` sites × `r ncol(hellinger_mat)` partition differences

Rows = sites; columns = partition differences (T2 vs T1, T3 vs T1, …).

```{r}
dim(hellinger_mat)
hellinger_mat
```

### Save HD into Matrix and Transpose

Now we have `r ncol(hellinger_mat)` multivariate (`r n_col`-dimensional) points
corresponding to the `r ncol(hellinger_mat)` partition differences.

```{r}
dat_mat_t <- t(hellinger_mat)
dat_mat_t
```

## HD Change Point Detection

### Method 1: Dynamic Programming via `ks.cp3o`

| **Parameter** | **Description**                                                              | **Default** |
|---------------|------------------------------------------------------------------------------|-------------|
| `Z`           | A `T × d` matrix containing the length-`T` time series with `d` dimensions. | —           |
| `K`           | Maximum number of change points.                                             | 1           |
| `minsize`     | Minimum segment size.                                                        | 30          |
| `verbose`     | Print status updates.                                                        | FALSE       |

```{r}
cp_m1 <- ks.cp3o(
  Z       = dat_mat_t,
  K       = 11,
  minsize = 2,
  verbose = FALSE
)

print("Change Points estimated by method 1:")
cp_m1$estimates

print("Other change point locations:")
cp_m1$cpLoc
```

### Method 2: Rolling ECP via `detect_changepoints_ecp`

Same as above, but starting with the first 3 partition differences and
expanding the window one step at a time.

```{r}
lower     <- 1
upper     <- 3
timesteps <- nrow(dat_mat_t) - upper

ECP_res <- detect_changepoints_ecp(
  data_matrix    = dat_mat_t,
  min_window     = lower,
  max_window     = upper,
  n_timesteps    = timesteps,
  rolling_window = FALSE,
  dynamic_k      = TRUE,
  minsize        = 2,
  verbose        = FALSE
)

ecp_indices <- ECP_res$ECP_est_list %>%
  unlist() %>%
  .[. >= 1] %>%
  unique() %>%
  sort()

print("Change points estimated by method 2:")
ecp_indices
```

### Method 3: Agglomerative Hierarchical ECP via `e.agglo` (preferred)

| **Parameter** | **Description**                                                              | **Default**  |
|---------------|------------------------------------------------------------------------------|--------------|
| `X`           | A `T × d` matrix containing the length-`T` time series with `d` dimensions. | —            |
| `member`      | Initial membership vector for the time series.                               | `1:nrow(X)`  |
| `alpha`       | Moment index used for determining distance between and within clusters.      | 1            |
| `penalty`     | Function used to penalise the goodness-of-fit statistics.                    | —            |

```{r}
cp_m3 <- e.agglo(
  X       = dat_mat_t,
  member  = seq_len(nrow(dat_mat_t)),  # naive: no prior knowledge of partitions
  alpha   = 1,
  penalty = function(cps) { 0 }
)

print("Change points estimated by method 3 (naive):")
cp_m3$estimates

print("Cluster membership per time point:")
cp_m3$cluster
```
