---
title: "Gower Distance Matrix Correlation Between Full 1,273-Site and GMM Entropy-Reduced Representations of SARS-CoV-2 Spike Protein Sequences"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    theme: flatly
    highlight: tango
    includes:
      in_header: meta_correlation_analysis.html
bibliography: references_correlation.bib
nocite: '@*'
---

# Overview

The ViralEntropR pipeline reduces the dimensionality of SARS-CoV-2 Spike
protein sequence data from 1,273 sites to a compact subset selected by
Gaussian mixture model (GMM) clustering of per-site Shannon entropy. A
central question for the validity of this reduction is how faithfully the
pairwise sequence relationships encoded in the full distance matrix are
preserved after dimensionality reduction. This script addresses that question
directly via the Mantel test — a permutation-based procedure that quantifies
the correlation between two distance matrices.

Gower distance matrices are computed on deduplicated unique sequence
haplotypes, both from the full 1,273-site representation (reference) and from
progressively expanded GMM entropy-reduced subsets. Site selection begins with
the single highest-entropy GMM class (class 1 after relabeling, where class 1
always denotes the highest-entropy group) and accumulates successively
lower-entropy classes through k = 1, 2, ..., N − 1, where N is the total
number of fitted GMM components and the lowest-entropy class is excluded as it
carries minimal evolutionary signal. For each k, a Pearson Mantel test
quantifies the correlation between the reduced and full distance matrices. At
the value of k achieving the maximum Pearson correlation, a Spearman Mantel
test is additionally computed as a non-parametric confirmation robust to
distributional assumptions.

Both NCBI US (January 2020 to September 2021, 109,536 sequences) and GISAID
US (March 2020 to June 2022, 129,371 sequences) post-processed datasets are
analysed independently. Preprocessing pipelines for both datasets are documented
separately: the NCBI pipeline is described in the `preprocessing_pipeline`
package vignette
(`vignette("preprocessing_pipeline", package = "ViralEntropR")`), and the GISAID
pipeline is recorded in
`analysis/GISAID_data_preprocessing/GISAID_Preprocessing_Pipeline.Rmd`.

All Mantel permutation tests are parallelised across available server cores.
Reduced distance matrices are written to disk incrementally — one per k
iteration — and removed from the R workspace immediately, so that peak memory
at any point equals `dist_full` plus a single `dist_red`, not the entire
collection. For GISAID US, where near-total sequence uniqueness (129,345 of
129,371 sequences) makes a full pairwise distance matrix infeasible (~67 GB),
a fixed random subsample of `N_MAX_UNIQUE` unique haplotypes is drawn
(reproducibly, via `set.seed(42L)`) before distance computation. The
comparison remains full 1,273-site vs entropy-reduced — both matrices are
computed on the same subsample — and the subsampling is documented
transparently in the results. Results are summarised numerically and
visualised as a faceted publication-quality figure.

# Setup

## Libraries

```{r libraries, message=FALSE, warning=FALSE}

library(ViralEntropR)
library(vegan)         # mantel()
library(cluster)       # daisy()
library(data.table)    # fast deduplication
library(tictoc)        # tic() / toc()
library(ggplot2)
library(scales)
library(here)

```

## Global Configuration

```{r configuration}

# ---------------------------------------------------------------------------
# All user-facing parameters are gathered here.
# ---------------------------------------------------------------------------

N_PERM       <- 9999L    # permutations per Mantel test (publication standard)
N_MAX_UNIQUE <- 10000L   # maximum unique sequences for Gower distance computation.
                         # Applied only when unique count exceeds this value
                         # (currently GISAID US: 129,345 unique sequences).
                         # NCBI US (8,155 unique) is unaffected.
SUBSAMPLE_SEED <- 42L    # reproducibility seed for subsampling

OUT_DIR <- here("analysis", "Correlation_Analysis", "outputs")
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# ---------------------------------------------------------------------------
# Core count: runtime RAM check (Linux /proc/meminfo).
# Each vegan::mantel() worker forks approximately 1.5 GB (two distance
# vectors + permutation overhead). Cap at 64 workers — diminishing returns
# from vegan's internal mclapply beyond that for 9999 permutations.
# ---------------------------------------------------------------------------
ram_available_gb <- tryCatch({
  as.numeric(system(
    "awk '/MemAvailable/ {print $2}' /proc/meminfo",
    intern = TRUE
  )) / 1e6
}, error = function(e) {
  warning("Could not read /proc/meminfo. Defaulting to 64 GB available RAM.")
  64
})

worker_ram_gb <- 3.5
N_CORES <- min(
  32L,
  floor(ram_available_gb / worker_ram_gb) - 4L
)
N_CORES <- max(1L, N_CORES)

cat(sprintf("Available RAM:              %.1f GB\n", ram_available_gb))
cat(sprintf("Est. RAM per worker:        %.1f GB\n", worker_ram_gb))
cat(sprintf("Cores for Mantel tests:     %d\n",      N_CORES))
cat(sprintf("Permutations:               %d\n",      N_PERM))
cat(sprintf("Max unique seqs (Gower):    %d\n",      N_MAX_UNIQUE))
cat(sprintf("Subsample seed:             %d\n",      SUBSAMPLE_SEED))
cat(sprintf("Output directory:           %s\n",      OUT_DIR))

```

## Dataset Configurations

```{r dataset-configs}

# Each entry specifies: display label, path to preprocessed RDS, and output
# paths for the distance matrix list and Mantel results.
datasets <- list(

  NCBI_US = list(
    label        = "NCBI US",
    rds_path     = here("NCBI_US_unaligned_feature_matrix_1273aa.rds"),
    dist_rds     = file.path(OUT_DIR, "NCBI_US_distance_matrices.rds"),
    results_rds  = file.path(OUT_DIR, "NCBI_US_mantel_results.rds")
  ),

  GISAID_US = list(
    label        = "GISAID US",
    rds_path     = here("GISAID_US_aligned_feature_matrix_1273aa.rds"),
    dist_rds     = file.path(OUT_DIR, "GISAID_US_distance_matrices.rds"),
    results_rds  = file.path(OUT_DIR, "GISAID_US_mantel_results.rds")
  )

)

```

## Helper Functions

```{r helpers}

# Runs a single Mantel test with timing output.
# Arguments:
#   dist_full : full reference Gower distance matrix
#   dist_red  : reduced Gower distance matrix
#   method    : "pearson" or "spearman"
#   n_perm    : number of permutations
#   n_cores   : parallel workers
#   label     : descriptive label for the tictoc message
run_mantel <- function(dist_full, dist_red, method, n_perm, n_cores,
                       label = "") {
  tic(sprintf("Mantel [%s] %s", toupper(method), label))
  res <- vegan::mantel(
    xdis         = dist_full,
    ydis         = dist_red,
    method       = method,
    permutations = n_perm,
    parallel     = n_cores
  )
  toc()
  res
}

```

# Analysis — NCBI US and GISAID US

Both datasets are processed sequentially. Within each dataset, Mantel tests
are parallelised across `r N_CORES` cores. Reduced distance matrices are
saved to disk incrementally after each k iteration and removed from the R
workspace immediately — peak memory at any point is `dist_full` plus a single
`dist_red`. The lowest GMM entropy class (class N after relabeling) is
excluded from the accumulation sweep.

```{r main-analysis, message=FALSE, warning=FALSE}

# Storage for all results — populated across the dataset loop.
all_results <- list()

for (ds_key in names(datasets)) {

  ds <- datasets[[ds_key]]

  # =========================================================================
  # Checkpoint: skip if results already exist for this dataset.
  # =========================================================================
  if (file.exists(ds$results_rds)) {
    message(sprintf("\n[%s] Checkpoint found — loading saved results.", ds$label))
    all_results[[ds_key]] <- readRDS(ds$results_rds)
    next
  }

  message(sprintf(
    "\n%s\n[%s] Starting analysis\n%s",
    strrep("=", 60L), ds$label, strrep("=", 60L)
  ))

  # =========================================================================
  # Step 1: Load preprocessed data
  # =========================================================================
  tic(sprintf("[%s] Load data", ds$label))
  AL_df <- readRDS(ds$rds_path)
  toc()

  # Retain site columns only — select by name to guard against column order
  # differences between NCBI and GISAID preprocessed matrices.
  site_cols  <- setdiff(colnames(AL_df), c("Date", "Country"))
  AL_df      <- AL_df[, site_cols, drop = FALSE]
  N_SITES_DS <- ncol(AL_df)
  cat(sprintf("[%s] Site columns retained: %d | Sequences: %d\n",
              ds$label, N_SITES_DS, nrow(AL_df)))

  # =========================================================================
  # Step 2: Shannon entropy and GMM clustering
  # Entropy is computed on the full dataset before any deduplication or
  # subsampling — it must reflect the population-level site variability.
  # =========================================================================
  tic(sprintf("[%s] Entropy computation", ds$label))
  entrp_full <- apply(AL_df, 2L, calculate_entropy)
  toc()

  # GMM: relabel so class 1 = highest-entropy group throughout.
  tic(sprintf("[%s] GMM clustering + relabeling", ds$label))
  clust_obj <- cluster_sites_by_entropy(entrp_full, nr = nrow(AL_df))
  clust_rl  <- relabel_entropy_classes(clust_obj$DataFrame)
  N_classes <- max(clust_rl$class[clust_rl$class != 999L])
  toc()

  cat(sprintf("[%s] GMM fitted %d entropy classes (sentinel 999 excluded).\n",
              ds$label, N_classes))

  # Pre-compute site indices per class for O(1) lookup in the k-loop.
  class_to_sites <- lapply(seq_len(N_classes), function(cl)
    sort(clust_rl$sites[clust_rl$class == cl])
  )
  names(class_to_sites) <- as.character(seq_len(N_classes))

  rm(entrp_full, clust_obj, clust_rl); gc()

  # =========================================================================
  # Step 3: Deduplication
  # Uses data.table for memory-efficient identification of unique rows.
  # =========================================================================
  tic(sprintf("[%s] Deduplication", ds$label))
  setDT(AL_df)
  unique_idx   <- which(!duplicated(AL_df))
  AL_df_unique <- as.data.frame(AL_df[unique_idx, ])
  rm(AL_df); gc()
  n_unique <- nrow(AL_df_unique)
  cat(sprintf("[%s] Unique sequences: %d (from %d total)\n",
              ds$label, n_unique, length(unique_idx)))
  toc()

  # =========================================================================
  # Step 3b: Subsampling (GISAID only — applied when unique count > N_MAX_UNIQUE)
  #
  # GISAID US has 129,345 unique sequences out of 129,371 total — near-total
  # uniqueness. A full pairwise Gower distance matrix for 129,345 sequences
  # requires ~67 GB, exceeding server RAM. A reproducible random subsample of
  # N_MAX_UNIQUE sequences is drawn; both dist_full and all dist_red are
  # computed on the same subsample, preserving the validity of the comparison.
  # =========================================================================
  subsampled <- FALSE
  if (n_unique > N_MAX_UNIQUE) {
    set.seed(SUBSAMPLE_SEED)
    subsample_idx <- sample(n_unique, N_MAX_UNIQUE)
    AL_df_unique  <- AL_df_unique[subsample_idx, , drop = FALSE]
    subsampled    <- TRUE
    cat(sprintf(
      "[%s] Subsampled %d unique sequences to %d (seed = %d) for Gower computation.\n",
      ds$label, n_unique, N_MAX_UNIQUE, SUBSAMPLE_SEED
    ))
    gc()
  }

  # =========================================================================
  # Step 4: Full Gower distance matrix (reference; computed once per dataset)
  # =========================================================================
  n_for_gower <- nrow(AL_df_unique)
  est_mb <- (n_for_gower * (n_for_gower - 1L) / 2L * 8L) / 1e6

  tic(sprintf("[%s] Full Gower distance matrix (%d sites, %d seqs, est. %.0f MB)",
              ds$label, N_SITES_DS, n_for_gower, est_mb))
  AL_full_fac <- AL_df_unique
  AL_full_fac[] <- lapply(AL_full_fac, as.factor)
  dist_full <- cluster::daisy(AL_full_fac, metric = "gower")
  rm(AL_full_fac); gc()
  cat(sprintf("[%s] dist_full: %.1f MB\n",
              ds$label, as.numeric(object.size(dist_full)) / 1e6))
  toc()

  # Initialise the distance list RDS with dist_full only.
  # Reduced matrices are appended one-at-a-time inside the k-loop.
  # This prevents accumulation of all matrices in RAM simultaneously.
  saveRDS(list(dist_full = dist_full), ds$dist_rds)
  message(sprintf("[%s] dist_full saved to: %s", ds$label, ds$dist_rds))

  # =========================================================================
  # Step 5: Pearson Mantel sweep — k = 1 to N_classes - 1
  #
  # At each k:
  #   - Sites = union of GMM classes 1:k (class 1 = highest entropy).
  #   - Reduced Gower distance matrix computed from those sites.
  #   - dist_red appended to the RDS checkpoint immediately, then removed
  #     from the R workspace — peak RAM = dist_full + one dist_red.
  #   - Pearson Mantel test run against dist_full.
  #
  # The lowest-entropy class (class N_classes) is excluded from the sweep.
  # =========================================================================
  pearson_results <- vector("list", N_classes - 1L)

  for (k in seq_len(N_classes - 1L)) {

    # Accumulate sites from classes 1 through k.
    sites_k   <- sort(unique(unlist(class_to_sites[seq_len(k)])))
    n_sites_k <- length(sites_k)

    cat(sprintf("[%s] k = %d | %d sites (classes 1:%d)\n",
                ds$label, k, n_sites_k, k))

    # Reduced feature matrix.
    AL_red_fac <- AL_df_unique[, as.character(sites_k), drop = FALSE]
    AL_red_fac[] <- lapply(AL_red_fac, as.factor)

    tic(sprintf("[%s] Reduced Gower k = %d", ds$label, k))
    dist_red <- cluster::daisy(AL_red_fac, metric = "gower")
    toc()
    rm(AL_red_fac); gc()

    # Append dist_red to the RDS checkpoint BEFORE running the Mantel test.
    # Reload, append, overwrite — only one dist_red is in RAM during append.
    dist_list_tmp <- readRDS(ds$dist_rds)
    dist_list_tmp[[paste0("k", k)]] <- list(
      k        = k,
      sites_k  = sites_k,
      classes  = seq_len(k),
      dist_red = dist_red
    )
    saveRDS(dist_list_tmp, ds$dist_rds)
    rm(dist_list_tmp); gc()

    # Pearson Mantel test.
    mantel_p <- run_mantel(
      dist_full = dist_full,
      dist_red  = dist_red,
      method    = "pearson",
      n_perm    = N_PERM,
      n_cores   = N_CORES,
      label     = sprintf("%s k=%d", ds$label, k)
    )

    pearson_results[[k]] <- list(
      k          = k,
      n_sites    = n_sites_k,
      classes    = seq_len(k),
      r_pearson  = mantel_p$statistic,
      p_pearson  = mantel_p$signif,
      mantel_obj = mantel_p
    )

    cat(sprintf("[%s] k = %d | Pearson r = %.4f, p = %.4f\n",
                ds$label, k,
                pearson_results[[k]]$r_pearson,
                pearson_results[[k]]$p_pearson))

    rm(dist_red, mantel_p); gc()
  }

  message(sprintf("[%s] Distance matrix list complete: %s", ds$label, ds$dist_rds))

  # =========================================================================
  # Step 6: Identify optimal k(s) — maximum Pearson r
  # =========================================================================
  r_vals <- vapply(pearson_results, `[[`, numeric(1L), "r_pearson")
  r_max  <- max(r_vals)
  k_opt  <- which(r_vals == r_max)

  cat(sprintf(
    "[%s] Maximum Pearson r = %.4f at k = %s (n_sites = %s)\n",
    ds$label, r_max,
    paste(k_opt, collapse = ", "),
    paste(vapply(pearson_results[k_opt],
                 `[[`, numeric(1L), "n_sites"),
          collapse = ", ")
  ))

  # =========================================================================
  # Step 7: Spearman Mantel at optimal k(s) only.
  # Reduced matrices are reloaded from the saved RDS — dist_full is the only
  # large object held in memory at this stage.
  # =========================================================================
  spearman_results <- vector("list", length(k_opt))

  for (j in seq_along(k_opt)) {

    k <- k_opt[j]
    cat(sprintf("[%s] Spearman Mantel at optimal k = %d\n", ds$label, k))

    dist_list_loaded <- readRDS(ds$dist_rds)
    dist_red_k       <- dist_list_loaded[[paste0("k", k)]]$dist_red
    rm(dist_list_loaded); gc()

    mantel_s <- run_mantel(
      dist_full = dist_full,
      dist_red  = dist_red_k,
      method    = "spearman",
      n_perm    = N_PERM,
      n_cores   = N_CORES,
      label     = sprintf("%s k=%d Spearman", ds$label, k)
    )

    spearman_results[[j]] <- list(
      k          = k,
      n_sites    = pearson_results[[k]]$n_sites,
      classes    = pearson_results[[k]]$classes,
      r_spearman = mantel_s$statistic,
      p_spearman = mantel_s$signif,
      mantel_obj = mantel_s
    )

    cat(sprintf("[%s] k = %d | Spearman r = %.4f, p = %.4f\n",
                ds$label, k,
                spearman_results[[j]]$r_spearman,
                spearman_results[[j]]$p_spearman))

    rm(dist_red_k, mantel_s); gc()
  }

  # Release dist_full and unique sequence matrix before next dataset.
  rm(dist_full, AL_df_unique); gc()

  # =========================================================================
  # Step 8: Save Mantel results checkpoint
  # =========================================================================
  result_obj <- list(
    dataset          = ds$label,
    N_classes        = N_classes,
    N_sites          = N_SITES_DS,
    n_unique_total   = n_unique,
    n_used_gower     = n_for_gower,
    subsampled       = subsampled,
    subsample_seed   = if (subsampled) SUBSAMPLE_SEED else NA_integer_,
    class_to_sites   = class_to_sites,
    pearson_results  = pearson_results,
    spearman_results = spearman_results,
    k_optimal        = k_opt,
    r_max_pearson    = r_max
  )

  saveRDS(result_obj, ds$results_rds)
  message(sprintf("[%s] Mantel results saved: %s", ds$label, ds$results_rds))

  all_results[[ds_key]] <- result_obj

  rm(result_obj, pearson_results, spearman_results,
     class_to_sites, unique_idx, r_vals,
     n_unique, n_for_gower, subsampled); gc()

}   # end dataset loop

```

# Results Summary

```{r results-summary}

for (ds_key in names(all_results)) {

  res <- all_results[[ds_key]]
  cat(sprintf("\n%s\n%s — Summary\n%s\n",
              strrep("-", 50L), res$dataset, strrep("-", 50L)))
  cat(sprintf("  GMM classes fitted:      %d\n",  res$N_classes))
  cat(sprintf("  Unique sequences total:  %d\n",  res$n_unique_total))
  cat(sprintf("  Sequences used (Gower):  %d%s\n", res$n_used_gower,
              if (isTRUE(res$subsampled))
                sprintf(" (subsampled, seed = %d)", res$subsample_seed)
              else " (all unique)"))
  cat(sprintf("  Optimal k(s):            %s\n",
              paste(res$k_optimal, collapse = ", ")))
  cat(sprintf("  Maximum Pearson r:       %.4f\n", res$r_max_pearson))

  for (j in seq_along(res$spearman_results)) {
    sp <- res$spearman_results[[j]]
    cat(sprintf(
      "  Spearman r at k=%d:      %.4f (p = %.4f, n_sites = %d)\n",
      sp$k, sp$r_spearman, sp$p_spearman, sp$n_sites
    ))
  }

  cat("\n  Pearson r by k:\n")
  for (pr in res$pearson_results) {
    flag <- if (pr$k %in% res$k_optimal) " <-- optimal" else ""
    cat(sprintf("    k = %2d | n_sites = %3d | r = %.4f | p = %.4f%s\n",
                pr$k, pr$n_sites, pr$r_pearson, pr$p_pearson, flag))
  }
}

```

# Visualisation

```{r visualisation, fig.width=12, fig.height=10, message=FALSE, warning=FALSE}

# ---------------------------------------------------------------------------
# Assemble long-format data frames for ggplot2
# ---------------------------------------------------------------------------

pearson_plot_df <- do.call(rbind, lapply(names(all_results), function(ds_key) {
  res <- all_results[[ds_key]]
  do.call(rbind, lapply(res$pearson_results, function(x) {
    data.frame(
      dataset    = res$dataset,
      k          = x$k,
      n_sites    = x$n_sites,
      r_pearson  = x$r_pearson,
      p_pearson  = x$p_pearson,
      is_optimal = x$k %in% res$k_optimal,
      subsampled = isTRUE(res$subsampled),
      stringsAsFactors = FALSE
    )
  }))
}))

spearman_plot_df <- do.call(rbind, lapply(names(all_results), function(ds_key) {
  res <- all_results[[ds_key]]
  do.call(rbind, lapply(res$spearman_results, function(x) {
    data.frame(
      dataset    = res$dataset,
      k          = x$k,
      n_sites    = x$n_sites,
      r_spearman = x$r_spearman,
      p_spearman = x$p_spearman,
      stringsAsFactors = FALSE
    )
  }))
}))

annotation_df <- merge(
  pearson_plot_df[pearson_plot_df$is_optimal, ],
  spearman_plot_df[, c("dataset", "k", "r_spearman", "p_spearman")],
  by    = c("dataset", "k"),
  all.x = TRUE
)
annotation_df$label <- sprintf(
  "r[S] == %.3f\np == %.4f",
  annotation_df$r_spearman,
  annotation_df$p_spearman
)

# Facet label: append subsampling note where applicable
facet_labels <- vapply(names(all_results), function(ds_key) {
  res <- all_results[[ds_key]]
  if (isTRUE(res$subsampled))
    sprintf("%s\n(subsampled to %s unique sequences)",
            res$dataset, format(res$n_used_gower, big.mark = ","))
  else
    sprintf("%s\n(%s unique sequences)",
            res$dataset, format(res$n_used_gower, big.mark = ","))
}, character(1L))
names(facet_labels) <- vapply(all_results, `[[`, character(1L), "dataset")
pearson_plot_df$facet_label <- facet_labels[pearson_plot_df$dataset]
annotation_df$facet_label   <- facet_labels[annotation_df$dataset]

# ---------------------------------------------------------------------------
# Colour palette
# ---------------------------------------------------------------------------
ds_colours <- c(
  "NCBI US"   = "#004B9E",
  "GISAID US" = "#D9A400"
)

# ---------------------------------------------------------------------------
# Build plot
# ---------------------------------------------------------------------------
p <- ggplot(pearson_plot_df,
            aes(x = k, y = r_pearson, colour = dataset, group = dataset)) +

  geom_line(linewidth = 1.2) +

  geom_point(aes(shape = is_optimal), size = 3.5) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 18),
    guide  = "none"
  ) +

  geom_vline(
    data      = annotation_df,
    aes(xintercept = k, colour = dataset),
    linetype  = "dashed",
    linewidth = 0.8,
    alpha     = 0.55
  ) +

  geom_label(
    data        = annotation_df,
    aes(x = k, y = r_pearson, label = label),
    parse       = TRUE,
    hjust       = -0.12,
    vjust       = 0.5,
    size        = 3.5,
    colour      = "black",
    fill        = "white",
    label.size  = 0.3,
    inherit.aes = FALSE
  ) +

  scale_colour_manual(values = ds_colours, name = "Dataset") +

  scale_x_continuous(
    breaks = function(lims) seq(ceiling(lims[1]), floor(lims[2]), by = 1L),
    name   = paste0(
      "Number of GMM Entropy Classes Included\n",
      "(k = 1: highest-entropy class only;  ",
      "k = N\u22121: all classes except lowest)"
    )
  ) +

  scale_y_continuous(
    name   = "Pearson Mantel Correlation Coefficient (r)",
    labels = scales::number_format(accuracy = 0.001)
  ) +

  facet_wrap(~ facet_label, ncol = 1L, scales = "free_x") +

  labs(
    title = paste0(
      "Gower Distance Matrix Correlation: Full (1,273 Sites) vs ",
      "GMM Entropy-Reduced\n",
      "Pearson Mantel Test by Number of Accumulated GMM Entropy Classes"
    ),
    subtitle = paste0(
      "Diamond markers and dashed lines indicate the optimal k (maximum Pearson r). ",
      "Spearman r and p-value are annotated at optimal k.\n",
      "Mantel test: ",
      format(N_PERM, big.mark = ","),
      " permutations, ",
      N_CORES, " parallel cores. ",
      "Sequences deduplicated prior to Gower distance computation."
    ),
    caption = paste0(
      "NCBI US: January 2020 \u2013 September 2021 (109,536 sequences).  ",
      "GISAID US: March 2020 \u2013 June 2022 (129,371 sequences)."
    )
  ) +

  theme_bw() +
  theme(
    plot.title       = element_text(hjust = 0.5, size = 13, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, size  = 8.5, colour = "grey40"),
    plot.caption     = element_text(hjust = 0,   size  = 8,   colour = "grey50"),
    axis.title       = element_text(size = 11),
    axis.text        = element_text(size = 10),
    strip.text       = element_text(size = 11, face = "bold"),
    strip.background = element_rect(fill = "grey92", colour = "grey70"),
    legend.position  = "none",
    panel.grid.minor = element_line(colour = "grey93")
  )

print(p)

fig_path <- file.path(OUT_DIR, "correlation_analysis_figure.png")
ggplot2::ggsave(
  filename = fig_path,
  plot     = p,
  width    = 12,
  height   = 10,
  dpi      = 300
)
message("Figure saved: ", fig_path)

```

# Session Information

```{r session-info}

sessionInfo()

```

# References
