---
title: "Reproducible Preprocessing of 16.7 Million GISAID SARS-CoV-2 Spike Protein Sequences in R"
author: "Vadim Tyuryaev, Jane Heffernan, Hanna Jankowski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    includes:
      in_header: meta_gisaid.html
bibliography: references.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Overview

This document records the complete preprocessing pipeline applied to GISAID
SARS-CoV-2 Spike protein sequences downloaded in April 2024. The goal is to
produce a clean, integer-encoded amino acid data matrix in the format required
by `partition_time_windows()` and the broader `ViralEntropR` detection
pipeline.

The pipeline proceeds in six stages:

1. Load the GISAID fasta file and the reference sequence.
2. Extract and validate collection dates and country annotations.
3. Filter to United States sequences within a biologically justified date
   window and sequence length range.
4. Remove sequences containing ambiguous amino acid residues.
5. Align short sequences (width < 1273 aa) against the reference using
   ClustalOmega; merge with full-length sequences.
6. Encode the aligned matrix and assemble the final data frame.

Elapsed times are recorded at each stage using `tictoc` to provide an honest
account of the computational cost of pre-processing at this scale.

> **Computational note.** Estimated peak RAM consumption is approximately 40 GB.
> Every effort has been made to release large intermediate objects explicitly
> via `rm()` and `gc()` to minimise memory pressure throughout the pipeline. The
> dominant wall-clock cost is the alignment step itself, which takes 
> approximately 6 hours on a Linux server. 

<!--
RENDERING NOTE
--------------
Do NOT knit from the RStudio Server "Knit" button — the session may be
aborted by the server, losing all progress after the last checkpoint save.

Instead, render from the terminal via nohup so the process survives
session disconnection:

  nohup Rscript -e \
    "rmarkdown::render('/home/path/to/GISAID_Preprocessing_Pipeline.Rmd')" \
    > render_log.txt 2>&1 &

Monitor progress:
  tail -f render_log.txt

If the knit is interrupted after Stage 5 has completed (i.e.
seq_US_GISAID_aligned_merged_1273aa.rds exists on disk), resume from
Stage 6 only by:
  1. Uncommenting the readRDS() line in the reextract-metadata-final chunk.
  2. Setting eval=FALSE on all Stage 1-5 chunks.
  3. Re-rendering via nohup as above.
-->

---

# Prerequisites

## Data download

The GISAID Spike protein dataset (`spikeprot0410.fasta`) must be downloaded
manually before running this pipeline. GISAID data access requires
registration and agreement to the GISAID Data Access Agreement.

1. Visit the [GISAID EpiCov Database](https://www.epicov.org) and log in.
2. Navigate to **EpiCov → Downloads**.
3. Download the file named `spikeprot0410.fasta` (or the nearest equivalent
   release).

A reference sequence (`Ref_Seq.fasta`) corresponding to the canonical
SARS-CoV-2 Spike protein [GenBank: YP_009724390.1, 1273 aa](https://www.ncbi.nlm.nih.gov/protein/1796318598) is also required.

## Install BiocManager and relevant packages

```{r, eval=FALSE}

# Install BiocManager if you don't already have it
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Biostrings")
BiocManager::install("msa")
BiocManager::install("DECIPHER")

```

## Libraries

```{r libraries}
library(Biostrings)
library(dplyr)
library(ggplot2)
library(lubridate)
library(msa)
library(tictoc)
library(zoo)
library(ViralEntropR)
```

---

# Stage 1: Load fasta_GISAID Data

The raw GISAID file contains sequences from all countries with dates spanning
the full pandemic period. We load the file and record the total sequence count 
before any filtering.

```{r load-fasta_GISAID}
tic("Stage 1: Load fasta_GISAID")
fasta_GISAID   <- readAAStringSet("spikeprot0410.fasta", format = "fasta")
Ref_Seq <- readAAStringSet("Ref_Seq.fasta")
toc()

n_seq_orig <- length(fasta_GISAID)
sprintf("Raw sequence count: %d", n_seq_orig)
```

---

# Stage 2: Date and Country Extraction

## Extract collection dates — first pass

GISAID sequence headers follow the pipe-delimited format:

```
Spike|hCoV-19/USA/OH-PLMI-HCFL5-xxxxxx/2021|2021-xx-xx|EPI_ISL_xxxxxxx|Original|hCoV-19^^Ohio|Human|
```

`extract_fasta_dates()` parses the date field using the `yyyy-mm-dd` pattern.
Dates where the day component is recorded as `00` (a common GISAID convention
indicating unknown collection day) are corrected to `01` and stored in
`$corrected_dates`. A leading zero for month or day is permitted.

```{r extract-dates-pass1}
tic("Stage 2a: Extract dates (pass 1)")
dates_result <- extract_fasta_dates(fasta_GISAID, option = 1)
toc()

dates_result$message

n_miss <- length(dates_result$missing_id)
sprintf("Sequences with missing or unparseable dates: %d", n_miss)

# Inspect a sample of headers with missing dates to confirm they are
# year-only records (e.g. "2021") rather than structurally malformed entries
sample(fasta_GISAID[dates_result$missing_id], 1000) %>% names() %>% head(10)
```

Sequences lacking complete month and day collection data (e.g., '2021-00-00') are
removed prior to downstream analysis. Reliable monthly resolution is necessary 
for downstream temporal partitioning.

```{r remove-missing-dates}
fasta_GISAID <- fasta_GISAID[-dates_result$missing_id]

n_seq_filtered_1 <- length(fasta_GISAID)

# Arithmetic sanity check: removed exactly n_miss sequences
stopifnot(n_miss == n_seq_orig - n_seq_filtered_1)
sprintf("Sequences remaining after date filtering: %d", n_seq_filtered_1)
```

## Extract country annotations

Country labels are encoded in the third pipe-delimited field of the GISAID
header. `extract_fasta_countries()` extracts this field for each sequence.
Date vectors are discarded at this point and will be re-extracted after the
country filter to ensure both metadata vectors remain aligned to the same
fasta_GISAID object.

```{r extract-countries}
tic("Stage 2b: Extract countries")
countries_result <- extract_fasta_countries(fasta_GISAID, position = 3)
toc()

countries_result$message
countries_result$missing_id        # expect NA (no missing country fields)

sum(is.na(countries_result$countries))             # 0 NAs
sum(countries_result$countries == "USA")           # 5,081,078 labelled "USA"

# Country distribution — reveals other US labels: "United States", "US"
countries_result$countries %>%
  table() %>%
  sort(decreasing = TRUE)%>%head(50)
```

## Re-extract dates and perform thorough validation

After country extraction, dates are re-extracted from the current fasta_GISAID object
(which had missing-date sequences removed) to produce two fully aligned
metadata vectors. Four independent NA checks confirm that the corrected date
vector is complete and that the `dd=00` correction propagated correctly.

```{r reextract-dates-validate}
tic("Stage 2c: Re-extract dates and validate")
dates_result <- extract_fasta_dates(fasta_GISAID, option = 1)
toc()

dates_result$message

# Four-way NA check: raw strings, corrected strings, raw dates, corrected dates
sum(is.na(dates_result$raw_date_strings))       # should be 0
sum(is.na(dates_result$corrected_date_strings)) # should be 0
sum(is.na(dates_result$raw_dates))              # non-zero: dd=00 entries
sum(is.na(dates_result$corrected_dates))        # should be 0 — confirmed

# Inspect headers where raw dates are NA (dd=00) to confirm the
# corrected_dates field handles them appropriately
# Set the seed using the date format YYYYMMDD (April 6, 1928)
set.seed(19280406)
sample(fasta_GISAID[which(is.na(dates_result$raw_dates))], 1000) %>% names() %>% head(5)

# dd is labelled 00 --> corrected_dates replaces 00 with 01
set.seed(19280406)
sample(dates_result$corrected_date_strings[which(is.na(dates_result$raw_dates))], 1000)%>% head(5)

sprintf("Date range: %s  to  %s",
        min(dates_result$corrected_dates),
        max(dates_result$corrected_dates))

# Verify metadata vectors are aligned
stopifnot(
  length(fasta_GISAID) == length(dates_result$corrected_dates),
  length(fasta_GISAID) == length(countries_result$countries)
)
```

## Date range filter — remove pre-pandemic and out-of-window sequences

The SARS-CoV-2 pandemic began in late 2019. Sequences dated before
`2019-12-01` cannot be genuine SARS-CoV-2 Spike protein records. Similarly, 
sequences beyond the date of data base download (`2024-04-10`) are out of scope. 
Both are identified and removed.

```{r date-range-filter}
start_date <- as.Date("2019-12-01")
end_date   <- as.Date("2024-04-10")

# Identify out-of-window sequences and inspect their headers
out_of_window_idx <- which(
  dates_result$corrected_dates < start_date |
  dates_result$corrected_dates > end_date
)

sprintf("Sequences outside the analysis window: %d", length(out_of_window_idx))

# Inspect suspicious sequence headers
if (length(out_of_window_idx) > 0) {
  cat("Sample of out-of-window sequence headers:\n")
  print(head(names(fasta_GISAID)[out_of_window_idx], 20))

  # Remove out-of-window sequences
  fasta_GISAID            <- fasta_GISAID[-out_of_window_idx]
  dates_result     <- extract_fasta_dates(fasta_GISAID, option = 1)
  countries_result <- extract_fasta_countries(fasta_GISAID, position = 3)
}

rm(out_of_window_idx)
gc()

sprintf("Sequences remaining after date range filter: %d", length(fasta_GISAID))
```

## Collection date distribution

```{r date-distribution, fig.width=10, fig.height=4}
date_df <- data.frame(
  Date = as.Date(dates_result$corrected_dates)
) %>%
  mutate(YearMonth = format(Date, "%Y-%m"))

p=ggplot(date_df, aes(x = YearMonth)) +
  geom_bar(stat = "count", fill = "steelblue") +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    plot.title   = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 1, colour = "grey45", size = 8)
  ) +
  labs(
    x       = "Collection Month",
    y       = "Number of Sequences",
    title   = "GISAID SARS-CoV-2 Spike Sequences — Collection Date Distribution",
    caption = "Day-00 dates corrected to the first day of the recorded month."
  )

print(p)

ggsave(filename = "GISAID_Collection_Distribution_Clean_All_Sequences.png", 
       plot = p, 
       width = 10,     # Width of the image in inches
       height = 6,     # Height of the image in inches
       dpi = 300,      # Resolution (300 is standard for publications)
       bg = "white")   # Ensures the background is white, not transparent

rm(date_df,p)
gc()

```

---

# Stage 3: Geographic and Length Filtering

## Filter to United States sequences

Analyses are restricted to United States sequences to minimise confounding
from inter-country sampling heterogeneity and surveillance intensity
differences. Analysis above reveals three country labels for the United States:
`"USA"`, `"United States"`, and `"US"`. Combined, these account for
5,081,317 sequences (5,081,078 + 121 + 118).

```{r filter-usa}
tic("Stage 3a: Filter to USA")
us_idx <- which(countries_result$countries %in% c("USA", "United States", "US"))
fasta_GISAID  <- fasta_GISAID[us_idx]
toc()

rm(us_idx)
gc()

# Re-align metadata
dates_result     <- extract_fasta_dates(fasta_GISAID, option = 1)
countries_result <- extract_fasta_countries(fasta_GISAID, position = 3)

sprintf("US sequences: %d", length(fasta_GISAID))
```

## Filter by sequence length

The canonical SARS-CoV-2 spike protein (`YP_009724390.1`) is 1273 amino acids in 
length. GISAID sequences of width 1274 include a terminal stop codon (*), which 
is removed during preprocessing. Sequences shorter than 1268 amino acids typically
reflect large deletions relative to the reference and disrupt the positional 
correspondence required by the fixed-width representation used in this pipeline,
while sequences longer than 1274 amino acids contain insertions relative to the 
reference. Both cases are excluded to preserve consistent site-wise comparability 
across sequences.

```{r filter-width}
tic("Stage 3b: Filter by sequence length")
fasta_GISAID <- fasta_GISAID[width(fasta_GISAID) >= 1268L & width(fasta_GISAID) <= 1274L]
toc()

# Re-align metadata
dates_result     <- extract_fasta_dates(fasta_GISAID, option = 1)
countries_result <- extract_fasta_countries(fasta_GISAID, position = 3)

# Width distribution after length filter
width(fasta_GISAID) %>% table() %>% sort(decreasing = TRUE)
```

## Remove sequences containing ambiguous residues

Sequences containing ambiguous amino acid codes — B (Asp/Asn), J (Leu/Ile), 
X (unknown), and Z (Glu/Gln) — are removed prior to analysis. These symbols 
encode uncertainty rather than distinct residues and, if treated as separate 
categories, can bias frequency-based measures such as Shannon entropy by 
inflating the effective alphabet size. This introduces noise into downstream 
distance calculations and clustering, potentially obscuring true patterns of 
sequence variability.

```{r filter-ambiguous-pre-alignment}
tic("Stage 3c: Remove ambiguous residues")
keep_idx <- !grepl("X|B|Z|J", as.character(fasta_GISAID), ignore.case = TRUE)
fasta_GISAID    <- fasta_GISAID[keep_idx]
toc()

rm(keep_idx)
gc()

# Re-align metadata
dates_result     <- extract_fasta_dates(fasta_GISAID, option = 1)
countries_result <- extract_fasta_countries(fasta_GISAID, position = 3)

# Width distribution after ambiguous residue removal
width(fasta_GISAID) %>% table() %>% sort(decreasing = TRUE)

saveRDS(fasta_GISAID,            "seqs_US_filtered_unaligned.rds")
saveRDS(dates_result,     "metadata_dates_US_filtered_unaligned.rds")
saveRDS(countries_result, "metadata_countries_US_filtered_unaligned.rds")
```

---

# Stage 4: Deduplication and Stop Codon Removal

Exact duplicate sequences provide no additional information for entropy
calculations and inflate computation time substantially. We deduplicate
separately for sequences requiring alignment (width ≤ 1273) and sequences of
canonical width (1274, including the stop codon).

```{r split-and-deduplicate}
tic("Stage 4a: Split and deduplicate")
fasta_GISAID_short   <- unique(fasta_GISAID[width(fasta_GISAID) != 1274L])
fasta_GISAID_regular <- unique(fasta_GISAID[width(fasta_GISAID) == 1274L])
toc()

rm(fasta_GISAID)
gc()

sprintf("Unique short sequences (width 1268-1273): %d", length(fasta_GISAID_short))
sprintf("Unique regular sequences (width 1274):    %d", length(fasta_GISAID_regular))
```

Terminal stop codons (*) are removed prior to alignment. While treated as valid 
symbols by Clustal Omega, their presence introduces non-biological variation at 
the C-terminus, which can lead to artificial mismatches and, in cases of 
inconsistent sequence lengths, may contribute to spurious gap placement at
terminal positions.

The `endoapply` function is a specialized iteration method in R that applies a
given function across all elements of a list-like object while returning an 
output of the exact same S4 class as the original input.

We utilize it in this step to ensure that removing the terminal stop codons 
preserves the native Bioconductor sequence class (`AAStringSet`) of the data,
preventing the sequences from being unintentionally coerced into a standard R list.

```{r remove-stop-codons}
remove_stop_codon <- function(seq) {
  AAString(gsub("\\*", "", as.character(seq)))
}

tic("Stage 4b: Remove stop codons (short sequences)")
fasta_GISAID_short <- endoapply(fasta_GISAID_short, remove_stop_codon)
toc()

tic("Stage 4c: Remove stop codons (regular sequences)")
fasta_GISAID_regular <- endoapply(fasta_GISAID_regular, remove_stop_codon)
toc()

# Confirm stop codons removed from both sets
stopifnot(!any(grepl("\\*", as.character(fasta_GISAID_short))))
stopifnot(!any(grepl("\\*", as.character(fasta_GISAID_regular))))

saveRDS(fasta_GISAID_short,   "seqs_US_GISAID_unaligned_unique_short.rds")
saveRDS(fasta_GISAID_regular, "seqs_US_GISAID_unaligned_unique_1273aa.rds")
```

---

# Stage 5: Multiple Sequence Alignment (MSA)

## Rationale

Sequences of width 1268–1273 carry one to five amino acid deletions relative to
the reference. To assign site positions consistently across all sequences, each
short sequence is aligned individually against the 1273-aa reference using 
Clustal Omega in pairwise mode. The reference sequence serves as an anchor for 
positional correspondence and is discarded from the output; only the aligned 
form of each query sequence is retained.

The parameter **threads = 1** is specified explicitly. For a two-sequence 
alignment (one query and one reference), Clustal Omega constructs a trivial 
guide tree and performs minimal profile alignment, resulting in a computational
workload that is inherently sequential. In this setting, increasing the number 
of threads provides no practical performance benefit and may introduce 
unnecessary overhead.

## Test run on 100 sequences

To validate the alignment logic and forecast computational requirements, a 
preliminary benchmark was conducted on a randomized sample of 100 sequences. 
This approach yields a robust empirical average of the per-sequence runtime, 
acknowledging that individual execution times may fluctuate depending on 
sequence divergence (e.g., insertion/deletion events) and dynamic hardware
allocation.

```{r alignment-test}
perform_alignment <- function(sequences, reference) {
  combined <- c(reference, sequences)
  env      <- environment()
  capture.output(suppressMessages(
    env$result <- msa(combined,
                      method  = "ClustalOmega",
                      order   = "input",
                      type    = "protein",
                      threads = 1L,
                      verbose = FALSE,
                      help    = FALSE)
  ))
  env$result@unmasked[-1L]
}

# 1. Set reproducible seed for the test sample
set.seed(19760727) 

# 2. Sample 100 distinct sequences
test_seq <- sample(fasta_GISAID_short, 100L)

# 3. Split into 100 individual chunks (each containing exactly 1 sequence)
test_chunks <- split(test_seq, seq_along(test_seq))

# 4. Time the one-by-one alignment process
tic("Test: 100 individual sequence alignments")
aligned_test <- lapply(test_chunks, perform_alignment, Ref_Seq)
aligned_test <- Reduce(c, aligned_test)
timer <- toc()

# 5. Extract elapsed time to calculate the per-sequence average
total_test_sec <- timer$toc - timer$tic
sec_per_seq    <- total_test_sec / 100
total_seqs     <- length(fasta_GISAID_short)

est_total_seconds <- sec_per_seq * total_seqs
est_total_hours   <- est_total_seconds / 3600

# 6. Print the projection
sprintf("Empirical average per sequence: %.4f seconds", sec_per_seq)
sprintf("Estimated production time for %d sequences: %.2f hours", 
            total_seqs, est_total_hours)

rm(test_seq,test_chunks,aligned_test)
gc()

```

## Production alignment (~6 hours)

```{r alignment-production, eval=FALSE}

tic("Stage 5 production: Align ~121,621 sequences")
prod_chunks  <- split(fasta_GISAID_short, seq_along(fasta_GISAID_short))
aligned_prod <- lapply(prod_chunks, perform_alignment, Ref_Seq)
aligned_prod <- Reduce(c, aligned_prod)
toc()

saveRDS(aligned_prod, "seq_US_GISAID_pairwise_aligned_short_vs_Ref.rds")

# Width distribution — note that some sequences may be longer than 1273 aa
# due to insertions introduced relative to the reference during alignment
width(aligned_prod) %>% table() %>% sort(decreasing = TRUE)

```

## Inspect alignment quality

```{r alignment-quality, eval=FALSE}
library(DECIPHER)

# Combine your reference and a random sample of 100 sequences
view_sample <- c(Ref_Seq, sample(aligned_prod, 100L))

# Open in browser. Setting highlight = 1 colors only the mutations 
# that differ from the first sequence (reference)!
BrowseSeqs(view_sample, highlight = 1)
```

```{r restart-after-crash, echo=FALSE, eval=TRUE}
aligned_prod <- readRDS("seq_US_GISAID_pairwise_aligned_short_vs_Ref.rds")
```

## Post-alignment width filter and merge

After alignment, sequences at irregular widths (outside 1273 aa) are discarded
— these are sequences for which ClustalOmega introduced an insertion relative
to the reference, which cannot be accommodated by the fixed-width pipeline.
The retained aligned sequences are merged with the full-length sequences.

```{r merge-aligned}
tic("Stage 5: Filter to width 1273 and merge")
aligned_prod_1273 <- aligned_prod[width(aligned_prod) == 1273L]

sprintf("Aligned short sequences retained at width 1273: %d",
        length(aligned_prod_1273))

fasta_GISAID_merged <- c(aligned_prod_1273, fasta_GISAID_regular)
toc()

# Confirm all merged sequences are width 1273
width(fasta_GISAID_merged) %>% table()

saveRDS(fasta_GISAID_merged, "seq_US_GISAID_aligned_merged_1273aa.rds")

rm(fasta_GISAID_short, fasta_GISAID_regular, aligned_prod,
   aligned_prod_1273, prod_chunks)
gc()
```

---

# Stage 6: Matrix Encoding and Data Frame Assembly

## Re-extract metadata for merged sequences

The merged fasta_GISAID object contains sequences from both the aligned-short and
full-length pools. Metadata vectors are re-extracted to ensure they are
aligned to the final merged sequence set. The reference sequence (Wuhan
isolate, position 1) is removed — it is not part of the surveillance dataset.

```{r reextract-metadata-final}

tic("Stage 6a: Re-extract metadata for merged set")
dates_final     <- extract_fasta_dates(fasta_GISAID_merged, option = 1)
countries_final <- extract_fasta_countries(fasta_GISAID_merged, position = 3)
toc()

dates_final$message
sprintf("Missing dates:     %d",
        if (all(is.na(dates_final$missing_id))) 0L
        else length(dates_final$missing_id))

sprintf("Missing countries: %d",
        if (all(is.na(countries_final$missing_id))) 0L
        else length(countries_final$missing_id))

sprintf("Final date range: %s  to  %s",
        min(dates_final$corrected_dates),
        max(dates_final$corrected_dates))
```

## Final date distribution

```{r date-distribution-final, fig.width=10, fig.height=4}
date_df_final <- data.frame(
  Date = as.Date(dates_final$corrected_dates)
) %>%
  mutate(YearMonth = format(Date, "%Y-%m"))

p=ggplot(date_df_final, aes(x = YearMonth)) +
  geom_bar(stat = "count", fill = "steelblue") +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    plot.title   = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 1, colour = "grey45", size = 8)
  ) +
  labs(
    x       = "Collection Month",
    y       = "Number of Sequences",
    title   = "GISAID USA Sequences — Final Date Distribution (Post-Processing)",
    caption = "Day-00 dates corrected to the first day of the recorded month."
  )

print(p)

ggsave(filename = "GISAID_Collection_Distribution_US_Seq_PostProcessing.png", 
       plot = p, 
       width = 10,     # Width of the image in inches
       height = 6,     # Height of the image in inches
       dpi = 300,      # Resolution (300 is standard for publications)
       bg = "white")   # Ensures the background is white, not transparent

rm(date_df_final,p)
gc()

```

## Convert to character matrix

Function `fasta_to_char_matrix()` converts the `AAStringSet` to a plain character
matrix (rows = sequences, columns = sites 1–1273). All sequences at this
stage have been verified to be 1273 aa.

```{r fasta_GISAID-to-char-matrix}
tic("Stage 6b: fasta_GISAID_to_char_matrix")
char_mat <- fasta_to_char_matrix(fasta_GISAID_merged)
rm(fasta_GISAID_merged)
toc()

gc()

sprintf("Character matrix dimensions: %d sequences x %d sites",
        nrow(char_mat), ncol(char_mat))
```

## Filter residual ambiguous residues

A second pass of `filter_ambiguous_sequences()` catches any ambiguous residues
or alignment gaps introduced during the ClustalOmega step. `option = 2`
specifies a character matrix input.

```{r filter-ambiguous-post-alignment}
tic("Stage 6c: filter_ambiguous_sequences (post-alignment)")
filtered <- filter_ambiguous_sequences(char_mat, option = 2)
rm(char_mat)
toc()

gc()

cat(filtered$OriginalDim,     "\n")
cat(filtered$NewDim,          "\n")
cat(filtered$NumberAmbiguous, "\n")

clean_char_mat  <- filtered$FilteredMatrix
deleted_rows    <- filtered$DeletedSeqId

rm(filtered)
gc()

# Align metadata
corrected_dates <- if (length(deleted_rows) > 0L) {
  dates_final$corrected_dates[-deleted_rows]
} else {
  dates_final$corrected_dates
}

countries <- if (length(deleted_rows) > 0L) {
  countries_final$countries[-deleted_rows]
} else {
  countries_final$countries
}

stopifnot(
  nrow(clean_char_mat) == length(corrected_dates),
  nrow(clean_char_mat) == length(countries)
)
```

## Encode amino acids

Function `encode_aa_sequence()` maps each single-letter amino acid code to its
integer representation using the 25-character `ViralEntropR` alphabet. After 
two-stage ambiguous residue filtering, only codes 1–20 (the 20 standard amino acids)
and code 25 (alignment gap, `-`) should be present.

```{r encode}
tic("Stage 6d: encode_aa_sequence")
int_mat <- encode_aa_sequence(clean_char_mat)
rm(clean_char_mat)
toc()

gc()

code_range <- range(int_mat)
sprintf("Encoded value range: %d to %d", code_range[1], code_range[2])

```

## Assemble the final data frame

The integer matrix is combined with aligned date and country metadata into a
single data frame. Dates are standardised to the first of each collection
month for compatibility with `partition_time_windows()`, and rows are sorted
by date. The unfiltered matrix is saved as an intermediate checkpoint before
the final cleaned object is written.

```{r assemble-dataframe}
tic("Stage 6e: Assemble data frame")
n_seq   <- nrow(int_mat)
n_sites <- ncol(int_mat)

AL_df           <- as.data.frame(int_mat)
colnames(AL_df) <- as.character(seq_len(n_sites))
AL_df[]         <- lapply(AL_df, as.integer)

AL_df$Date    <- as.Date(format(as.Date(corrected_dates), "%Y-%m-01"))
AL_df$Country <- "USA"

rownames(AL_df) <- as.character(seq_len(n_seq))

AL_df <- AL_df[order(AL_df$Date), ]
toc()

# Apply final date-ordered row reset
rownames(AL_df) <- NULL

sprintf("Final data frame: %d sequences x %d columns (%d sites + Date + Country)",
        nrow(AL_df), ncol(AL_df), n_sites)

saveRDS(AL_df, "GISAID_US_aligned_feature_matrix_1273aa.rds")

rm(int_mat)
gc()
```

---

# Session Information

```{r session-info}
sessionInfo()
```

# References

